\documentclass{ximera}

\input{../../preamble.tex}

\author{Jim Talamo}
\license{Creative Commons 3.0 By-NC}


\outcome{Set up an integral or sum of integrals that gives the volume of a solid with known cross sections.}
\outcome{Explore how types cross sections affects the volume.}

\begin{document}
\begin{exercise}


The region $R$ in the first quadrant is bounded by $y=\frac{12}{x}$, $x=1$ and $x=3$.  This exercise considers three solids whose base is $R$, but whose cross sections differ from each other. 


          \begin{image}
\begin{tikzpicture}
\begin{axis}[
            domain=0.25:4, ymax=14,xmax=4, ymin=-1, xmin=-0.5,
            axis lines =center, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top,
          ]
          
\addplot [draw=penColor,very thick,smooth] {12/x};
\addplot [draw=penColor2,very thick,smooth] coordinates {(1,0)(1,12)};
\addplot [draw=penColor3,very thick,smooth] coordinates {(3,0)(3,4)};

           
\addplot [name path=A,domain=1:3,draw=none] {12/x};  
\addplot [name path=B,domain=1:3,draw=none] {0};
\addplot [fillp] fill between[of=A and B];

         
\node at (axis cs:2,9) [penColor] {$y=\frac{12}{x}$};
\end{axis}
\end{tikzpicture}
\end{image}


For Solid 1, cross sections through the solid taken parallel to the $y$-axis are squares.

For Solid 2, cross sections through the solid taken parallel to the $y$-axis are equilateral triangles.

For Solid 3, cross sections through the solid taken parallel to the $y$-axis are semicircles.

\begin{exercise}
Using geometric reasoning only (i.e. without performing any calculations):

Which solid has the greatest volume?
\begin{multipleChoice}
\choice[correct]{Solid 1}
\choice{Solid 2}
\choice{Solid 3}
\end{multipleChoice}

Which solid has the least volume?
\begin{multipleChoice}
\choice{Solid 1}
\choice{Solid 2}
\choice[correct]{Solid 3}
\end{multipleChoice}

\begin{hint}
How do the areas of a square of side length $s$, and semicircle with diameter $s$, and an equilateral triangle with side length $s$ compare?  How can this be used to compare the volumes of the three given solids?
\end{hint}

\end{exercise}

\begin{exercise}
To verify your observations quantitatively, calculate the volume of each solid.

The volume of Solid 1 is $\answer{96}$ cubic units.
	
The volume of Solid 2 is $\answer{24 \sqrt{3}}$ cubic units.

The volume of Solid 3 is $\answer{12 \pi }$ cubic units.

\end{exercise}


\end{exercise}

\end{document}