\documentclass{ximera}

\input{../preamble.tex}

\title[Dig-In:]{L'H\^{o}pital's rule}

\outcome{Recall how to find limits for forms that are not indeterminate.}
\outcome{Define an indeterminate form.}
\outcome{Convert indeterminate forms to the form zero over zero or infinity over infinity.}
\outcome{Define L'Hopital's Rule and identify when it can be used.}
\outcome{Use L'Hopital's Rule to find limits.}

\begin{document}
\begin{abstract}
  We use derivatives to give us a ``short-cut'' for computing limits.
\end{abstract}
\maketitle

Derivatives allow us to take problems that were once difficult to
solve and convert them to problems that are easier to solve. Let us
consider L'H\^{o}pital's rule:

\begin{theorem}[L'H\^{o}pital's Rule]\index{L'H\^opital's Rule} 
	Let $f(x)$ and $g(x)$ be functions that are differentiable near $a$.  If
	\[
	\lim_{x \to a} f(x) = \lim_{x \to a}g(x) = 0 \qquad \text{or} \pm \infty,
	\]
	 and $g'(x) \neq 0$
	for all $x$ near $a$, then 
	\[
	\lim_{x \to a} \frac{f(x)}{g(x)} = \lim_{x \to a} \frac{f'(x)}{g'(x)},
	\]
	provided that $\lim_{x \to a} \frac{f'(x)}{g'(x)}$ exists or $\pm \infty$.
\end{theorem}

This theorem is somewhat difficult to prove, in part because it
incorporates so many different possibilities, so we will not prove it
here. 
\begin{remark}
  L'H\^{o}pital's rule applies even when $\lim_{x\to a}f(x) = \pm \infty$
  and $\lim_{x\to a}g(x) = \mp \infty$.
\end{remark}


L'H\^{o}pital's rule allows us to investigate limits of \textit{indeterminate form}.

\begin{definition}[List of Indeterminate Forms]\index{indeterminate form}\hfil
	\begin{description}
		\item[\zeroOverZero] This refers to a limit of the form $\lim_{x\to a}
		  \frac{f(x)}{g(x)}$ where $f(x)\to 0$ and $g(x)\to 0$ as $x\to a$.
		\item[\inftyOverInfty] This refers to a limit of the form $\lim_{x\to a}
		  \frac{f(x)}{g(x)}$ where $f(x)\to \infty$ and $g(x)\to \infty$ as $x\to a$.
		\item[\zeroTimesInfty] This refers to a limit of the form $\lim_{x\to a}
		  \left(f(x)\cdot g(x)\right)$ where $f(x)\to 0$ and $g(x)\to \infty$ as $x\to a$.
		\item[\inftyMinusInfty] This refers to a limit of the form $\lim_{x\to a}\left(
		  f(x)-g(x)\right)$ where $f(x)\to \infty$ and $g(x)\to \infty$ as $x\to a$.
		\item[\oneToInfty] This refers to a limit of the form $\lim_{x\to a}
		  f(x)^{g(x)}$ where $f(x)\to 1$ and $g(x)\to \infty$ as $x\to a$.
		\item[\zeroToZero] This refers to a limit of the form $\lim_{x\to a}
		  f(x)^{g(x)}$ where $f(x)\to 0$ and $g(x)\to 0$ as $x\to a$.
		\item[\inftyToZero] This refers to a limit of the form $\lim_{x\to a}
		  f(x)^{g(x)}$ where $f(x)\to \infty$ and $g(x)\to 0$ as $x\to a$.
	\end{description}
	In each of these cases, the value of the limit is \textbf{not} immediately
	obvious. Hence, a careful analysis is required!
\end{definition}

\section{Basic indeterminant forms}


Our first example is the computation of a limit that was somewhat
difficult before.

\begin{example}
	Compute
	\[ \lim_{x\to 0} \frac{\sin(x)}{x}. \]
	\begin{explanation}
		Set $f(x) = \sin(x)$ and $g(x) = x$.  Since both $f(x)$ and $g(x)$ are
		differentiable functions at $0$, and 
		\[
		\lim_{x \to 0} f(x) = \lim_{x \to 0}g(x) = 0,
		\]
		this situation is ripe for L'H\^{o}pital's Rule. Now
		\[
		f'(x) = \answer[given]{\cos(x)}
		\]
		and
		\[
		g'(x) = \answer[given]{1}.
		\] 
		L'H\^{o}pital's rule tells us that 
		\[ \lim_{x \to 0} \frac{\sin(x)}{x} = \lim_{x \to 0} \frac{\cos(x)}{1} = 1. \]
	\end{explanation}
\end{example}

\begin{remark}
  Note, the astute mathematician will notice that in our example
  above, we are somewhat cheating. To apply L'H\^ opital's rule, we
  need to know the derivative of sine; however, to know the derivative
  of sine we must be able to compute the limit:
  \[  \lim_{x\to 0}\frac{\sin(x)}{x} \]
  Hence using L'H\^{o}pital's Rule to compute this limit is a circular
  argument! We encourage the gentle reader to view L'H\^{o}pital's rule
  a ``reminder'' as to what is true, not as the formal derivation of
  the result.
\end{remark}

\begin{example}
	Compute:
	\[ \lim_{x\to \infty} \dfrac{3x^2 + 1}{2xe^x}.\]

	\begin{explanation}
		Set $f(x) =3x^2+1$ and $g(x)=2xe^x$. 
		Since $\displaystyle \lim_{x\to \infty}f(x) = \infty$ and $\displaystyle \lim_{x\to \infty} g(x) = \infty$
		this is a limit with form $\inftyOverInfty$, so L'H\^{o}pital's Rule (LHR) applies. This gives:
		\[ \lim_{x\to \infty} \dfrac{3x^2 + 1}{2xe^x} = \lim_{x\to \infty}\dfrac{6x}{2e^x + 2xe^x}. \]
		This right-hand side is another limit with form $\inftyOverInfty$, so we can use LHR again.
		\[ \lim_{x\to \infty}\dfrac{6x}{2e^x + 2xe^x} = \lim_{x\to \infty}\dfrac{6}{4e^x + 2xe^x}. \]
		This limit has numerator tending to $6$ and denominator tending to $\infty$, making this limit have value $0$.
		
		\[ \lim_{x\to \infty}\dfrac{3x^2+1}{2xe^x} = 0\]
	\end{explanation}
\end{example}



Our next set of examples will run through the remaining indeterminate
forms one is likely to encounter.
\section{Indeterminate forms involving multiplication}

$\zeroTimesInfty$ forms arise from a limit of the form $\displaystyle \lim_{x\to a}f(x)g(x)$. One way to write $f(x)g(x)$ as a fraction remember that
\[ f(x)g(x) = \dfrac{f(x)}{\left(\dfrac{1}{g(x)}\right)} = \dfrac{g(x)}{\left(\dfrac{1}{f(x)}\right)}. \]

\begin{example}
  Compute 
\[
\lim_{x\to \pi/2^+} \frac{\sec(x)}{\tan(x)}.
\]
\begin{explanation}
Set $f(x) = \sec(x)$ and $g(x) = \tan(x)$. Both $f(x)$ and $g(x)$
are differentiable near $\pi/2$. Additionally,
\[
\lim_{x \to \pi/2^+} f(x) = \lim_{x \to \pi/2^+}g(x) = -\infty.
\]
This situation is ripe for L'H\^opital's Rule. Now 
\[
f'(x) = \answer[given]{\sec(x)\tan(x)}
\]
and
\[
g'(x) = \answer[given]{\sec^2(x)}.
\]
L'H\^{o}pital's rule tells us that 
\begin{align*}
\lim_{x\to \pi/2^+} \frac{\sec(x)}{\tan(x)} &= \lim_{x\to \pi/2^+}
\frac{\sec(x)\tan(x)}{\sec^2(x)} \\
&= \lim_{x\to \pi/2^+} \sin(x)\\
&=\answer[given]{1}.
\end{align*}
\end{explanation}
\end{example}



\begin{example}\label{example:xlnx infty} 
Compute 
\[
\lim_{x\to 0^+} x\ln(x).
\]
\begin{explanation}
This doesn't appear to be suitable for L'H\^{o}pital's Rule. As $x$
approaches zero, $\ln(x)$ goes to $-\infty$, so the product looks like
\[
(\text{something very small})\cdot (\text{something very large and
  negative}).
\] 
This product could be anything. A careful analysis is required.
Write
\[
x\ln(x) = \frac{\ln(x)}{x^{-1}}.
\]
Set $f(x) = \ln(x)$ and $g(x) = x^{-1}$.  Since both functions are differentiable near zero and 
\[
\lim_{x\to 0+} \ln(x) = -\infty\qquad\text{and}\qquad \lim_{x\to 0+} x^{-1} = \infty,
\]
we may apply L'H\^{o}pital's rule. Write with me
\[
f'(x) = \answer[given]{x^{-1}}
\]
and
\[
g'(x) = \answer[given]{-x^{-2}},
\]
so
\begin{align*}
  \lim_{x\to 0^+} x\ln(x) &= \lim_{x\to 0^+} \frac{\ln(x)}{x^{-1}} \\
  &= \lim_{x\to 0^+} \frac{x^{-1}}{-x^{-2}}\\
  &=\lim_{x\to 0^+} -x \\
  &= 0.
\end{align*}
One way to interpret this is that since $\lim_{x\to 0^+}x\ln(x) = 0$,
the function $x$ approaches zero much faster than $\ln(x)$ approaches
$-\infty$.
\end{explanation}
\end{example}


\section{Indeterminate forms involving subtraction}

There are two basic cases here, we'll do an example of each.

\begin{example}
Compute
\[
\lim_{x\to 0} \left(\cot(x) - \csc(x)\right).
\]
\begin{explanation}
Here we simply need to write each term as a fraction,
\begin{align*}
\lim_{x\to 0} \left(\cot(x) - \csc(x)\right) &= \lim_{x\to 0} \left(\frac{\cos(x)}{\sin(x)} - \frac{1}{\sin(x)}\right)\\
&= \lim_{x\to 0} \frac{\cos(x)-1}{\sin(x)} 
\end{align*}
Setting $f(x) = \cos(x)-1$ and $g(x)=\sin(x)$, both functions are differentiable near zero and 
\[
\lim_{x\to 0}(\cos(x)-1)=\lim_{x\to 0}\sin(x) = 0.
\]
We may now apply L'H\^{o}pital's rule. Write with me
\[
f'(x) = \answer[given]{-\sin(x)}
\]
and
\[
g'(x) = \answer[given]{\cos(x)},
\]
so
\begin{align*}
  \lim_{x\to 0} \left(\cot(x) - \csc(x)\right) &= \lim_{x\to 0} \frac{\cos(x)-1}{\sin(x)}\\
  &= \lim_{x\to 0} \frac{-\sin(x)}{\cos(x)} \\
  &=0.
\end{align*}
\end{explanation}
\end{example}


Sometimes one must be slightly more clever. 

\begin{example}
Compute
\[
\lim_{x\to\infty}\left(\sqrt{x^2+x}-x\right).
\]
\begin{explanation}
Again, this doesn't appear to be suitable for L'H\^{o}pital's Rule. A
bit of algebraic manipulation will help. Write with me
\begin{align*}
\lim_{x\to\infty}\left(\sqrt{x^2+x}-x\right) &= \lim_{x\to\infty}\left(x\left(\sqrt{1+1/x}-1\right)\right)\\
&=\lim_{x\to\infty}\frac{\sqrt{1+1/x}-1}{x^{-1}}
\end{align*}
Now set $f(x) = \sqrt{1+1/x}-1$, $g(x) = x^{-1}$. Since both
  functions are differentiable for large values of $x$ and 
\[
\lim_{x\to\infty} (\sqrt{1+1/x}-1) = \lim_{x\to\infty}x^{-1} = 0, 
\]
we may apply L'H\^{o}pital's rule. Write with me
\[
f'(x) = \answer[given]{(1/2)(1+1/x)^{-1/2}\cdot(-x^{-2})}
\]
and
\[
g'(x) = \answer[given]{-x^{-2}}
\]
so
\begin{align*}
\lim_{x\to\infty}\left(\sqrt{x^2+x}-x\right) &= \lim_{x\to\infty}\frac{\sqrt{1+1/x}-1}{x^{-1}} \\
&= \lim_{x\to\infty}\frac{(1/2)(1+1/x)^{-1/2}\cdot(-x^{-2})}{-x^{-2}} \\
&= \lim_{x\to\infty} \frac{1}{2\sqrt{1+1/x}}\\
&= \frac{1}{2}.
\end{align*}
\end{explanation}
\end{example}

\section{Exponential Indeterminate Forms}

There is a standard trick for dealing with the indeterminate forms
\[
\oneToInfty,\quad \zeroToZero,\quad \inftyToZero.
\]
Given $u(x)$ and $v(x)$ such that
\[
\lim_{x\to a}u(x)^{v(x)}
\]
falls into one of the categories described above, rewrite as
\[
    \lim_{x\to a}e^{\ln\left(u(x)^{v(x)}\right)}
\]
and then we rewrite as
\[
    \lim_{x\to a}e^{v(x)\ln{(u(x))}}.
\]
 Recall that the exponential function is continuous. Therefore
\[
 \lim_{x\to a}e^{v(x)\ln{(u(x))}}=e^{\left[\lim_{x\to a} v(x)\ln(u(x))\right]}. 
\]
Now we will focus on the limit
\[
\lim_{x\to a} v(x)\ln(u(x))
\]
using L'H\^{o}pital's rule. (After calculating this limit, don't forget to plug back into the exponential.)

In the last two examples below, notice that they are different exponential indeterminate forms, but we approach them with the same
technique.
\begin{example}
	First determine the form of the limit, then compute the limit.
	\[ \lim_{x\to 0^+} x ^{\sin(x)}. \]
	 Select the correct choice. The form of the limit is
	 \begin{multipleChoice}
		\choice{$\zeroOverZero$}
		 \choice{ \inftyOverInfty}
		 \choice{$\zeroTimesInfty$}
		 \choice{ \inftyMinusInfty}
		  \choice {$\oneToInfty$}
		   \choice{$ \inftyToZero$}
		   \choice[correct]{$\zeroToZero$}
	 \end{multipleChoice}
  
	\begin{explanation}
		Write 
		\[\lim_{x\to 0^+} x^{\sin(x)} = e^{\left[ \lim_{x\to 0^+} \sin(x) \ln(x) \right]}.\]
		Now we look at the limit in the exponent.
		\[ \lim_{x\to 0^+} \sin(x) \ln(x) = \lim_{x\to 0^+} \dfrac{\ln(x)}{(\sin(x))^{-1}}
				=\lim_{x\to 0^+} \dfrac{\ln(x)}{\csc(x)}\]
		Since $\displaystyle \lim_{x\to 0^+} \ln(x) = \answer[given]{-\infty}$ and $\displaystyle \lim_{x\to 0^+}\csc(x) = \answer[given]{\infty}$, we know that
		$\displaystyle \lim_{x\to 0^+} \dfrac{\ln(x)}{\csc(x)}$ has form $\inftyOverInfty$, so we can apply l'H\^{o}pital's Rule.
		
		That gives:
		\begin{align*}
			\lim_{x\to 0^+} \dfrac{\ln(x)}{\csc(x)} &= \lim_{x\to 0^+} \dfrac{ \frac{1}{x} }{-\csc(x)\cot(x)}, \textrm{ l'H\^{o}pital's Rule}\\
				&= \lim_{x\to 0^+} \dfrac{- \sin(x)\tan(x)}{x}, \textrm{ algebra}\\
				&= \lim_{x\to 0^+} \dfrac{-\sin(x)\sec^2(x) - \cos(x)\tan(x)}{1}, \textrm{l'H\^{o}pital's Rule}\\
				&= \dfrac{0 - 0}{1} = 0.
		\end{align*}
		Notice that $\displaystyle \lim_{x\to 0^+} \dfrac{- \sin(x)\tan(x)}{x}$ had form $\zeroOverZero$, we used l'H\^{o}pital's Rule again
		in that step.
		
		Remember that $\displaystyle \lim_{x\to 0^+} x^{\sin(x)} = e^{\left[ \lim_{x\to 0^+} \sin(x) \ln(x) \right]}$. We just calculated that
		$\displaystyle  \lim_{x\to0^+} \sin(x)\ln(x) = 0$, which means:
		\[\lim_{x\to 0^+} x^{\sin(x)} = \answer[given]{1}\]
	\end{explanation}

\end{example}


\begin{example}
First determine the form of the limit, then compute the limit.
\[
\lim_{x\to \infty}\left(1 + \frac{1}{x}\right)^x.
\]
  Select the correct choice. The form of the limit is
 \begin{multipleChoice}
     \choice{$\zeroOverZero$}
 \choice{ \inftyOverInfty}
 \choice{$\zeroTimesInfty$}
 \choice{ \inftyMinusInfty}
  \choice[correct] {$\oneToInfty$}
   \choice{$ \inftyToZero$}
   \choice{$\zeroToZero$}
  \end{multipleChoice}
\begin{explanation}
Write
\[
\lim_{x\to \infty}\left(1 + \frac{1}{x}\right)^x = e^{\left[\lim_{x\to \infty}x\ln\left(1 + \frac{1}{x}\right)\right]}.
\]
So now look at the limit of the exponent
\[
\lim_{x\to\infty} x\ln\left(1 + \frac{1}{x}\right) = \lim_{x\to\infty} \frac{\ln\left(1 + \frac{1}{x}\right)}{x^{-1}}.
\]
Setting $f(x) = \ln\left(1 + \frac{1}{x}\right)$ and $g(x) = x^{-1}$,
both functions are differentiable for large values of $x$ and
\[
\lim_{x\to \infty}\ln\left(1 + \frac{1}{x}\right)=\lim_{x\to \infty}x^{-1} = 0.
\]
We may now apply L'H\^{o}pital's rule. Write
\[
f'(x) = \answer[given]{\frac{-x^{-2}}{1 + \frac{1}{x}}}
\]
and
\[
g'(x) = \answer[given]{-x^{-2}},
\]
so
\begin{align*}
\lim_{x\to\infty} \frac{\ln\left(1 + \frac{1}{x}\right)}{x^{-1}} &= \lim_{x\to\infty} \frac{\frac{-x^{-2}}{1 + \frac{1}{x}}}{-x^{-2}} \\
&=\lim_{x\to\infty} \frac{1}{1 + \frac{1}{x}}\\
&=1.
\end{align*}
Hence, 
\begin{align*}
  \lim_{x\to \infty}\left(1 + \frac{1}{x}\right)^x &=e^{ \eval{\lim_{x\to \infty}x\ln\left(1 + \frac{1}{x}\right)}} \\
  &=e^{ \eval{\lim_{x\to\infty} \frac{\ln\left(1 + \frac{1}{x}\right)}{x^{-1}}}} \\
  &=e^{1} \\
  &= e.
\end{align*}
\end{explanation}
\end{example}




\end{document}

