\documentclass{ximera}

\title{Exam 3}

\begin{document}
\begin{abstract}
Click here to learn about exam 3.
\end{abstract}
\maketitle

\thispagestyle{empty}


\vfill

\textbf{How to take the exam:}
\begin{itemize}
  \item Print out the PDF of the exam (or work on a separate sheet of paper).
  \item Take the exam for around $55$ minutes. If it is easy and you
    have solved all the problems, you may finish early. If you are
    stuck, try to fight-it-out for the full time.
  \item Check your answers online with Ximera. The Ximera version of
    the practice exam will be posted online sometime after the PDF is
    posted.
\end{itemize}

\textbf{Things to know:}
\begin{itemize}
\item The real exam is \textbf{not} limited to questions from the practice
  exam.
\item \textbf{Any topic from the course is fair-game for an in-class
  examination}.
\item At least 50\% of the real exam will be very similar to content
  found on the practice exams.
\item The exam will cover content from Friday, October $19$th to
  Wednesday, November $7$th.
\end{itemize}

\noindent\textsf{{\fontsize{32}{36}\textbf{Your Name}}} 

\vspace{1cm}

\noindent\textsf{\large Recitation Leader: Your TA} 


\noindent\textsf{\large Recitation Time: Your Recitation Time} 


\noindent\textsf{\large Instructor: Your Instructor}


%% \vspace{1cm}

%% \begin{center}
%% \Large
%% \begin{tabular}{|l|l|l|l|l|l|l|l|}
%% \hline
%% Page 1 & Page 2 & Page 3 & Page 4 & Page 5 & Total: \\ \hline
%% \rule{0pt}{4em} & & & & &  \hspace*{7em}\\
%%  \hspace*{2em}  /21 & \hspace*{2em}  /22 & \hspace*{2em}  /21 & \hspace*{2em}  /12& \hspace*{2em} /24& \hspace*{7em}/100\\ \hline
%% \end{tabular}
%% \end{center}

\vspace{1cm}

{\Large
%% \item Keep in mind that if you are feeling anxious now
%%     or while taking this exam, this anxiety could be the result of
%%     negative stereotypes that are widely known in society and have
%%     nothing to do with your actual ability to do well on the exam.
  \noindent \textbf{Do not look inside the exam} until instructed to do so.\\[1em]
  \noindent By taking this exam, you are acknowledging that you are the person whose name appears above.\\[1em]
\noindent You have $55$ minutes to complete the exam.\\[1em]
\noindent Calculators or cell phones are prohibited.\\[1em]
\noindent Justify your answers for full credit.\\[1em]
\noindent Show your work for partial credit.\\[1em]
\noindent Write your answers on the included pages.\\[1em]
\noindent \textbf{Provide answers for all questions asked.}
}
\end{document}
