\documentclass{ximera}

\input{../../preamble.tex}

\author{Jason Miller}
\license{Creative Commons 3.0 By-NC}


\outcome{}


\begin{document}
\begin{exercise}
Use the method of partial fractions to determine the integral.
\[
\int \frac{2x-8}{x^2+8x+12} \d x
\]

We must decompose the rational function into simpler rational pieces which we can integrate. 
Note that the degree of the numerator is smaller than the degree of the denominator so we do not need 
to use long division. 

First we see if we can factor the denominator. 

In this case we can factor and we get:
\[
x^{2}+8x+12=(x+6)(x+2)
\]

That means, for some constants $A$ and $B$, we have:

\[
\frac{2x-8}{x^{2}+8x+12}= \frac{A}{x+6} + \frac{B}{x+2}
\]

We need to determine the constants $A$ and $B$. 

We clear denominators by multiplying both sides of the above equation by $\answer{(x+6)(x+2)}$. 

This gives us 

\[
2x-8=A(x+2) + B(x+6)
\]

We can determine the unknown coefficients $A$ and $B$ in two ways. 
We could expand the right hand side and then identify coefficients of like terms on the left and right.
Instead, for this case, we try certain convenient values of $x$ to simplify the above equation and then solve for the coefficients. 

Let us try plugging in $x=-2$ into the above equation. 

Then we get 
\[
\answer{-12 }= B\answer{4}
\]

This tells us that $B=\answer{ -3}$. 


In a similar fashion we can plug in $x=-6$. 

Then we have 

\[
\answer{-20}=A\answer{-4}
\]

Solving, we obtain $A=\answer{5}$. 

\begin{exercise}
This means our original integral can be rewritten as 

\[
\int \frac{2x-8}{x^{2}+8x+12} \d x= \int \frac{5}{x+6} \d x + \int \frac{-3}{x+2} \d x
\]

 Finally computing the antiderivative we get 

\[
\int \frac{2x-8}{x^{2}+8x+12} \d x= \answer{ 5\ln|x+6|-3\ln|x+2| + C }
\]
(Use $C$ for the constant of integration and put absolute values for natural log.)

\end{exercise}
\end{exercise}
\end{document}
