\documentclass{ximera}

\input{../preamble.tex}

\author{Bart Snapp}

\outcome{Compute tangent planes.}
\outcome{Be familiar with theorems about differentiability.}

\title[Dig-In:]{Differentiability}

\begin{document}
\begin{abstract}
  We introduce differentiability for functions of several variables and find tangent planes.
\end{abstract}
\maketitle


\section{Differentiability}

In the past you may have learned
\begin{quote}\index{tangent line}
Given a function $f$ and a number $a$ in the domain of $f$, if one can
``zoom in'' on the graph at $(a, f(a))$ sufficiently so that it
appears to be a straight line, then the function is
\dfn{differentiable}, and that line is the \dfn{tangent line} to
$f(x)$ at the point $(a,f(a))$.
\end{quote}
We illustrate this informal definition with the following diagram:
\begin{image}
\begin{tikzpicture}
  \begin{axis}[
            domain=0:6, range=0:7,
            ymin=-.2,ymax=7,
            width=6in,
            height=2.5in, %% Hard coded height! Moreover this effects the aspect ratio of the zoom--sort of BAD
            axis lines=none,
          ]   
          \addplot [draw=none, fill=textColor!10!background] plot coordinates {(.8,1.6) (2.834,5)} \closedcycle; %% zoom fill
          \addplot [draw=none, fill=textColor!10!background] plot coordinates {(2.834,5) (4.166,5)} \closedcycle; %% zoom fill
          \addplot [draw=none, fill=background] plot coordinates {(1.2,1.6) (4.166,5)} \closedcycle; %% zoom fill
          \addplot [draw=none, fill=background] plot coordinates {(.8,1.6) (1.2,1.6)} \closedcycle; %% zoom fill

          \addplot [draw=none, fill=textColor!10!background] plot coordinates {(3.3,3.6) (5.334,5)} \closedcycle; %% zoom fill
          \addplot [draw=none, fill=textColor!10!background] plot coordinates {(5.334,5) (6.666,5)} \closedcycle; %% zoom fill
          \addplot [draw=none, fill=background] plot coordinates {(3.7,3.6) (6.666,5)} \closedcycle; %% zoom fill
          \addplot [draw=none, fill=background] plot coordinates {(3.3,3.6) (3.7,3.6)} \closedcycle; %% zoom fill
          
          \addplot [draw=none, fill=textColor!10!background] plot coordinates {(3.7,2.4) (6.666,1)} \closedcycle; %% zoom fill
          \addplot [draw=none, fill=textColor!10!background] plot coordinates {(3.3,2.4) (3.7,2.4)} \closedcycle; %% zoom fill
          \addplot [draw=none, fill=background] plot coordinates {(3.3,2.4) (5.334,1)} \closedcycle; %% zoom fill          
          \addplot [draw=none, fill=background] plot coordinates {(5.334,1) (6.666,1)} \closedcycle; %% zoom fill
          

          \addplot [draw=none, fill=textColor!10!background] plot coordinates {(.8,.4) (2.834,1)} \closedcycle; %% zoom fill
          \addplot [draw=none, fill=textColor!10!background] plot coordinates {(2.834,1) (4.166,1)} \closedcycle; %% zoom fill
          \addplot [draw=none, fill=background] plot coordinates {(1.2,.4) (4.166,1)} \closedcycle; %% zoom fill
          \addplot [draw=none, fill=background] plot coordinates {(.8,.4) (1.2,.4)} \closedcycle; %% zoom fill

          \addplot[very thick,penColor, smooth,domain=(0:1.833)] {-1/(x-2)};
          \addplot[very thick,penColor, smooth,domain=(2.834:4.166)] {3.333/(2.050-.3*x)-0.333}; %% 2.5 to 4.333
          %\addplot[very thick,penColor, smooth,domain=(5.334:6.666)] {11.11/(1.540-.09*x)-8.109}; %% 5 to 6.833
          \addplot[very thick,penColor, smooth,domain=(5.334:6.666)] {x-3}; %% 5 to 6.833
          
          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(1,1)};  %% point to be zoomed
          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(3.5,3)};  %% zoomed pt 1
          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(6,3)};  %% zoomed pt 2

          \addplot [->,textColor] plot coordinates {(0,0) (0,6)}; %% axis
          \addplot [->,textColor] plot coordinates {(0,0) (2,0)}; %% axis
          
          \addplot [textColor!50!background] plot coordinates {(.8,.4) (.8,1.6)}; %% box around pt
          \addplot [textColor!50!background] plot coordinates {(1.2,.4) (1.2,1.6)}; %% box around pt
          \addplot [textColor!50!background] plot coordinates {(.8,1.6) (1.2,1.6)}; %% box around pt
          \addplot [textColor!50!background] plot coordinates {(.8,.4) (1.2,.4)}; %% box around pt
          
          \addplot [textColor!50!background] plot coordinates {(2.834,1) (2.834,5)}; %% zoomed box 1
          \addplot [textColor!50!background] plot coordinates {(4.166,1) (4.166,5)}; %% zoomed box 1
          \addplot [textColor!50!background] plot coordinates {(2.834,1) (4.166,1)}; %% zoomed box 1
          \addplot [textColor!50!background] plot coordinates {(2.834,5) (4.166,5)}; %% zoomed box 1

          \addplot [textColor] plot coordinates {(3.3,2.4) (3.3,3.6)}; %% box around zoomed pt
          \addplot [textColor] plot coordinates {(3.7,2.4) (3.7,3.6)}; %% box around zoomed pt
          \addplot [textColor] plot coordinates {(3.3,3.6) (3.7,3.6)}; %% box around zoomed pt
          \addplot [textColor] plot coordinates {(3.3,2.4) (3.7,2.4)}; %% box around zoomed pt

          \addplot [textColor] plot coordinates {(5.334,1) (5.334,5)}; %% zoomed box 2
          \addplot [textColor] plot coordinates {(6.666,1) (6.666,5)}; %% zoomed box 2
          \addplot [textColor] plot coordinates {(5.334,1) (6.666,1)}; %% zoomed box 2
          \addplot [textColor] plot coordinates {(5.334,5) (6.666,5)}; %% zoomed box 2

          \node at (axis cs:2.2,0) [anchor=east] {$x$};
          \node at (axis cs:0,6.6) [anchor=north] {$y$};
        \end{axis}
\end{tikzpicture}
%% \caption{Given a function $f$, if one can ``zoom in'' at $(a, f(a))$
%% sufficiently so that the graph seems to be a straight line,
%% then that line is the \textbf{tangent line} to $f$ at the point
%% determined by $(a, f(a))$.}
%% \label{figure:informal-tangent}
\end{image}
%% \todo{This image should be interactive.}

**INTERACTIVE IMAGE**

We would like to generalize this notion of a tangent line to functions of several variables.  We would like to say $F$  is differentiable when one can ``zoom-in'' and the graph of the function looks "flat".  

Let's start by considering functions from $\R^2\to \R$ as these are the easiest to visualize. In this case, if our function is differentiable, when we "zoom-in", graph of $F: \R^2\to \R$ should look like a plane. We call this plane the \dfn{tangent plane}. 

**INTERACTIVE IMAGE**
%%BADBAD would like an image

In fact, this is what the technical definition of differentiability of a function $F: \R^2\to \R$ says.

\begin{definition}
Let $F: \R^2\to \R$.  We say $F$ is  \dfn{differentiable} at a vector $\vec{a}$ if there exists a tangent plane $T_a(\vec{x})$ such that 
\[
\lim_{\vec{x} \rightarrow \vec{a}} \frac{|F(\vec{x})-T_a(\vec{x})|}{|\vec{x}-\vec{a}|} = 0
\]
A function is \dfn{differentiable} on an open set if it is differentiable at every point in that set.
\end{definition}

We can interpert this definition as saying that a function $F$ is differentiable at a vector$\vec{a}$ if there is a tangent plane $T_a(\vec{x})$ at that point such that $F$ approaches its  tangent plane $T_a(\vec{x})$ faster than $\vec{x}$ approaches $\vec{a}$.  That is, if one ``zooms in'' on the graph of $F$ at $(\vec{a}, F(\vec{a}))$ sufficiently, it  looks just like the tangent plane.

Now let's generalize this statement to higher dimensions.  When we zoomed in a differentiable curve from $\R \to \R$, we saw a line. When we zoomed in on a differentiable surface from $\R^2 \to \R$, we saw a plane.  Similarly, when we zoom in on a differentiable surface $F:\R^n \to \R$, we should expect to see a flat surface which looks like $\R^n$.  Formalizing this, we have

\begin{definition}
Let $F: \R^n\to \R$ and let $\vec{x}=\vector{x_1, x_2, ..., x_n}$ and $\vec{a} = \vector{a_1, a_2, ..., a_n}$.  We say $F$ is  \dfn{differentiable} at a vector $\vec{a}$ if there exists a $T_a(\vec{x})=c_0+\sum_{i=1}^n c_i (x_i-a_i)$ for some real constants $c_i$ such that 
\[
\lim_{\vec{x} \rightarrow \vec{a}} \frac{|F(\vec{x})-T_a(\vec{x})|}{|\vec{x}-\vec{a}|} = 0
\]
A function is \dfn{differentiable} on an open set if it is differentiable at every point in that set.
\end{definition}

Keep in mind that this statement of differentiability is much stronger than just saying the partial derivatives of $F$ exist at $\vec{a}$.  Because this is a limit in a higher dimension, we are saying this limit must exist and be equal for every path by which $\vec{x}$ could approach $\vec{a}$, not just the paths along the axes.  Because of this, it is very difficult to use this definition to show that a function is differentiable.  Thankfully we have the following theorem:

\begin{theorem}[A Criterion for Differentiability]
  Let $F:\R^n\to\R$ be defined on an open set $S$ containing
  $\vec{a}$.  If
  \[
  \pp[F]{x_1},\pp[F]{x_2},\dots,\pp[F]{x_n}
  \]
  are all continuous on $S$, then $F$ is differentiable on $S$.
\end{theorem}

Once we know a function is differentable we can use the following theorems to conclude it is a also continuous and the partial derivatives all exist.

\begin{theorem}[Differentiability Implies Partial Derivatives Exist]
Let $F:\R^n\to\R$ be defined on an open set $S$ containing
$\vec{a}$.  If $F$ is differentiable at $\vec{a}$, then all the partial derivatives of $F$ exist at $\vec{a}$.
\end{theorem}

\begin{theorem}[Differentiability Implies Continuity]
Let $F:\R^n\to\R$ be defined on an open set $S$ containing
$\vec{a}$.  If $F$ is differentiable at $\vec{a}$, then $F$ is
continuous at $\vec{a}$.
\end{theorem}

These theorems assure us that essentially all functions that we see in
the course of our studies here are differentiable (and hence
continuous) on their natural domains. There is a small caveat to the
theorems above: it is possible for a function $F$ to be differentiable
yet some partials may not be continuous. Such strange behavior of
functions is a source of delight for many mathematicians.

\subsection{Finding tangent planes}

In your earlier calculus courses, you often found tangent lines to
curves. To do this, you were given a function $f$, a point $x=a$, and
then you produced
\[
y = f'(a)\cdot (x-a) + f(a)
\]
as your tangent line. Using our new notation, we would write this same equation as
\[
T_a(x)=f(a)+f^{(1)}(a)\cdot (x-a)
\]


Now, for a function $F:\R^2\to\R$ which is differentiable at a vector $\vec{a}=\vector{a_1, a_2}$ we would like to use partial derivatives to find the tangent
  plane. 

We know from the definition above that our tangent plane is of the form:
\[
T_a(\vec{x})=c_0+ c_1 (x_1-a_1)+ c_2 (x_2-a_2)
\]

To find the constants, we use the following criteria about the tangent plane:

\[
F(\vec{a}) =T_a(\vec{a})
\]
\[
 F^{(1,0)}(\vec{a})=T_a^{(1,0)}(\vec{a})
\]
\[
F^{(0,1)}(\vec{a})= T_a^{(0,1)}(\vec{a})
\]
 That is, the tangent plane at $\vec{a}$ touches the surface at $\vec{a}$ and the partial derivatives all agree at $\vec{a}$.  Saying the tangent plane has the same partial derivatives as the function is another way of saying that tangent plane has to contain the tangent lines we can find using the partial derivatives.

**INSERT IMAGE**\\
It is now an easy computation to find the constants $c_0, c_1,$ and $c_2$.
\[
F(\vec{a})=T_a(\vec{a})=c_0+ c_1 (a_1-a_1)+ c_2 (a_2-a_2) =\answer[given]{c_0}
\]
\[
F^{(1,0)}(\vec{a})= T_a^{(1,0)}(\vec{a})=\answer[given]{c_1}
\]
\[
F^{(0,1)}(\vec{a})= T_a^{(0,1)}(\vec{a})=\answer[given]{c_2}
\]

Therefore, we now have the equation of the tangent plane 
\[
T_a(\vec{x})=F(\vec{a})+ F^{(1,0)}(\vec{a}) (x_1-a_1)+ F^{(0,1)}(\vec{a}) (x_2-a_2)
\]
We can rewrite this equation to look more like the equation from single variable calculus if we use the dot product.
\[
T_a(\vec{x})=F(\vec{a})+ \vector{F^{(1,0)}(\vec{a}), F^{(0,1)}(\vec{a})} \cdot (\vec{x}-\vec{a})
\]

\begin{question}
  Find a tangent plane to $F(x,y) = 3\cos(x)\sin(y)$ at $(x,y) =
  (\pi/3,\pi/6)$.
  \begin{prompt}
    \[
    z = \answer{(-3\sqrt{3}/4)\cdot (x-\pi/3) + (3\sqrt{3}/4)\cdot (y-\pi/6) + 3/4}
    \]
  \end{prompt}
\end{question}


We can also find tangent planes of surfaces that are defined
parameterically.
\begin{example}
Consider the parameterization of the unit sphere,
centered at the origin where $0\le \theta< 2\pi$ and $0\le\phi\le \pi$:
\begin{align*}
  x(\theta,\phi) &= \cos(\theta)\sin(\phi)\\
  y(\theta,\phi) &= \sin(\theta)\sin(\phi)\\
  z(\theta,\phi) &= \cos(\phi)
\end{align*}
Find a plane tangent to the sphere when $\theta = \pi/4$ and $\phi =
\pi/3$.
\begin{explanation}
  Here we'll use the parametric formula for a plane:
  \[
  \vec{L}(s,t) = \vec{p}+ s \vec{v} + t\cdot \vec{w}
  \]
  The the point $\vec{p}$ (denoted by a vector) is:
  \[
  \vec{p} = \vector{\answer[given]{\sqrt{3/8}},\answer[given]{\sqrt{3/8}},\answer[given]{1/2}}
  \]
  The vector $\vec{v}$ is given by
  \begin{align*}
    \vec{v} &=\eval{\pp{\theta}\vector{x(\theta,\phi),y(\theta,\phi),z(\theta,\phi)}}_{\substack{\theta = \pi/4\\ \phi = \pi/3}}\\
      &= \eval{\vector{\answer[given]{-\sin(\theta)\sin(\phi)},\answer[given]{\cos(\theta)\sin(\phi)},\answer[given]{0}}}_{\substack{\theta = \pi/4\\ \phi = \pi/3}}\\
      &= \vector{-\sqrt{3/8},\sqrt{3/8},0}
  \end{align*}
  And vector $\vec{w}$ is given by
  \begin{align*}
    \vec{w} &=\eval{\pp{\phi}\vector{x(\theta,\phi),y(\theta,\phi),z(\theta,\phi)}}_{\substack{\theta = \pi/4\\ \phi = \pi/3}}\\
      &= \eval{\vector{\answer[given]{\cos(\theta)\cos(\phi)},\answer[given]{\sin(\theta)\cos(\phi)},\answer[given]{-\sin(\phi)}}}_{\substack{\theta = \pi/4\\ \phi = \pi/3}}\\
      &= \vector{1/\sqrt{8},1/\sqrt{8},-\sqrt{3}/2}
  \end{align*}
  Hence our desired plane is given by:
  \[
  \vec{L}(s,t) = \begin{bmatrix}
    \answer[given]{\sqrt{3/8}-s\sqrt{3/8}+ t/\sqrt{8}}\\
    \answer[given]{\sqrt{3/8} +s\sqrt{3/8} + t/\sqrt{8}}\\
    \answer[given]{1/2 -t\sqrt{3}/2}
  \end{bmatrix}
  \]
\end{explanation}
\end{example}



\end{document}
