\documentclass{ximera}

\input{../preamble.tex}

\author{Bart Snapp}

\title[Dig-In:]{Planes in space}

\outcome{Give an implicit formula for a plane.}
\outcome{Read-off the normal vector for a plane.}
\outcome{Give a parametric formula for a plane.}


\begin{document}
\begin{abstract}
  We discuss how to find implicit and explicit formulas for planes.
\end{abstract}
\maketitle


With our modern outlook, we think of \textit{slope} as being
synonymous with lines and derivatives. In single variable calculus,
the derivative is the slope of the tangent line. However, this
identification of ``slope'' as being the ``key-player'' is something
of a new idea. Before calculus, people were often interested in
finding normal lines to curves. While this might seem odd at first,
consider this: A line is the set of all points normal to some vector.
Check it out:
\begin{image}
  \begin{tikzpicture}
    \begin{axis}[
        xmin=-.5,xmax=6.5,ymin=-.5,ymax=4.5,
        unit vector ratio*=1 1 1,
        axis lines=center,
        xlabel=$x$, ylabel=$y$,
        ticks=none,
        every axis y label/.style={at=(current axis.above origin),anchor=south},
        every axis x label/.style={at=(current axis.right of origin),anchor=west},
      ]
      \addplot[dashed,very thick,penColor,domain=-.5:6.5] plot {-2*x/3+4};
      \addplot[ultra thick,penColor2,->] plot coordinates {(3,2) (3.87,3.3)};
      \addplot[dashed,->] plot coordinates {(0,0) (3,2)};
      \addplot[dashed,->] plot coordinates {(0,0) (1.73,2.85)};
      \addplot[ultra thick,penColor3,->] plot coordinates {(3,2) (1.73,2.85)};

      \node[rotate=33.7] at (axis cs: 1.7,.8) {$\vec{p}=\vector{x_0,y_0}$};
      \node[rotate=58.74] at (axis cs: .7,1.7) {$\vec{x}=\vector{x,y}$};

      \node[below right,penColor2] at (axis cs: 3.44,2.7) {$\vec{n}$};
      \node[rotate=-33,penColor3] at (axis cs: 2.5,2.75) {$\vec{x}-\vec{p}$}; 
    \end{axis}
  \end{tikzpicture}
\end{image}
From this we see that if $\vec{p}=\vector{x_0,y_0}$ is a point
(denoted by a vector) on the line, and our normal vector is
$\vec{n}=\vector{a,b}$, then
\begin{align*}
  \vec{n}\dotp(\vec{x}-\vec{p}) &=0\\
  \vector{a,b}\dotp(\vector{x,y}- \vector{x_0,y_0}) &= 0\\
  a(x-x_0)+ b(y-y_0) &= 0
\end{align*}
is an implicit formula for the line. This means that if points $(x,y)$
make the equation above true, then those points are on the line. As we
will see, a nearly identical implicit formula defines a plane.

\begin{question}
  Give an implicit equation for a line that goes through $(-2,3)$ and
  is perpendicular to the vector $\vector{4,-5}$.
  \begin{prompt}
    \[
    \answer{4(x+2) - 5(y-3)} = 0
    \]
  \end{prompt}
  \begin{question}
    Quick, tell me a vector normal to the line $2x-3y = 4$.
    \begin{prompt}
      \[
      \vec{n} = \vector{\answer{2},\answer{-3}}
      \]
    \end{prompt}
  \end{question}
\end{question}


\section{Implicit planes}\index{plane!implicit formula}

We would like to know the implicit formula for a plane. Remember an
implicit function in $\R^3$ is one of the form:
\[
F(x,y,z) = 0
\]
Here the dot product saves the day. Recall that if $\vec{n} =
\vector{a,b,c}$ is any vector, and $\vec{x}= \vector{x,y,z}$, then the
equation
\[
\vec{n}\dotp\vec{x} = 0
\]
is solved by all vectors $\vec{x}$ that are orthogonal to
$\vec{n}$. We plotted several such vectors below:
\begin{image}%%BADBAD we should move the vectors to the origin, or change the explanation
  \begin{tikzpicture}
    \begin{axis}%
      [%tick label style={font=\scriptsize},axis on top,
	%axis lines=center,
	view={135}{25},
        %axis=false,
	name=myplot,
	%xtick={-3,3},minor tick num=2,
	%ytick={-3,3},
	%ztick={-3,3},
	ymin=-4,ymax=4,
	xmin=-4,xmax=4,
	zmin=-4, zmax=4,
	%% every axis x label/.style={at={(axis cs:\pgfkeysvalueof{/pgfplots/xmax},0,0)},xshift=-3pt,yshift=-3pt},
	%% xlabel={\scriptsize $x$},
	%% every axis y label/.style={at={(axis cs:0,\pgfkeysvalueof{/pgfplots/ymax},0)},xshift=0pt,yshift=-5pt},
	%% ylabel={\scriptsize $y$},
	%% every axis z label/.style={at={(axis cs:0,0,\pgfkeysvalueof{/pgfplots/zmax})},xshift=0pt,yshift=4pt},
	%% zlabel={\scriptsize $z$}
      ]
      \draw[thick,->,penColor2] (axis cs: 2,1,1) -- (axis cs: 2/3,2/3,2/3);
      \draw[thick,->,penColor2] (axis cs: 2,1,1) -- (axis cs: 1.057,.057,1.47);
      \draw[thick,->,penColor2] (axis cs: 2,1,1) -- (axis cs: 2,0,2);
      \draw[thick,->,penColor2] (axis cs: 2,1,1) -- (axis cs: 2.943,.529,1.943);
      \draw[thick,->,penColor2] (axis cs: 2,1,1) -- (axis cs: 10/3,4/3,4/3);
      \draw[thick,->,penColor2] (axis cs: 2,1,1) -- (axis cs: 2.943,1.943,0.529);
      \draw[thick,->,penColor2] (axis cs: 2,1,1) -- (axis cs: 2,2,0);
      \draw[thick,->,penColor2] (axis cs: 2,1,1) -- (axis cs: 1.057,1.471,0.057);
      
            
      \draw[thick,->,penColor] (axis cs: 2,1,1) -- (axis cs: 1,3,3);
    \end{axis}
  \end{tikzpicture}
\end{image}
From this we see that
\begin{align*}
  \vec{n}\dotp\vec{x} &=0\\
  \vector{a,b,c}\dotp\vector{x,y,z} &=0\\
  ax+by+cz &= 0
\end{align*}
gives the formula for a plane. Since $\vec{0} = \vec{x}$ is a
solution, this plane must pass through the origin. If we want our
plane to be located anywhere in space, we must know a point on the
plane, call it $\vec{p}=\vector{x_0,y_0,z_0}$. Putting this together, we can
now see:
\begin{quote}
  If you know
\begin{itemize}
  \item a vector $\vec{n} = \vector{a,b,c}$ and
  \item a point (given by a vector) $\vec{p} = \vector{x_0,y_0,z_0}$
\end{itemize}
then,
\begin{align*}
  \vec{n}\dotp(\vec{x}-\vec{p}) &= 0\\
  \vector{a,b,c}\dotp(\vector{x,y,z}-\vector{x_0,y_0,z_0})&=0\\
  a(x-x_0) + b(y-y_0) + c(z-z_0) &= 0
\end{align*}
is an implicit equation for a plane passing through the point
$(x_0,y_0,z_0)$ with normal vector $\vec{n}$.
\end{quote}

\begin{question}
  Find the implicit equation of a plane that passes through the point
  $(-1,-3,2)$ and with normal vector $\vector{3,-1,1}$.
\begin{onlineOnly}
  Check your answer by moving the slides below:
  \begin{center}
    \geogebra{mdB4N6Dz}{800}{600} %% https://ggbm.at/mdB4N6Dz
  \end{center}
\end{onlineOnly}
\begin{question}
  Find an explicit formula for the plane above.
  \begin{prompt}
    \[
    z = \answer{2+y-3x}
    \]
  \end{prompt}
\end{question}
\end{question}

Normal vectors not only allow us to define equations for planes but also 
they help us describe properties of planes.

\begin{definition}
  Two planes are said to be \dfn{parallel} if their normal vectors are
  parallel. Two planes are said to be \dfn{orthogonal} if their normal
  vectors are orthogonal.
\end{definition}

\begin{question}
  What is the (most obvious) normal vector for the plane
  \[
  -6x+10y-2z = 4?
  \]
  \begin{prompt}
    \[
    \vec{n} = \vector{\answer{-6},\answer{10},\answer{-2}}
    \]
  \end{prompt}
  \begin{question}
    Which of the following planes are parallel to the plane $-6x+10y-2z = 4$?
    \begin{selectAll}
      \choice{$-2.46x+3.9y-0.82z=3$}
      \choice[correct]{$-2.07x+3.45y-0.69z = 8$}
      \choice[correct]{$10.38x-17.3y+3.46z=0$}
      \choice{$15.03x-25.1y+5.02z=0$}
    \end{selectAll}
      \begin{question}
    Which of the following planes are orthogonal to the plane $-6x+10y-2z = 4$?
    \begin{selectAll}
      \choice[correct]{$34x+16y-22z=0$}
      \choice{$-54x-32y+8z=-3$}
      \choice[correct]{$-36x-42y-102z=-17$}
      \choice{$54x+31y-2z=4$}
    \end{selectAll}
  \end{question}
  \end{question}
\end{question}


\section{Parametric planes}


Given \textbf{any} two nonzero vectors in $\R^3$, $\vec{v}$ and
$\vec{w}$, such that
\[
\vec{v}\cross \vec{w} \ne \vec{0},
\]
we can produce a parametric formula for a plane by writing
\[
\vec{L}(s,t) = \vec{p} + s\cdot \vec{v} + t\cdot\vec{w},
\]
where $\vec{p}$ is a vector whose ``tip'' is on the plane, and
$\vec{v}$ and $\vec{w}$ are in the plane.
\begin{question}
  Given two nonzero vectors, $\vec{v}$ and $\vec{w}$, what does it
  mean for $\vec{v}\cross \vec{w} \ne \vec{0}$?
  \begin{prompt}
    \begin{multipleChoice}
      \choice{It means these vectors are parallel.}
      \choice[correct]{It means these vectors are not parallel.}
      \choice{It means these vectors are orthogonal.}
      \choice{It means these vectors are not orthogonal.}
    \end{multipleChoice}
  \end{prompt}
\end{question}

The vector-valued formula for a plane\index{plane!vector-valued formula}
\[
\vec{L}(s,t) = \vec{p} + s\cdot \vec{v} + t\cdot\vec{w},
\]
is very similar to our formula for a line,
\[
\vecl(t) = \vec{p} + t\vec{v}
\]
where $\vec{v}$ is a vector that points in the direction of the line,
both represent linear relationships, and hence we use similar notation
for both.

\begin{image}
  \begin{tikzpicture}
    \begin{axis}[
        xmin=-2,xmax=4,ymin=-1,ymax=3,
        unit vector ratio*=1 1 1,
        ticks=none,
        axis line style={draw=none}
      ]
      \addplot[dashed,black] plot {x/2 +3};
      \addplot[dashed,black] plot {x/2 +1.5};
      \addplot[dashed,black] plot {x/2};
      \addplot[dashed,black] plot {x/2-1.5};

      \addplot[dashed,black] plot {2*x+6};
      \addplot[dashed,black] plot {2*x+3};
      \addplot[dashed,black] plot {2*x};
      \addplot[dashed,black] plot {2*x-3};
      \addplot[dashed,black] plot {2*x-6};

      \addplot[ultra thick, penColor, ->] plot coordinates {(0,0) (2,1)};
      \addplot[ultra thick, penColor2, ->] plot coordinates {(0,0) (1,2)};
      \node[left,penColor2] at (axis cs: .5,1) {$\vec{w}$};
      \node[below,penColor] at (axis cs: 1,.5) {$\vec{v}$};
    \end{axis}
  \end{tikzpicture}
\end{image}


Now that we have \textbf{two} methods of graphing planes, let's use
both of the representations at once!

\begin{question}
  Let $\vec{v} = \vector{1,-2,1}$ and $\vec{w} =
  \vector{-1,-2,1}$. Compute $\vec{v}\cross\vec{w}$.
  \begin{prompt}
    \[
    \vec{v}\cross\vec{w} = \vector{\answer{0},\answer{-2},\answer{-4}}
    \]
  \end{prompt}
  \begin{question}
    Use your answer above to give an \textbf{implicit} equation for
    the plane that passes through the point $(3,2,1)$ that is normal
    to $\vec{v}\cross\vec{w}$.
    \begin{onlineOnly}
      Check your answer by moving the sliders below:
      \begin{center}
        \geogebra{fssBbva6}{800}{600} %https://ggbm.at/fssBbva6
      \end{center}
      \end{onlineOnly}
    \begin{question}
      Now give a vector-valued formula for the same plane using the vectors given above.
      \begin{prompt}
        \[
        \vec{L}(s,t) = \vector{3+s-t,\answer{2-2s-2t},\answer{1+s+t}}
        \]
      \end{prompt}
    \end{question}
  \end{question}
\end{question}
\end{document}
