\documentclass{ximera}

\input{../preamble.tex}

\outcome{Identify situations in which an absolute maximum or minimum is guaranteed.}

\title[Break-Ground:]{More coffee}

\begin{document}
\begin{abstract}
Two young mathematicians witness the perils of drinking too much
coffee.
\end{abstract}
\maketitle

Check out this dialogue between two calculus students (based on a true
story):

\begin{dialogue}
\item[Devyn] Riley! 
\item[Riley] Yes Devyn?
\item[Devyn] Do you like coffee? I like coffee! Sometimes I feel
  really ``bad,'' sluggish and tired. Then I drink coffee and I feel
  good! Sometimes I drink a lot of coffee!
\item[Riley] Um?
\item[Devyn] But here's the problem, see---if I drink too much, I
  become over excited and can't stop talking. I just drink coffee,
  then talk. Then drink more coffee. Then I start to feel sick. Ugh. I
  have a love hate relationship with coffee.
\item[Riley] If only there were a calculus solution to this problem!
\end{dialogue}

Remember, calculus is about studying functions. If we can ``see'' a function in the work above, maybe we can figure out how to solve it. 

\begin{problem}
  If we were to try to solve Devyn's coffee problem, what would be the best
  function to know?
  \begin{multipleChoice}
    \choice{How many donuts Devyn eats.}
    \choice[correct]{How ``good'' Devyn feels after $x$ cups of coffee.}
    \choice{How many cups of coffee Devyn drinks when Devyn feels $x$ ``good.''}
    \choice{Impossible to say.}
  \end{multipleChoice}
\begin{problem}
  If we let $f(x)$ be ``How `good' Devyn feels after $x$ cups of
  coffee.'' And we think about what Devyn says above, is there an amount Devyn can drink and feel maximally ``good?''
  \begin{multipleChoice}
    \choice[correct]{yes}
    \choice{no}
  \end{multipleChoice}
  \begin{feedback}
    ``Yes'' is probably the best answer, though we are assuming that
    Devyn feels ``bad'' without enough coffee, ``good'' with some, and
    ``bad'' again with the function continuously growing and then
    decreasing.
  \end{feedback}
\end{problem}
\end{problem}




\input{../leveledQuestions.tex}


\end{document}
