\documentclass{ximera}

\input{../preamble.tex}

\outcome{Explain why the product rule is not given by multiplying the
	derivatives of the products.}
\outcome{Apply the sum rule repeatedly to find the derivative of a product.}
\outcome{Relate the sum rule, the constant multiple rule, and the product rule.}


\title[Break-Ground:]{Derivatives of products are tricky}

\begin{document}
\begin{abstract}
Here we see a dialogue where young mathematicians talk about
derivatives of products and products of derivatives.
\end{abstract}
\maketitle

Check out this dialogue between two calculus students (based on a true
story):

\begin{dialogue}
\item[Devyn] Hey Riley, remember the sum rule for derivatives?
\item[Riley] You know I do.
\item[Devyn] What do you think that the ``product rule'' will be?
\item[Riley] Let's give this a spin:
  \[
  \dd{x} \left(f(x)\cdot g(x)\right) = f'(x) \cdot g'(x)?
  \]
\item[Devyn] Hmmm, let's give this theory an acid test. Let's try
  \[
  f(x) = x^2+1\qquad\text{and}\qquad g(x) = x^3-3x
  \]
  Now
  \begin{align*}
    f'(x)g'(x) &= (2x)(3x^2-3)\\
    &= 6x^3-6x.
  \end{align*}
\item[Riley] On the other hand,
  \begin{align*}
    f(x)g(x) &= (x^2+1)(x^3-3x)\\
    &=x^5-3x^3+x^3-3x\\
    &=x^5-2x^3-3x.
  \end{align*} 
\item[Devyn] And so, 
  \[
  \ddx \left(f(x) \cdot g(x)\right) = 5x^4-6x^2-3.
  \]
\item[Riley] Wow. Hmmm. It looks like our guess was incorrect.
\item[Devyn] I've got a feeling that the so-called ``product rule''
  might be a bit tricky.
\end{dialogue}

\begin{problem}
  Above, our intrepid young mathematicians guess that the ``product rule'' might be:
  \[
  \dd{x} \left(f(x)\cdot g(x)\right) = f'(x) \cdot g'(x)?
  \]
  Does this \textbf{ever} hold true?
  \begin{freeResponse}
  \end{freeResponse}
\end{problem}

\begin{problem}
  Our mathematicians mentioned the sum rule at the beginning of their conversation.
  The sum rule can be used on certain functions that look like products.  Find five
  examples of such functions, and take their derivatives.  What other rule could be 
  used in this case?
  \begin{freeResponse}
  \end{freeResponse}
\end{problem}

\input{../leveledQuestions.tex}


\end{document}
