\documentclass{ximera}

\input{../preamble.tex}

\outcome{}


\title[Break-Ground:]{Roxy and Yuri like food}

\begin{document}
\begin{abstract}
Here we see a dialog where two calculus students discuss the eating
habits of their cats.
\end{abstract}
\maketitle

Check out this dialog between two calculus students (based on a true
story):

\begin{itemize}
  \item[\textbf{Devyn}] Yo Riley, I was watching my two cats
    \textit{Roxy} and \textit{Yuri} eat their dry cat-kibble last
    night.
\item[\textbf{Riley}] Cats love food!
\item[\textbf{Devyn}] I know! I noticed something kinda funny though:
  Both Roxy and Yuri eat at the same time; and while I gave Yuri a
  little less food, more food was left in Yuri's dish when they
  stopped eating.

  I wonder, is there is a point in time when Roxy and Yuri have the
  exact same amount of \textbf{food} in their dishes?
\item[\textbf{Riley}] Hmmmmm. Do Roxy and Yuri both drink their water
  at the same time?
  \item[\textbf{Devyn}] They do!
\item[\textbf{Riley}] Interesting. I wonder, is there is a point in
  time when Roxy and Yuri have the exact same amount of \textbf{water}
  in their dishes?
\end{itemize}

\begin{problem}
  Write down \textbf{five} examples where this ``trick'' works, and
  \textbf{five} examples when it doesn't work.
  \begin{freeResponse}
\end{freeResponse}
\end{problem}

\begin{problem}
  What is the pattern for when the ``trick'' works and when it does not work?
  \begin{freeResponse}
\end{freeResponse}
\end{problem}



\begin{xarmaBoost}
  Write down at least \textbf{five} questions for this lecture. After
  you have your questions, label them as ``Level 1,'' ``Level 2,'' or
  ``Level 3'' where:
\begin{description}
\item[Level 1] Means you know the answer, or know exactly how to do
  this problem.
\item[Level 2] Means you think you know how to do the problem, or will
  soon learn how to do the problem.
\item[Level 3] Means you have no idea how to do the problem.
\end{description}
\begin{freeResponse}
\end{freeResponse}
\end{xarmaBoost}


\end{document}
