\documentclass{ximera}

\input{../../preamble.tex}

\author{Jim Talamo}
\license{Creative Commons 3.0 By-NC}


\outcome{Understand the relationship between sequences, sequences of partial sums, and remainders.}

\begin{document}

\begin{exercise}

Consider the series $\sum_{k=1}^{\infty} \frac{1}{k^5}$.  

Does the integral test apply?

\begin{multipleChoice}
\choice[correct]{Yes.}
\choice{No.}
\end{multipleChoice}

What does the integral test tell us?
\begin{multipleChoice}
\choice{The series converges to $\frac{1}{4}$.}
\choice[correct]{The series converges, but we do not know its value yet.}
\choice{The series diverges.}
\end{multipleChoice}

Suppose we want to approximate $\sum_{k=1}^{\infty} \frac{1}{k^5}$ by using $\sum_{k=1}^{6} \frac{1}{k^5}$.  

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{exercise}
Using the remainder results for the integral test, find the \emph{maximum} possible error made by using $\sum_{k=1}^{6} \frac{1}{k^5}$ to approximate $\sum_{k=1}^{\infty} \frac{1}{k^5}$.  Report your answer to six decimal places.

\[
\textrm{maximum possible error} \leq \answer[tolerance=.000002]{.000193}
\]


Thus, to six decimal places, $\sum_{k=1}^{\infty} \frac{1}{k^5} \leq \answer[tolerance=.000002]{1.036983}$

\begin{hint}
We can determine an upper bound for the error from the inequality $ r_n \leq \int_{n}^{\infty} \frac{1}{x^5} \d x$. 

\[
\int_n^\infty \frac{1}{x^5} \d x = \lim_{b \to \infty} \eval{\answer{-\frac{1}{4x^4}}}_n^b = \lim_{b \to \infty} \eval{-\frac{1}{4b^4}+\frac{1}{4n^4}} =\answer{\frac{1}{4n^4}}. 
\]

Since we want to approximate $\sum_{k=1}^{\infty} \frac{1}{k^5}$ by using $\sum_{k=1}^{6} \frac{1}{k^5}$, we should use $n=\answer{6}$, and doing so gives us the maximum possible error.
\end{hint}
\end{exercise}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{exercise}
Using the remainder results for the integral test, find the \emph{minimum} possible error made by using $\sum_{k=1}^{6} \frac{1}{k^5}$ to approximate $\sum_{k=1}^{\infty} \frac{1}{k^5}$.  Report your answer to six decimal places.

\[
\textrm{minimum possible error} \geq \answer[tolerance=.000002]{.000104}
\]

Thus, to six decimal places, $\sum_{k=1}^{\infty} \frac{1}{k^5} \geq \answer[tolerance=.000002]{ 1.036895}$

\begin{hint}
We can determine a lower bound for the error from the inequality $ r_n \geq \int_{n+1}^{\infty} \frac{1}{x^5} \d x$. 

\[
\int_{n+1}^\infty \frac{1}{x^5} \d x = \lim_{b \to \infty} \eval{\answer{-\frac{1}{4x^4}}}_{n+1}^b = \lim_{b \to \infty} \eval{-\frac{1}{4b^4}+\frac{1}{4n^4}} =\answer{\frac{1}{4(n+1)^4}}. 
\]

Since we want to approximate $\sum_{k=1}^{\infty} \frac{1}{k^5}$ by using $\sum_{k=1}^{6} \frac{1}{k^5}$, we should use $n=\answer{6}$, and doing so gives us the minimum possible error.

Since $\sum_{k=1}^{6} \frac{1}{k^5} = \answer[tolerance=.000002]{1.036794}$, we see that 

\begin{align*}
\sum_{k=1}^{\infty} \frac{1}{k^5} & \geq \sum_{k=1}^{6} \frac{1}{k^5} + \textrm{minimum possible error} \\
& \leq 1.036794 + .000104 \\
\end{align*}

\end{hint}
\end{exercise}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{exercise}
Use $\sum_{k=1}^{6} \frac{1}{k^5}$ and the error bounds above to approximate $\sum_{k=1}^{\infty} \frac{1}{k^5}$.

A reasonable approximation for $\sum_{k=1}^{\infty} \frac{1}{k^5}$ is $\answer[tolerance=0.000044]{1.036939}$.

\begin{hint}
From the results in the previous parts of the exercise, we know that  $\sum_{k=1}^{\infty} \frac{1}{k^5} \leq \answer[tolerance=.000002]{1.036983}$ and $\sum_{k=1}^{\infty} \frac{1}{k^5} \geq \answer[tolerance=.000002]{ 1.036895}$.  So, we can simply choose a value for the series that lies between these minimum and maximum possible values.  A common choice is the average of them.


\end{hint}
\end{exercise}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{exercise}
What is the smallest value of $N$ that should be chosen so we can provide an approximation to $\sum_{k=1}^{\infty} \frac{1}{k^5}$ to within $.001$?

\[
N=\answer{4}
\]

\begin{hint}
We want $N$ so $\int_N^{N+1} \frac{1}{x^5} \d x \leq .001$.  

\[
\int_N^{N+1} \frac{1}{x^5} \d x = \eval{\answer{-\frac{1}{4x^4}}}_N^{N+1} = \eval{\frac{1}{4N^4}-\frac{1}{4(N+1)^4}}.
\]


Thus, we set $\answer{\frac{1}{4N^4}-\frac{1}{4(N+1)^4}} \leq .001$ and find $N \geq 3.54$, and we should thus choose $N = \answer{4}$.

\begin{feedback}
To obtain an approximate value, we can calculate the error bounds, then construct the interval in which the value of the series must lie.  We can then estimate the series by using any value in that interval, but a good choice would be the midpoint.
\end{feedback}
\end{hint}
\end{exercise}

\end{exercise}
\end{document}


