\documentclass{ximera}

\input{../../preamble.tex}

\author{Jim Talamo}
\license{Creative Commons 3.0 By-NC}

%This exercise parallels remainderIntegralTest1 but only discusses the upper bound
\outcome{Understand the relationship between sequences, sequences of partial sums, and remainders.}

\begin{document}

\begin{exercise}

Consider the series $\sum_{k=1}^{\infty} \frac{1}{k^5}$.  

Does the integral test apply?

\begin{multipleChoice}
\choice[correct]{Yes.}
\choice{No.}
\end{multipleChoice}

What does the integral test tell us?
\begin{multipleChoice}
\choice{The series converges to $\frac{1}{4}$.}
\choice[correct]{The series converges, but we do not know its value yet.}
\choice{The series diverges.}
\end{multipleChoice}

Suppose we want to approximate $\sum_{k=1}^{\infty} \frac{1}{k^5}$ by using $\sum_{k=1}^{6} \frac{1}{k^5}$.  

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{exercise}
Using the remainder results for the integral test, find the \emph{maximum} possible error made by using $\sum_{k=1}^{6} \frac{1}{k^5}$ to approximate $\sum_{k=1}^{\infty} \frac{1}{k^5}$.  Report your answer to six decimal places.

\[
\textrm{maximum possible error} \leq \answer[tolerance=.000002]{.000193}
\]


Thus, to six decimal places, $\sum_{k=1}^{\infty} \frac{1}{k^5} \approx  \sum_{k=1}^{6} \frac{1}{k^5} = \answer[tolerance=.000002]{1.036790}$

\begin{hint}
We can determine an upper bound for the error from the inequality $ r_n \leq \int_{n}^{\infty} \frac{1}{x^5} \d x$. 

\[
\int_n^\infty \frac{1}{x^5} \d x = \lim_{b \to \infty} \eval{\answer{-\frac{1}{4x^4}}}_n^b = \lim_{b \to \infty} \eval{-\frac{1}{4b^4}+\frac{1}{4n^4}} =\answer{\frac{1}{4n^4}}. 
\]

Since we want to approximate $\sum_{k=1}^{\infty} \frac{1}{k^5}$ by using $\sum_{k=1}^{6} \frac{1}{k^5}$, we should use $n=\answer{6}$.

\end{hint}
\end{exercise}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{exercise}
If we use the integral test estimate what is the smallest value of $N$ that should be chosen so $\sum_{k=1}^{N} \frac{1}{k^5}$ is within $.001$ of the value of $\sum_{k=1}^{\infty} \frac{1}{k^5}$?

\[
N=\answer{8}
\]

\begin{hint}
We want $N$ so $\int_N^{\infty} \frac{1}{x^5} \d x \leq .001$.  

\[
\int_N^{\infty} \frac{1}{x^5} \d x = \lim_{b \to \infty} \eval{\answer{-\frac{1}{4x^4}}}_N^{b} = \lim_{b \to \infty} \eval{-\frac{1}{4b^4}+\frac{1}{4N^4}}= \answer{\frac{1}{4N^4}}.
\]


Thus, we set $\answer{\frac{1}{4N^4}} \leq .001$ and find that the smallest integer $N$ that satisfies this inequality is $N \geq 7.95$, and we should thus choose $N = \answer{8}$.

\end{hint}
\end{exercise}

\end{exercise}
\end{document}


