\documentclass{ximera}

\title{Practice Exam 2}

\begin{document}
\begin{abstract}
This is your practice exam.
\end{abstract}
\maketitle
\thispagestyle{empty}

How to take the exam:
\begin{enumerate}
  \item Set aside $55$--$80$ minutes of time.
  \item Print out the PDF of the exam (or work on a separate sheet of paper).
  \item Take the exam for $55$--$80$ minutes. If it is easy and you have
    solved all the problems, you may finish early. If you are stuck,
    try to fight-it-out for the full time.
  \item Check your answers online with Ximera. The Ximera version of
    the practice exam will be posted online $24$ hours after the PDF is
    posted.
\end{enumerate}


When you come to your exam you will need to know:
\begin{enumerate}
\item Your name.
\item Your lastname.number.
\item Your instructor's name.
\item Your recitation leader's name.
\item Your recitation time.
\end{enumerate}
One point will be taken off for each missing piece of information.

The real exam will have the following written on it:

\begin{enumerate}
%% \item Keep in mind that if you are feeling anxious now
%%     or while taking this exam, this anxiety could be the result of
%%     negative stereotypes that are widely known in society and have
%%     nothing to do with your actual ability to do well on the exam.
\item Do not look inside the exam until instructed to do so.
\item Write your name above.
\item You have $55$ minutes to complete the exam.
\item Calculators or cell phones are prohibited.
\item Justify your answers for full credit.
\item Show your work for partial credit.
\item Write your answers on the included pages.
\item Provide answers for all questions asked.
\end{enumerate}

The real exam is \textbf{not} limited to question from the practice
exam. \textbf{Any topic from the course is fair-game on an in-class
  examination}.  However, at least 50\% of the real exam will be very
similar to the practice exam.



\end{document}
