\documentclass{ximera}

\input{../preamble.tex}

\title[Dig-In:]{The mean value theorem}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle


Here are some interesting questions involving derivatives:

\begin{enumerate}
\item Suppose you toss a ball into the air and then catch it. Must the
  ball's vertical velocity have been zero at some point?
\item Suppose you drive a car from toll booth on a toll road to
  another toll booth $30$ miles away in half of an hour. Must you have
  been driving at $60$ miles per hour at some point?
\item Suppose two different functions have the same derivative. What
  can you say about the relationship between the two functions?
\end{enumerate}

While these problems sound very different, it turns out that the
problems are very closely related. We'll start simply:

\begin{theorem}[Rolle's Theorem]\index{Rolle's Theorem} 
Suppose that $f(x)$ is differentiable on the interval $(a,b)$, is
continuous on the interval $[a,b]$, and $f(a)=f(b)$. Then 
\[
f'(c)=0
\]
for some $a<c<b$.
\label{thm:rolle}
\end{theorem}
%\begin{marginfigure}[0in]
\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            xmin=0, xmax=4.5,ymin=1,ymax=5,
            axis lines =left, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={1,2,3}, xticklabels={$a$,$c$,$b$},
            ytickmin=1, ytickmax=0,
            axis on top,
          ]       
          \addplot [draw=none, fill=fill2,domain=(1:3)] {5} \closedcycle;       
	  \addplot [very thick,penColor, smooth] {-(x-2)^2+4};
          \addplot [very thick,penColor2, smooth] {4};
          \node at (axis cs:.4,2.5) [penColor] {$f(x)$}; 
          \addplot [textColor,dashed] plot coordinates {(2,0) (2,4)};
          \addplot [textColor,dashed] plot coordinates {(1,3) (3,3)};
          \addplot[color=penColor3,fill=penColor3,only marks,mark=*] coordinates{(2,4)};  %% closed hole          
          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(1,3)};  %% closed hole          
          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(3,3)};  %% closed hole          
        \end{axis}
\end{tikzpicture}
%% \caption{A geometric interpretation of Rolle's Theorem.}
%% \label{figure:geoRolle}
%% \end{marginfigure}
\end{image}

\begin{proof}
By the Extreme Value Theorem, Theorem~\ref{theorem:evt}, we know that
$f(x)$ has a maximum and minimum value on $[a,b]$.

If maximum and minimum both occur at the endpoints, then
$f(x)=f(a)=f(b)$ at every point in $[a,b]$. Hence the function is a
horizontal line, and it has derivative zero everywhere on
$(a,b)$. We may choose any $c$ at all to get $f'(c)=0$.

If the maximum or minimum occurs at a point $c$ with $a<c<b$, then by
Fermat's Theorem, Theorem~\ref{theorem:fermat}, $f'(c)=0$.
\end{proof}

We can now answer our first question above.

\begin{example}
Suppose you toss a ball into the air and then catch it. Must the
ball's vertical velocity have been zero at some point?


If $p(t)$ is the position of the ball at time $t$, then we may apply
Rolle's Theorem to see at some time $c$, $p'(c)=0$. Hence the velocity
must be zero at some point.
\end{example}

Rolle's Theorem is a special case of a more general theorem.

\begin{theorem}[Mean Value Theorem]\label{thm:mvt}\index{Mean Value Theorem}
Suppose that $f(x)$ has a derivative on the interval $(a,b)$ and is
continuous on the interval $[a,b]$.  Then
\[
f'(c)=\frac{f(b)-f(a)}{b-a}
\]
for some $a<c<b$. 
\end{theorem}
%\begin{marginfigure}[.5in]
\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            xmin=.5, xmax=5.5,ymin=0,ymax=3.1,
            axis lines =center, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={1,2.04,5}, xticklabels={$a$,$c$,$b$},
            ytickmin=1, ytickmax=0,
            axis on top,
          ] 
          \addplot [draw=none, fill=fill2,domain=(1:5)] {3.1} \closedcycle;       
          \addplot [penColor2!40!background,very thick,dashed] plot coordinates {(1,.84+1.5) (5,1.5-.96)};        
          \addplot [textColor,dashed] plot coordinates {(2.04,0) (2.04,1.5+.89)};        
	  \addplot [very thick,penColor, smooth,domain=(1:5)] {sin(deg(x))+1.5};
          \addplot [very thick,penColor2,domain=(.5:5.5)] {-.45*(x-2.04)+.89+1.5};
          %\node at (axis cs:.4,2.5) [penColor] {$f(x)$}; 
          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(1,.84+1.5)};  %% closed hole          
          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(5,-.96+1.5)};  %% closed hole          
          \addplot[color=penColor3,fill=penColor3,only marks,mark=*] coordinates{(2.04,.89+1.5)};  %% closed hole          
        \end{axis}
\end{tikzpicture}
%% \caption{A geometric interpretation of the Mean Value Theorem}
%% \label{figure:geoMVT}
%% \end{marginfigure}
\end{image}
\begin{proof}
Let 
\[
m=\frac{f(b)-f(a)}{b-a},
\] 
and consider a new function $g(x)=f(x) - m(x-a)-f(a)$.  We know that
$g(x)$ has a derivative on $[a,b]$, since $g'(x)=f'(x)-m$. We can
compute $g(a)=f(a)- m(a-a)-f(a) =0$ and
\begin{align*}
g(b)=f(b)-m(b-a)-f(a)&=f(b)-{f(b)-f(a)\over b-a}(b-a)-f(a) \\
&=f(b)-(f(b)-f(a))-f(a)\\
&=0. 
\end{align*}
So $g(a) = g(b) = 0$. Now by Rolle's Theorem, that at some $c$,
\[
g'(c)=0\qquad\text{for some $a<c<b$}.
\]
But we know that $g'(c)=f'(c)-m$, so
\[
0=f'(c)-m=f'(c)-\frac{f(b)-f(a)}{b-a}.
\]
Hence
\[
f'(c)=\frac{f(b)-f(a)}{b-a}.
\]
\end{proof}

We can now answer our second question above.  

\begin{example}
Suppose you drive a car from toll booth on a toll road to another toll
booth $30$ miles away in half of an hour. Must you have been driving
at $60$ miles per hour at some point?


If $p(t)$ is the position of the car at time $t$, and $0$ hours is
the starting time with $1/2$ hours being the final time, the Mean Value Theorem states there is a time $c$
\[
p'(c) = \frac{30-0}{1/2} = 60\qquad \text{where $0<c<1/2$.}
\]
Since the derivative of position is velocity, this says that the car
must have been driving at $60$ miles per hour at some point.
\end{example}

Now we will address the unthinkable, could there be a function $f(x)$
whose derivative is zero on an interval that is not constant? As we will
see, the answer is ``no.''

\begin{theorem} 
If $f'(x)=0$ for all $x$ in an interval $I$, then $f(x)$ is constant
on $I$.
\end{theorem}

\begin{proof}
Let $a< b$ be two points in $I$. By the Mean Value Theorem we know
\[
\frac{f(b)-f(a)}{b-a} = f'(c)
\]
for some $c$ in the interval $(a,b)$. Since $f'(c)=0$ we see that
$f(b)=f(a)$. Moreover, since $a$ and $b$ were arbitrarily chosen,
$f(x)$ must be the constant function.
\end{proof}

Now let's answer our third question.

\begin{example}
Suppose two different functions have the same derivative. What can you
say about the relationship between the two functions?



Set $h(x) = f(x) - g(x)$, so $h'(x) = f'(x) -g'(x)$. Now $h'(x) = 0$
on the interval $(a,b)$. This means that $h(x) = k$ where $k$ is some
constant. Hence
\[
g(x) = f(x) + k.
\]
\end{example}


\begin{example}
Describe all functions whose derivative is $\sin(x)$.



One such function is $-\cos(x)$, so all such functions have the form
$-\cos(x)+k$, see Figure~\ref{figure:cos+k}.
\end{example}
%% \begin{marginfigure}[0in]
\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            xmin=0, xmax=6.2,ymin=-4,ymax=4,domain=(0:6.2),
            axis lines =center, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top,
          ] 
          \addplot [very thick,penColor, smooth] {-cos(deg(x))};
          \addplot [very thick,penColor2!30!background, smooth] {-cos(deg(x))+1};
          \addplot [very thick,penColor3!30!background, smooth] {-cos(deg(x))-1};
          \addplot [very thick,penColor4!30!background, smooth] {-cos(deg(x))+2};
          \addplot [very thick,penColor5!30!background, smooth] {-cos(deg(x))-2};         
        \end{axis}
\end{tikzpicture}
%% \caption{Functions of the form $-\cos(x)+k$, each of whose derivative is $\sin(x)$.}
%% \label{figure:cos+k}
%% \end{marginfigure}
\end{image}



\end{document}
