\documentclass{ximera}

\input{../preamble.tex}

\outcome{Second derivatives}


\title[Break-Ground:]{Rates of rates}

\begin{document}
\begin{abstract}
Two young mathematicians discuss what happens when they differentiate
multiple times.
\end{abstract}
\maketitle

Check out this dialogue between two calculus students (based on a true
story):

\begin{dialogue}
  \item[Devyn] Riley, I have a pressing question: What does the
    derivative of derivative mean?
  \item[Riley] Hmmm, well if we imagine that a function gives the
    position of an object with respect to time,
    \[
    s(t) = \text{position with respect to time}
    \]
    then the derivative is the change in position, so that should be
    velocity. Meaning
    \[
    s'(t) = \text{velocity with respect to time}.
    \]
  \item[Devyn] Ah! Right! A derivative is something like a
    ``speedometer'' for the growth of a function.
  \item[Riley] Ooooh! So if we want to know the derivative of
    velocity, we want to know the change in velocity with respect to
    time, and that is acceleration! So we have
    \[
    s''(t) = \text{acceleration with respect to time}.
    \]
  \item[Devyn] Hmmm what does $s'''(t)$ mean?
    \item[Riley] Good question! I wonder if we can recover $s(t)$ if
      we know $s'(t)$, and/or $s''(t)$?
\end{dialogue}

\begin{problem}
  Suppose that $s$ represents the position of a ball tossed at time
  $t=0$. Recalling that the acceleration for $t>0$ is only due to
  gravity, and knowing that the acceleration due to gravity is
  $-9.8~\mathrm{m}/\mathrm{s}^2$, what is $s''(t)$?
  \begin{prompt}
  \[
  s''(t) = \answer{-9.8}
  \]
  \end{prompt}
\end{problem}


\input{../leveledQuestions.tex}


\end{document}
