\documentclass{ximera}

\input{../../preamble.tex}

\author{Jim Talamo}
\license{Creative Commons 3.0 By-bC}


\outcome{}


\begin{document}
\begin{exercise}

Consider the series $\sum_{k=1}^{\infty} \frac{k^2}{2^k}$.  

\begin{multipleChoice}
\choice[correct]{This series converges.}
\choice{This series diverges.}

Which of the following tests would allow us to conclude that the series converges?
\end{multipleChoice}
\begin{selectAll}
\choice{Divergence Test}
\choice[correct]{Ratio Test}
\choice[correct]{Root Test}
\choice{This is a geometric series with $|r|<1$, so it converges.}
\choice{This is a $p$-series with $p>1$, so it converges.}
\end{selectAll}

\begin{hint}
You should be able to provide the details for all of the tests that can be used above!  Note that since the Comparison Test requires us to be careful with our estimates, it is usually better to see if a solution can be provided with either Ratio, Root, or the Limit Comparison Test.  For the Limit Comparison test, what would be a good choice to use for a comparison?  Generally, when there are exponential terms, it's better to try to use the Ratio or Root test instead!

\begin{question}
By the growth rates results:

\[
\lim_{n \to \infty} \frac{n^2}{2^n} = \answer{0}
\]
What does the divergence test tell us?

\end{question}

\begin{question}
For the Ratio Test, the limit we must compute is:

\[
L = \lim_{n \to \infty} \frac{a_{n+1}}{a_n} =\lim_{n \to \infty}  \frac{\answer{\frac{(n+1)^2}{2^{n+1}}}}{\answer{\frac{n^2}{2^n}}} = \lim_{n \to \infty}  \frac{(n+1)^2}{n^2} \cdot \frac{2^n}{2^{n+1}} = \answer{\frac{1}{2}}
\]

Thus:

\begin{multipleChoice}
\choice[correct]{The series converges by the Ratio Test.}
\choice{The series diverges by the Ratio Test.}
\choice{The Ratio Test is inconclusive.}
\end{multipleChoice}
\end{question}

\begin{question}
For Root Test, the limit we must compute is:

\[
L = \lim_{n \to \infty} \sqrt[n]{a_n} =\lim_{n \to \infty} \sqrt[n]{\frac{n^2}{2^n}} = \lim_{n \to \infty} \frac{\answer{\sqrt[n]{n^2}}}{\sqrt[n]{2^n}}
\]

We have seen previously that if $p(x)$ is a polynomial, then $\lim_{n \to \infty} \sqrt[n]{p(n)} = \answer{1}$.  Also: $\sqrt[n]{2^n} = \answer{2}$, so the above limit is $\answer{\frac{1}{2}}$.

Thus:
\begin{multipleChoice}
\choice[correct]{The series converges by the Root Test.}
\choice{The series diverges by the Root Test.}
\choice{The Root Test is inconclusive.}
\end{multipleChoice}

\end{question}


\end{hint}

\end{exercise}

\end{document}
