\documentclass{ximera}

\input{../../preamble.tex}

\outcome{Understand what is meant by the form of a limit.}
\outcome{Calculate limits of the form zero over zero.}
\outcome{Identify determinate and indeterminate forms.}
\outcome{Distinguish between determinate and indeterminate forms.}

\author{Nela Lakos}

\begin{document}


\begin{exercise}
Let $g(x) =\cos{(x-4)}$, and $h(x) = (x-4)^{2}$.\\\\

\begin{enumerate}
\item Evaluate the limit.
\[
\lim_{x\to4}g(x)  =\answer{1}
\]

\item Choose all correct statements regarding  the form of the limit. 
\[
\lim_{x\to4}\frac{\cos{(x-4)}}{(x-4)^{2}}
\]
Choose all correct statements.
\begin{selectAll} 
\choice[correct]{The limit is of determinate form.}
\choice{The limit is of indeterminate form.}
\choice{The limit is of the form $\dfrac{0}{0}$.}
\choice[correct]{The limit is of the form $\dfrac{\#}{0}$.}
\end{selectAll}
\noindent\rule[0.5ex]{\linewidth}{0.2pt}
\end{enumerate}
\begin{exercise}
 Evaluate the limit. Possible answers include a number, $+\infty$, $-\infty$ and $DNE$.
\[
\lim_{x\to4}\frac{\cos{(x-4)}}{(x-4)^{2}}=\answer{+\infty}
\]
 Justify your answer  by choosing all correct statements.
 \begin{selectAll} 
\choice{The numerator is negative and the denominator is positive and approaching zero.}
\choice[correct]{The numerator is positive and the denominator is positive and approaching zero.}
\choice{The numerator is positive and the denominator is negative and approaching zero.}
\choice{The numerator is negative and the denominator is negative and approaching zero.}
\end{selectAll}
\noindent\rule[0.5ex]{\linewidth}{0.2pt}


\end{exercise}
\end{exercise}
\end{document}