\documentclass{ximera}

\input{../preamble.tex}

\outcome{State the Second Fundamental Theorem of Calculus.}
\outcome{Evaluate definite integrals using the Second Fundamental Theorem of Calculus.}
\outcome{Understand how the area under a curve is related to the antiderivative.}
\outcome{Understand the relationship between indefinite and definite integrals.}

\title[Dig-In:]{Second Fundamental Theorem of Calculus}

\begin{document}
\begin{abstract}
Accumulated area is computed by undoing the derivative. 
\end{abstract}
\maketitle




There is a another common form of the Fundamental Theorem of Calculus:

\begin{theorem}[Second Fundamental Theorem of Calculus]
  Let $f$ be continuous on $[a,b]$. If $F$ is \textbf{any}
  antiderivative of $f$, then
  \[
  \int_a^b f(x)\d x = F(b)-F(a).
  \]
  \begin{explanation}
    Let $a\le c\le b$ and write
    \begin{align*}
      \int_a^b f(x) \d x &= \int_a^c f(x) \d x + \int_c^b f(x) \d x \\
      &= \int_c^b f(x) \d x - \int_c^a f(x) \d x.
    \end{align*}
    By the First Fundamental Theorem of Calculus, we have
    \[
    F(b) = \int_c^b f(x) \d x\qquad\text{and}\qquad F(a) = \int_c^a f(x) \d x
    \] 
    for some antiderivative $F$ of $f$. So
    \[
    \int_a^b f(x) \d x = F(b)-F(a)
    \]
    for this antiderivative. However, \textbf{any} antiderivative
    could have be chosen, as antiderivatives of a given function
    differ only by a constant $k$, and this constant \textit{always}
    cancels out of the expression when evaluating $F(b)-F(a)$.
\end{explanation}
\end{theorem}

From this you should see that the two versions of the Fundamental
Theorem are very closely related. In reality, the two forms are
\textbf{equivalent}, just differently stated. Hence people often
simply call them both ``The Fundamental Theorem of Calculus.''
One way of thinking about the Second Fundamental Theorem of Calculus is:
\begin{center}
  \begin{tikzpicture}[scale=2,every node/.style={transform shape}]
    \node at (0,0) {
      $\color{green!70!black!70!blue}\int_a^b\color{blue!70!green}f'(t)\color{green!70!black!70!blue}\d t\color{black} = 
      \color{purple!50!blue!90!black}f(b) - f(a)$
      };
  \end{tikzpicture}
\end{center}
This could be read as:%%BADBAD I want this all in bold and with colors
\begin{quote}\large\textbf{The \textcolor{green!70!black!70!blue}{accumulation} of a \textcolor{blue!70!green}{rate} is given by the \textcolor{purple!50!blue!90!black}{change in the amount}.}
\end{quote}


When we compute a definite integral, we first find an antiderivative
and then evaluate at the limits of integration. It is convenient to
first display the antiderivative and then evaluate.  A special
notation is often used in the process of evaluating definite integrals
using the Fundamental Theorem of Calculus. Instead of explicitly
writing $F(b)-F(a)$, we often write
\[
\eval{F(x)}_a^b
\]
meaning that one should evaluate $F(x)$ at $b$ and then subtract
$F(x)$ evaluated at $a$
\[
\eval{F(x)}_a^b = F(b)-F(a).
\]

Let's see some examples of the fundamental theorem in action.

\begin{example}
  Compute:
  \[
  \int_{-2}^2 x^3\d x
  \]
  \begin{explanation}
    We start by finding an antiderivative of
    \[
    x^3.
    \]
    A correct choice is $\frac{x^{4}}{4}$, one could verify
    this by taking the derivative. Hence
    \begin{align*}
      \int_{-2}^2 x^3 \d x &=\eval{\frac{x^4}{4}}_{-2}^2\\
      &= \frac{2^4}{4} - \frac{(-2)^4}{4}\\
      &= 0.
    \end{align*}
  \end{explanation}
\end{example}


\begin{example}
  Compute:
  \[
  \int_0^\pi \sin \theta \d \theta
  \]
  \begin{explanation}
    We start by finding an antiderivative of
    \[
    \sin \theta.
    \]
    A correct choice is $-\cos\theta$, one could verify
    this by taking the derivative. Hence
    \begin{align*}
      \int_0^\pi \sin\theta \d \theta &= \eval{-\cos\theta}_0^\pi\\
      &=\answer[given]{-\cos(\pi)} - (-\cos(0))\\
      &=\answer[given]{2}.
    \end{align*}
    This is interesting: It says that the area under one ``hump'' of a
    sine curve is $2$.
  \end{explanation}
\end{example}

\begin{example}
  Compute:
  \[
  \int_0^5 e^t\d t
  \]
  \begin{explanation}
    We start by finding an antiderivative of
    \[
    e^t.
    \]
    A correct choice is $\answer[given]{e^t}$, one could verify this by taking the
    derivative. Hence
    \begin{align*}
      \int_0^5 e^t \d t &= \eval{e^t}_0^5 \\
      &= e^5-\answer[given]{e^0}\\
      &= e^5-\answer[given]{1}.
    \end{align*}
  \end{explanation}
\end{example}


\begin{example}
Compute:
\[
\int_1^2\left(x^9 + \frac{1}{x}\right) \d x
\]
\begin{explanation}
We start by finding an antiderivative of
\[
x^9 + \frac{1}{x}.
\]
A correct choice is $\frac{x^{10}}{10} + \ln(x)$, one could verify
this by taking the derivative. Hence
\begin{align*}
\int_1^2\left(x^9 + \frac{1}{x}\right) \d x &= \eval{\frac{x^{10}}{10} + \ln(x)}_1^2 \\
&= \frac{2^{10}}{10} + \ln(2) - \answer[given]{\frac{1}{10}}.
\end{align*}
\end{explanation}
\end{example}




\section{Understanding motion with the Fundamental Theorem of Calculus}

We know that
\begin{itemize}
\item The derivative of a position function is a velocity function.
\item The derivative of a velocity function is an acceleration
  function.
\end{itemize}
Now consider definite integrals of velocity and acceleration
functions. Specifically, if $v(t)$ is a velocity function, what does
$\int_a^b v(t)\d t$ mean?

The Second Fundamental Theorem of Calculus states that
\[
\int_a^b v(t)\d t = V(b) - V(a),
\]
where $V(t)$ is any antiderivative of $v(t)$. Since $v(t)$ is a
velocity function, $V(t)$ must be a position function, and $V(b) -
V(a)$ measures a \textbf{change in position}, or \dfn{displacement}.

\begin{example}
  A ball is thrown straight up with velocity given by $v(t) =
  -32t+20$ft/s, where $t$ is measured in seconds. Find, and interpret,
  $\int_0^1 v(t)\d t$.
    \begin{explanation}
      Using the Second Fundamental Theorem of Calculus, we have
      \begin{align*}
        \int_0^1 v(t)\d t &= \int_0^1 (-32t+20)\d t \\
	&= \eval{\answer[given]{-16t^2 + 20t}}_0^1 \\
	&= 4
      \end{align*}
      Thus if a ball is thrown straight up into the air with velocity
      \[
      v(t) = \answer[given]{-32t+20},
      \]
      the height of the ball, $1$ second later, will be $4$ feet above the
      initial height. Note that the ball has \textit{traveled} much
      farther. It has gone up to its peak and is falling down, but the
      difference between its height at $t=0$ and $t=1$ is $4$ft. 
    \end{explanation}
\end{example}    



\section{A tale of three integrals}

At this point we have three different ``integrals.'' Let's see if we
can sort out the differences.

\subsection{Indefinite integrals}

Indefinite integrals, also called \dfn{antiderivatives} compute
classes of functions:
\[
\int f(x) \d x = \text{``a class of functions whose derivative is $f$''}
\]
Here there are no limits of integration, and your answer will have a
``$+C$'' at the end.

\subsection{Accumulation functions}

Accumulation functions, also called \dfn{area functions} compute accumuated area:
\[
\int_a^x f(t) \d t = \text{``a function $F$ whose derivative is $f$''}
\]
This is a function of $x$ whose derivative is $f$, with the additional
property that $F(a)=0$.


\subsection{Definite integrals}

Definite integrals compute signed area:
\[
\int_a^b f(x) \d x = \text{``the signed area between the $x$-axis and $f$''}
\]
Here we always have limits of integration, both of which are
numbers. Moreover, definite integrals have definite values, the signed
area between $f$ and the $x$-axis


Now we know that to solve certain kinds of problems, those that
involve accumulation of some form, we ``merely'' find an
antiderivative and substitute two values and subtract. Unfortunately,
finding antiderivatives can be quite difficult. While there are a
small number of rules that allow us to compute the derivative of any
common function, there are no such rules for antiderivatives. There
are some techniques that frequently prove useful, but we will never be
able to reduce the problem to a completely mechanical process.







\end{document}
