\documentclass{ximera}

\input{../preamble.tex}

\outcome{Compute basic antiderivatives.}
\outcome{Solve basic initial value problems.}
\title[Break-Ground:]{Jeopardy of calculus}

\begin{document}
\begin{abstract}
  Two young mathematicians discuss a `\textit{Jeopardy}' version of calculus.
\end{abstract}
\maketitle

Check out this dialogue between two calculus students:

% BADBAD: Jeopardy of Calculus

\begin{dialogue}
\item[Devyn] (Pretending to Alex Trebek) I've got a new costume.
\item[Riley] Woah! You look just like Trebek!
\item[Devyn] So you know how in \textit{Jeopardy} Trebek, uh I mean, \textit{I} give you an answer, and you must tell me the questions?
\item[Riley] Uh Alex, `What are the rules of \textit{Jeopardy}?'
\item[Devyn] Ha. Exactly! So I'll tell you a derivative, and you tell
  me the function? Are we ready to play ``Calculuperty?''
\item[Riley] I'll take ``Formulas for slope'' for $600$.
\item[Devyn] $3\cdot e^{3x}$  
\item[Riley] I've got an answer!  actually, I've got three different answers!
  \[
  e^{3x},\qquad e^{3x}+1,\qquad e^{3x}-1.
  \]
\item[Devyn] Hmmm. Now I'm not sure which one it was.
\item[Riley] What about if you had given me $\frac{\sin(x)}{x}$?
\end{dialogue}


\begin{problem}
  How many functions whose derivative is $3\cdot e^{3x}$ are there?
  \begin{multipleChoice}
    \choice{Zero}
    \choice{One}
    \choice{Two}
    \choice{Three}
    \choice{Four}
    \choice[correct]{Infinitely many}
  \end{multipleChoice}
\end{problem}

\begin{problem}
  How many functions whose derivative is $3\cdot e^{3x}$ that equal
  $1$ at $x=0$ are there?
  \begin{multipleChoice}
    \choice{Zero}
    \choice{One}
    \choice{Two}
    \choice{Three}
    \choice{Four}
    \choice[correct]{Infinitely many}
  \end{multipleChoice}
\end{problem}


\input{../leveledQuestions.tex}

\end{document}
