\documentclass{ximera}

\input{../preamble.tex}

\outcome{}

\title[Dig-In:]{Anti-Derivatives}

\begin{document}
	\begin{definition}
		We say that $F$ is an \textbf{antiderivative} of $f$ if the derivative of $F$ is $f$.
	\end{definition}
	
%	\begin{question}
%		BADBAD 
%		Want a question here where the student is presented with $f$ and has to sketch an antiderivative.
%	\end{question}
	
	\begin{question}
		An antiderivative of $f(x) = x^2$ is $F(x) = \antiderivativeanswer{x^2}$
			\begin{hint}
				You have to think backwards.  What function can you think of that, when differentiated, has an $x^2$ term in it?  Can you modify that function to give you the antiderivative you seek?
			\end{hint}
			\begin{hint}
				$\frac{d}{dx} x^3 = 3x^2$, so this is close.  What could one do to $x^3$ to get rid of that factor of $3$?
			\end{hint}
			\begin{hint}
				$\frac{d}{dx} \frac{1}{3} x^3 = x^2$, so $F(x) = \frac{1}{3} x^3$ is an antiderivative.  Note that there are other antiderivatives, like $F(x) = \frac{1}{3}x^3 + 7$, for instance.
			\end{hint}
	\end{question}
	
	\begin{question}
		%I do not think we have the ability to access multiple answers yet, even in a custom solution environment.
		
		What are two different antiderivatives of $f(x) = 0$?  $F_1(x) =  \antiderivativeanswer{0}$ and $F_2(x) = \antiderivativeanswer{0}$.
		
		\begin{hint}
			The derivative of any constant function is $0$, so you can just use two different constant functions.
		\end{hint}
	\end{question}
	
	\begin{question}
		Use the mean value theorem for derivatives to show that if $F$ is an antiderivative of $f(x) = 0$, then $F$ must be a constant function.  This should make intuitive sense, since basically the statement says "If the slope is always 0, then the graph is a horizontal line", but here you should provide a rigorous justification using the mean value theorem.
		
		\begin{hint}
			Showing that $F$ is constant is the same as showing that $F(a) = F(b)$ for every $a,b \in \mathbb{R}$.
		\end{hint}
		\begin{hint}
			If you assume to the contrary that $F(a) \neq F(b)$ for some $a,b \in \mathbb{R}$, then the slope of the secant line connecting them will be nonzero.  Can you use the mean value theorem for derivatives to get a contradiction?
 		\end{hint}
		
		\begin{free-response}
			Showing that $F$ is constant is the same as showing that $F(a) = F(b)$ for every $a,b \in \mathbb{R}$.
			
			Assume to the contrary that $F(a) \neq F(b)$ for some $a,b \in \mathbb{R}$.
			
			Then the slope of the secant line connecting $(a,F(a))$ to $(b,F(b))$ is $\frac{F(b)-F(a)}{b-a}$ which is nonzero.
			
			By the mean value theorem for derivatives, there is a point $c$ between $a$ and $b$ with $F'(c) = \frac{F(b) - F(a)}{b-a}$.
			
			But $F'(c) = f(c)$ since $F$ is an antiderivative of $f$, and $f(c) = 0$.
			
			Thus assuming $F$ was not constant has lead to a contradiction.
			
			Thus $F$ must be a constant function. 
		\end{free-response}
	\end{question}
	
	\begin{question}
		Using the previous exercise, show that if $F_1$ and $F_2$ are both antiderivatives of $f$ on an interval $(a,b)$, then $F_1 - F_2$ is constant on that interval. 
		\begin{free-response}
			Since $F_1$ and $F_2$ are both antiderivatives of $f$, we have
			
			$\frac{d}{dx}  \left[ F_1(x) - F_2(x)\right]  =f(x) - f(x) = 0$
			
			This $F_1 - F_2$ is an antiderivative of $0$, which by the previous exercise must be a constant function.
		\end{free-response}
	\end{question}
	
	It is worth stating the contents of the last question as a theorem:
	
	\begin{theorem}
		Let $F_1$ and $F_2 $ be antiderivatives of $f$ on an interval $(a,b)$.  Then $F_1(x) = F_2(x)+C$ for some constant $C$.
	\end{theorem}
	
	\begin{question}
		Let $f(x)$ be defined on $[-3,6]$.  Let $F(x)$ and $G(x)$ be antiderivatives of $f$.  Mark all that necessarily  are true.
		\begin{multiple-response}
			\choice[correct]{ F(1) - G(1) = F(2)  - G(2)}
			\choice[correct]{F(3) - F(0) = G(3) - G(0)}
			\choice{F(1) - G(0) = G(1) - F(0)}
			\choice{F(1) - F(0) = F(2) - F(1)} 
		\end{multiple-response}
	\end{question}
	
	\begin{question}
		$\frac{d}{dx} $
	\end{question}
	
	\begin{question}
		Let $f(x)$ be defined on $(-7,0) \cup (3,9)$. Let  $F$ and $G$ be antiderivatives of $f$.  Mark all that are necessarily true:
		\begin{multiple-response}
			\choice[correct]{F(-2) - F(-5) = G(-2) - G(-5)}
			\choice{F(5) - F(-2) = G(5) - G(-2) }
			\chocie[correct]{F(7) - G(7) = F(5) - G(5) }
			\choice{There is a constant $C$ such that $F(x) = G(x)+C$}
		\end{multiple-response}
	\end{question}
	
	As the last exercise shows, if $f$ is defined on a disjoint set of intervals, its antiderivatives can differ by a different constant on each piece.  
	For example, in the following picture $F$ and $G$ are both antiderivatives of $f$, but they have a different constant on each interval.
		
	\begin{theorem}
		Let $f(x)$ be defined on $(a_1,b_1) \cup (a_2,b_2) \cup ... \cup (a_n,b_n)$ with all the intervals $(a_i,b_i)$ disjoint.  If $F$ and $G$ are antiderivatives of $f$, 
		then there are constants $C_1,C_2,...,C_n$ such that
		
		$G(x) = \begin{cases}
			F(x)+C_1 \text{on $(a_1,b_1)$}\\
			F(x)+C_2 \text{on $(a_2,b_2)$}\\
			\vdots\\
			F(x)+C_n \text{on $(a_n,b_n)$}\\
		\end{cases}$
	\end{theorem}
	
	\begin{example}
		We know that $\frac{d}{dx} \log( |x |) =  \frac{1}{x}$ on its domain.   Since its domain consists of the two disjoint intervals $(-\infty,0) \cup (0,\infty)$, we have that any antiderivative of $\frac{1}{x}$ can be written 
		\[
		F(x) = \begin{cases}
			\log(x)+C_1 if $x >0$\\
			\log(-x)+C_2 if $x<0$
		\end{cases}
		\]
	\end{example}
	
	\begin{question}
		Let $a$ be a positive number.  Show that $\log(ax)$ and $\log(x)$ are both antiderivatives of $\frac{1}{x}$ on $(0,\infty)$.  Conclude that $\log(ax)$ and $\log(x)$ must differ by a constant.  What constant do they differ by?  Is your discovery already familiar to you?
		
		\begin{free-response}
			\begin{align*}
				\frac{d}{dx} \log(ax) &= a\frac{1}{ax}\\
				&=\frac{1}{x}
				&=\frac{d}{dx} \log(x)
			\end{align*}
			
			Thus $\log(ax)$ and $\log(x)$are both antiderivatives of $\frac{1}{x}$ on $(0,\infty)$.
			
			Thus they must differ by a constant, so $\log(ax) = \log(x)+C$ for some constant $C$.
			
			Plugging in $x=1$, we have $\log(a) = \log(1)+C$, so $\log(a) = C$.
			
			Thus we can conclude that $\log(ax) = \log(x)+\log(a)$.
			
			This is one of the laws of logarithms!  It is kind of neat that we can derive this law of logarithms in this way.
			
		\end{free-response}
	\end{question}
	
	

\end{document}