\documentclass{ximera}

\input{../preamble.tex}

\outcome{Understand exponential growth and decay models.}
\outcome{Model a process using an exponential growth or exponential decay equations.}
\outcome{Find doubling-time or half-life.}

\title[Dig-In:]{Exponential models}

\begin{document}
\begin{abstract}
We investigate how the exponential functions model phenomena in the
real world.
\end{abstract}
\maketitle

A \textit{differential equation}\index{differential equation} is
simply an equation with a derivative in it. Here is an example:
\[
a\cdot f''(x) + b\cdot f'(x) + c\cdot f(x) = g(x). 
\]
\begin{question}
  What is a differential equation?
  \begin{multipleChoice}
    \choice{An equation that you take the derivative of.}
    \choice[correct]{An equation that relates the rate of a function to other values.}
    \choice{It is a formula for the slope of a tangent line at a given point.}  
  \end{multipleChoice}
\end{question}

When a mathematician solves a differential equation, they are finding
\textit{functions} satisfying the equation.
\begin{question}
  Which of the following functions solve the differential equation
  \[
  f''(x) = -f(x)?
  \]
  \begin{selectAll}
    \choice{$e^x$}
    \choice[correct]{$\sin(x)$}
    \choice[correct]{$\cos(x)$}
  \end{selectAll}
\end{question}






\section{Exponential growth and decay}

A function $f$ exhibits \textit{exponential growth}\index{exponential
  growth} if its growth rate is proportional to its value. As a
differential equation, this means
\[
f'(x) = k f(x)\qquad\text{for some constant of proportionality $k$.}
\]
We claim that this differential equation is solved by
\[
f(x) = Ae^{kx},
\]
where $A$ and $k$ are constants.  Check it out, if $f(x) = Ae^{kx}$,
then
\begin{align*}
f'(x) &= Ak e^{kx}\\
&= k\left(Ae^{kx} \right)\\
&= k f(x).
\end{align*}


\begin{example}
A culture of yeast starts with $100$ cells. After $160$ minutes, there
are $350$ cells. Assuming that the growth rate of the yeast is
proportional to the number of yeast cells present, estimate when the
culture will have $1000$ cells.

\begin{explanation}
Since the growth rate of the yeast is proportional to the number of
yeast cells present, we have the following differential equation
\[
p'(t) = k p(t)
\]
where $p(t)$ is the population of the yeast culture at time $t$ with
$t$ measured in minutes. We know that this differential equation is
solved by the function
\[
p(t) = A e^{k\cdot t}
\]
where $A$ and $k$ are yet to be determined constants. Since
\[
100 = p(0) = Ae^{k\cdot 0}
\]
we see that $A = 100$. So 
\[
p(t) = 100 e^{k\cdot t}.
\]
Now we must find $k$. Since we know that 
\[
\answer[given]{350} = p(160) = 100e^{k\cdot 160}
\]
we need to solve for $k$. Write
\begin{align*}
350 &= 100 e^{k\cdot 160}\\
3.5 &= e^{k\cdot 160}\\
\ln(3.5) &= k\cdot 160\\
\ln(3.5)/160 &= k. 
\end{align*}
Hence
\[
p(t) = 100 e^{t\ln(3.5)/160} = 100 \cdot 3.5^{t/160}.
\]
To find out when the culture has 1000 cells, write
\begin{align*}
1000 &= 100 \cdot 3.5^{t/160}\\
10 &= 3.5^{t/160}\\
\ln(10) &= \frac{t\ln(3.5)}{160}\\
\frac{160\ln(10)}{\ln(3.5)} &= t.
\end{align*}
From this we find that after approximately $294$ minutes, there are
around $1000$ yeast cells present.
\end{explanation}
\end{example}

One concept related to this sort of exponential models is that of \textit{doubling-time}.

\begin{definition}
  The \dfn{doubling-time} of an exponetial function
  \[
  f(t) = c\cdot e^{k\cdot t}
  \]
  where $c\ne 0$, $b>0$, and $k>0$ is the value $T_2 =t$ such that
  \[
  f(T_2+t) = 2 f(t)
  \]
  for all values of $t$.
\end{definition}

\begin{example}
  Find the doubling-time of the function $f(t) = c\cdot e^{k\cdot t}$.
  \begin{explanation}
    We want to find $T_2$ such that
    \[
    f(T_2+t) = 2 f(t).
    \]
    Write with me:
    \begin{align*}
    c\cdot e^{k\cdot (T_2+t)} &= 2 c\cdot e^{k\cdot t}\\
    e^{k\cdot T_2+k\cdot t} &= 2 e^{k\cdot t}\\
    k\cdot T_2+k\cdot t &= \ln (2) + k\cdot t\\
    T_2 &= \frac{\ln (2)}{k}
    \end{align*}
  \end{explanation}
\end{example}

What is somewhat remarkable is that for the exponential function
  \[
  f(t) = c\cdot e^{k\cdot t}
  \]
  where $c\ne 0$, $b>0$, and $k>0$ this doubling time only depends on
  $k$!




It is worth seeing an example of exponential decay as well. Consider
this: Living tissue contains two types of carbon, a stable isotope
carbon-12 and a radioactive (unstable) isotope carbon-14. While an
organism is alive, the ratio of one isotope of carbon to the other is
always constant. When the organism dies, the ratio changes as the
radioactive isotope decays.  This is the basis of radiocarbon dating.


\begin{example}
The half-life of carbon-14 (the time it takes for half of an amount of
carbon-14 to decay) is about 5730 years. Moreover, the rate of decay
of carbon-14 is proportional to the amount of carbon-14.

If we find a bone with $1/70$th of the amount of carbon-14 we would
expect to find in a living organism, approximately how old is the
bone?

\begin{explanation}
Since the rate of decay of carbon-14 is proportional to the amount of
carbon-14 present, we can model this situation with the differential
equation
\[
f'(t) = k f(t).
\]
We know that this differential equation is solved by the function
defined by
\[
f(t) = A e^{k\cdot t}
\]
where $A$ and $k$ are yet to be determined constants. Since the
half-life of carbon-14 is about $5730$ years we write
\[
\frac{1}{2} = e^{k 5730}.
\]
Solving this equation for $k$, gives
\[
k = \frac{-\ln(2)}{5730}.
\]
Since we currently have $1/70$th of the original amount of carbon-14
we write
\[
\answer[given]{\frac{1}{70}} = 1\cdot e^{\frac{-\ln(2)t}{5730}}.
\]
Solving this equation for $t$, we find $t \approx -35121$. This means
that the bone is approximately $35121$ years old.
\end{explanation}
\end{example}

In our example above we briefly defined \textit{half-life}. Let's do
this justice.

\begin{definition}
  The \dfn{half-life} of an exponetial function
  \[
  f(t) = c\cdot e^{-k\cdot t}
  \]
  where $c\ne 0$, $b>0$, and $k>0$ is the value $T_{\frac{1}{2}} =t$ such that
  \[
  f(T_{\frac{1}{2}}+t) =  \frac{f(t)}{2}
  \]
  for all values of $t$.
\end{definition}

\begin{example}
  Find the half-life of the function $f(t) = c\cdot e^{-k\cdot t}$.
  \begin{explanation}
    We want to find $T_{\frac{1}{2}}$ such that
    \[
    f(T_{\frac{1}{2}}+t) = \frac{f(t)}{2}.
    \]
    Write with me:
    \begin{align*}
    c\cdot e^{-k\cdot (T_{\frac{1}{2}}+t)} &= \frac{1}{2} c\cdot e^{-k\cdot t}\\
    e^{-k\cdot T_{\frac{1}{2}}-k\cdot t} &= \frac{1}{2} e^{-k\cdot t}\\
    -k\cdot T_{\frac{1}{2}}-k\cdot t &= \ln \left(\frac{1}{2}\right) - k\cdot t\\
    T_{\frac{1}{2}} &= \frac{\ln \left(\frac{1}{2}\right)}{-k}.
    \end{align*}
    Noting that $\ln(1/2) = -\ln(2)$ we see that
    \[
    T_{\frac{1}{2}} = \frac{\ln(2)}{k}.
    \]
  \end{explanation}
\end{example}

Here again it is somewhat remarkable is that for the exponential
function
  \[
  f(t) = c\cdot e^{-k\cdot t}
  \]
  where $c\ne 0$, $b>0$, and $k>0$ the half-life only depends on $k$!


  While exponetial models are somewhat limited, they apply in many
  cases. Here are some examples of real-world phenomena that can be
  modeled by exponential growth:
\begin{itemize}
  \item The rate that a population grows is proportional to the size
    of the population.
  \item The rate that a radioactive substance decays is proportional
    to the amount of the substance.
  \item The rate that an investment grows is proportional to the
    amount of the investment.
  \item The rate that a pathogen spreads is proportional to the number
    of people with the pathogen.
  \item The rate that glucose is processed in one's blood is
    proportional to the concentration of glucose in the blood.
\end{itemize}


  
  
\end{document}
