\documentclass{ximera}

\input{../../preamble.tex}

\author{Nela Lakos \and Bart Snapp}

\outcome{Find the domain and range of a function.}
\outcome{Perform basic operations and compositions on functions.}

\begin{document}
\begin{exercise}
  A box with an open top is to be constructed from a rectangular piece
  of cardboard with dimensions $12\unit{in}$ by $20\unit{in}$ by
  cutting out equal squares of side $x$ at each corner and then folding
  up the sides:
  \begin{image}
    \begin{tikzpicture}
      \draw [dashed] (0,0) rectangle (1,1);
      \draw [dashed] (6,0) rectangle (7,1);
      \draw [dashed] (6,4) rectangle (7,5);
      \draw [dashed] (0,4) rectangle (1,5);

      \draw [penColor, very thick, fill=fill3] (0,1) rectangle (1,4);
      \draw [penColor, very thick, fill=fill3] (1,0) rectangle (6,1);
      \draw [penColor, very thick, fill=fill3] (1,4) rectangle (6,5);
      \draw [penColor, very thick, fill=fill3] (6,1) rectangle (7,4);
      
      \draw [penColor,very thick,fill=fill1] (1,1) rectangle (6,4);

      \draw[decoration={brace,mirror,raise=.2cm},decorate,thin] (0,0)--(7,0);
      \draw[decoration={brace,raise=.2cm},decorate,thin] (0,0)--(0,5);
      
      \node [penColor] at (3.5,-.5) {$20$};
      \node [penColor] at (-.5,2.5) {$12$};

      \draw[decoration={brace,mirror,raise=.2cm},decorate,thin] (7,4)--(7,5);
      \draw[decoration={brace,raise=.2cm},decorate,thin] (6,5)--(7,5);
      
      \node [penColor] at (7.5,4.5) {$x$};
      \node [penColor] at (6.5,5.5) {$x$};
    \end{tikzpicture}
  \end{image}
  Express the volume $V$ of the box as a function of $x$.
  \[
  V(x) = \answer{x(20-2x)(12-2x)}
  \]
\end{exercise}
\begin{exercise}
  Simplify your answer above:
  \[
  V(x) = \answer{4}x^3 + \answer{-64}x^2 + \answer{240}x
  \]
\end{exercise}
\begin{exercise}
  What is the (natural) domain of $V$?
  \[
    \left[\answer{0},\answer{6}\right]
  \]
\end{exercise}
\begin{exercise}
  Is the function $V$ a polynomial?
  \begin{multipleChoice}
    \choice{yes}
    \choice[correct]{no}
  \end{multipleChoice}
  \begin{feedback}
    $V$ is not a polynomial. The domain of a polynomial is
    $(-\infty,\infty)$. $V$ is a polynomial on a restricted domain.
  \end{feedback}
\end{exercise}
\begin{exercise}
  Evaluate (either write a number or ``DNE''):
  \begin{align*}
    V(1) &= \answer{180}\\
    V(5) &= \answer{100}\\
    V(10) &= \answer{DNE}
  \end{align*}
\end{exercise}
\begin{exercise}
  If $x = 1$ and increases by $h$, by how much will volume of the box change? Simplify.
  \begin{hint}
    You may find the following formulas useful:
    \begin{align*}
      (a+b)^2 &= a^2 + 2ab + b^2\\
      (a + b)^3 &= a^3 + 3 a^2 b + 3 ab^2 + b^3
    \end{align*}
  \end{hint}
  \[
  V(1+h) - V(1) = \answer{4}h^3 +\answer{-52}h^2 + \answer{124}h
  \]
\end{exercise}

\begin{exercise}
  As $x$ increases from $x=1$ to $x=5$, what is $h$?
  \[
  h=\answer{4}
  \]
  Now, compute
  \begin{align*}
    V(5) &= \answer{100}\\
    V(1) &= \answer{180}\\
    V(5) - V(1) &= \answer{-80}
  \end{align*}
  \begin{exercise}
    We also know from your work above that
    \begin{align*}
      V(1+h) -V(1) &= 4h^3 -52h^2 + 124h\\
      V(1+4) - V(1) &= \answer{-80}
    \end{align*}
  \end{exercise}
\end{exercise}

\begin{exercise}
  Express the total surface area (inside and outside) of the open box
  as a function of $x$.
  \[
  S(x) = \answer{480-8x^2}
  \]
\end{exercise}
\begin{exercise}
  As $x$ increases from $1$ to $(1+h)$, the total surface area of the
  box changes by
  \[
  S(1+h) - S(1) = \answer{-8}h^2 + \answer{-16}h
  \]
  So as $x$ increases by $h$, the surface areas decreases by
  $\answer{16hx+8h^2}$.

  As a final exercise, illustrate this change in a picture and share your picture with a classmate.
  \begin{multipleChoice}
    \choice[correct]{I have done this.}
    \choice{I have not done this.}
  \end{multipleChoice}
\end{exercise}
\end{document}
