\documentclass{ximera}

\input{../../preamble.tex}

\author{Gregory Hartman \and Matthew Carr\and Nela Lakos}
\license{Creative Commons 3.0 By-NC}
\acknowledgement{https://github.com/APEXCalculus}

\outcome{Add up a large number of terms quickly using sigma notation.}
\outcome{Compute left, right, and midpoint Riemann Sums with many rectangles.}
\outcome{Understand the relationship between area under a curve and sums of rectangles.}
\outcome{Approximate area under a curve.}
\outcome{Understand how the area under a curve is related to the antiderivative.}
\outcome{Use limits of Riemann sums to find the exact area under a curve.}
\outcome{Understand how Riemann sums are used to find exact area.}

\begin{document}
\begin{exercise}


Answer the following, using the formula $\sum_{k=1}^{n}k=\frac{n(n+1)}{2}$.
\begin{enumerate}
\item Find a right Riemann sum for $3x+1$ with $n$ subintervals on the interval $[-1,3]$.
\begin{hint}
Recall: $\Delta x=\frac{3-(-1)}{n}=\frac{4}{n}$

$x_k^*=x_k=-1+k\Delta x=-1+\frac{4k}{n}$, for $k=1,2,...,n.$
\end{hint}
\begin{hint}
\[
\sum_{k=1}^n f(x_k^*)\Delta x=\sum_{k=1}^n f(x_k)\Delta x=\sum_{k=1}^n f\left(-1+\frac{4k}{n}\right)\frac{4}{n}
\]
where $f(x)=3x+1$.
\end{hint}
\begin{prompt}
\[
\sum_{k=1}^n \left(3\cdot\left(\answer{-1+\frac{4k}{n}}\right)+\answer{1}\right)\cdot\frac{4}{n}
\]
\end{prompt}

\item Compute the limit as $n\to\infty$ of your answer above. Here $f(x)=3x+1$.
\begin{hint}
Use properties of sums and the given formula to evaluate the right Riemann sum. Then take the limit as $n\to\infty$.
\end{hint}
\begin{prompt}

$\lim_{n\to\infty}\sum_{k=1}^n f(x_k^*)\Delta x= \answer{16}$

\end{prompt}
\item Does the answer above agree with the value of the integral
  $\int_{-1}^{3}(3x+1)\d x$? \begin{multipleChoice}
  \choice[correct]{Yes} \choice{No}
\end{multipleChoice}
\end{enumerate}
\end{exercise}
\end{document}
