\documentclass{ximera}

\input{../preamble.tex}

\outcome{Define a slant asymptote.}
\outcome{Approximate a slant asymptote from the graph of a function.}
\outcome{Find slant asymptotes algebraically and graphically.}

\title[Dig-In:]{Slant asymptotes}

\begin{document}
\begin{abstract}
We explore functions that ``shoot to infinity'' at certain points in
their domain.
\end{abstract}
\maketitle

If we think of an asymptote as a ``line that a function resembles
when the input is large,'' then there are three types of asymptotes:
\begin{enumerate}
\item Vertical Asymptotes.
\item Horizontal Aymptotes.
\item Slant Asymptotes.
\end{enumerate}

Here, we are classifying lines into essentially three types: vertical lines, 
horizontal lines, and lines that are not horizontal or vertical.  We could 
call this last type ``slant lines''.

Recall, a function $f$ has a vertical asymptote at $x=a$ if

\[
\lim_{x\to a} f(x) = \pm\infty, \qquad \lim_{x\to a+} f(x) = \pm\infty, \qquad\text{or}\qquad \lim_{x\to a-} f(x) = \pm\infty,
\]
In this case, the asymptote is the vertical line
\[
x = a.
\]
We have also seen that a function $f$ has a horizontal asymptote if
\[
\lim_{x\to \infty} f(x) = L \qquad\text{or}\qquad \lim_{x\to -\infty}
f(x) = L,
\]
and in this case, the asymptote is the horizontal line
\[
\l(x) = L.
\]
On the other hand, a \textit{slant asymptote} is a somewhat different
beast.

\begin{definition}\index{slant asymptote}\index{asymptote!slant}
  If there is a nonhorizontal line $\l(x) = m\cdot x+b$ such that
  \[
  \lim_{x\to \infty}\left(f(x) - \l(x)\right) = 0 \qquad\text{or}\qquad \lim_{x\to -\infty} \left( f(x) - \l(x)\right) = 0,
  \]
  then $\l$ is a \dfn{slant asymptote} for $f$.
\end{definition}
\begin{question}
  Consider the graph of the following function. 
  \begin{image}
    \begin{tikzpicture}
      \begin{axis}[
          samples=100,
          axis lines =middle, xlabel=$x$, ylabel=$y$,
          every axis y label/.style={at=(current axis.above origin),anchor=south},
          every axis x label/.style={at=(current axis.right of origin),anchor=west},
        ]
	\addplot [very thick, penColor, smooth,domain=0:5] {x/2-1};
        \addplot [very thick, penColor, smooth,domain=-5:0] {sqrt(-x)};
        \addplot[color=penColor,fill=background,only marks,mark=*] coordinates{(0,-1)};  %% open hole
        \addplot[color=penColor,fill=penColor,only marks,] coordinates{(0,0)};  %% closed hole
      \end{axis}
    \end{tikzpicture}
  \end{image}
  What is the slant asymptote of this function?
  \begin{prompt}
    \[
    \l(x) = \answer{x/2 -1}
    \]
  \end{prompt}
\end{question}

To analytically find slant asymptotes, one must find the required
information to determine a line:
\begin{itemize}
\item The slope.
\item The $y$-intercept.
\end{itemize}
While there are several ways to do this, we will give a method that is
fairly general.


\begin{example}
  Find the slant asymptote of
  \[
f(x) = \frac{3x^2+x+2}{x+2}.
  \]
  \begin{explanation}
    We are looking to see if there is a line $\l$ such that
    \[
    \lim_{x\to \infty}\left(f(x) - \l(x)\right) = 0\qquad\text{or}\qquad \lim_{x\to -\infty} \left( f(x) - \l(x)\right) = 0.
    \]
    First, let's consider the limit as $x$ approaches positive infinity.  We will imagine that we have such a line
    \[
    \l(x) = m\cdot x + b
    \]
    and attempt to find the correct values for $m$ and $b$.  Let's look again at our limit.  We are assuming:
\[
    \lim_{x \to \infty} \left (\frac{3x^2 + x + 2}{x+2} - (mx + b) \right) = 0.
\]
    We know that $f(x)$ is continuous everywhere except at $x = -2$ and $\l(x)$ is continuous everywhere, so 
    we can apply our limit laws away from $x = -2$.  We're looking at large values of $x$, so this is no problem.  We use the fact that the sum of the limits is the limit of the sums.
    \begin{align*}
    \lim_{x \to \infty} \left ( \frac{3x^2 + x + 2}{x+2} \right ) - &\lim_{x\to\infty} (mx+b) = 0 \\
    \lim_{x \to \infty} \frac{3x^2 + x + 2}{x+2} = &\lim_{x \to \infty} (mx+b)
    \end{align*}
    
    We are assuming these two limits are equal.  Dividing by $x$ on the right hand side makes the limit equal to $m$:
    \begin{align*}
    m = \lim_{x \to \infty} \left(\frac{mx}{x} + \frac{b}{x}\right) = \lim_{x \to \infty} \left(\frac{mx+b}{x}\right).
    \end{align*}
    To find the value of $m$, then, we can divide the left hand side by $x$ and evaluate the limit.   We see the following. 
    \begin{align*}
      m &=\lim_{x\to\infty}\frac{\frac{3x^2+x+2}{x+2}}{x}\\
      &= \lim_{x\to\infty}\frac{\frac{3x^2+x+2}{x+2}}{x}\\
      &= \lim_{x\to\infty}\frac{3x^2+x+2}{x^2+2x}\\
      &= \lim_{x\to\infty}\left(\frac{3x^2+x+2}{x^2+2x}\cdot\frac{1/x^2}{1/x^2}\right)\\
      &= \lim_{x\to\infty}\frac{3+1/x+2/x^2}{1+2/x}\\
      &= \answer[given]{3}.
    \end{align*}
    So $m=3$.  We now know that
    \[
    \lim_{x \to \infty}\frac{3x^2 +x+2}{x+2} = \lim_{x \to \infty} (3x + b)
    \]
    for some value of $b$.  To find the $y$-intercept $b$, we use a similar method.  Notice that
    \[
    \lim_{x \to \infty} 3x + b - 3x = \lim_{x \to \infty} b = b,
    \]
    so if we subtract $3x$ from the right hand side, we are left with just $b$.  Since the two sides are equal, subtracting $3x$ from the left hand side and evaluating the limit will give us the value for $b$.  We write the following.
    \begin{align*}
      b &=\lim_{x\to\infty} \left(\frac{3x^2+x+2}{x+2} - 3x\right)\\
      &=\lim_{x\to\infty} \left( \frac{3x^2+x+2}{x+2} - \frac{3x^2 + 6x}{x+2}\right) \\
      &=\lim_{x\to\infty} \frac{3x^2+x+2-3x^2-6x}{x+2}\\
      &=\lim_{x\to\infty}  \frac{-5x+2}{x+2}\\
      &=\lim_{x\to\infty} \left(\frac{-5x+2}{x+2}\cdot\frac{1/x}{1/x}\right)\\
      &=\lim_{x\to\infty} \frac{-5+2/x}{1+2/x}\\
      &= \answer[given]{-5}.
    \end{align*}
    By this method, we have determined that 
    \[
    \lim_{x\to\infty}\left(\frac{3x^2+x+2}{x+2} - (3x-5) \right) = 0
    \]
   In other words, $\l(x) = x-3$ is a slant asymptote for our function $f(x)$.  
   
   You should check that we get the same slant asymptote $\l(x) = 3x-5$ when we take the limit to negative infinity as well.
  \end{explanation}
\end{example}





\end{document}
