\documentclass{ximera}

\input{../../preamble.tex}

\author{Jim Talamo}
\license{Creative Commons 3.0 By-NC}


\outcome{Understand the relationship between the sequence of remainders and the convergence of the series.}

\begin{document}

\begin{exercise}

Consider the series $\sum_{k=1}^{\infty} \frac{1}{k^5}$.  

<<<<<<< HEAD
Does the integral test apply?

\begin{multipleChoice}
\choice[correct]{Yes.}
\choice{No.}
\end{multipleChoice}

What does the integral test tell us?
\begin{multipleChoice}
\choice{The series converges to $\frac{1}{4}$.}
\choice[correct]{The series converges, but we do not know its value yet.}
\choice{The series diverges.}
\end{multipleChoice}

Suppose we want to approximate $\sum_{k=1}^{\infty} \frac{1}{k^5}$ by using $\sum_{k=1}^{6} \frac{1}{x^5}$.  

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{exercise}
Using the remainder results for the integral test, find the \emph{maximum} possible error made by using $\sum_{k=1}^{6} \frac{1}{k^5}$ to approximate $\sum_{k=1}^{\infty} \frac{1}{k^5}$.  Report your answer to six decimal places.

\[
\textrm{maximum possible error} \leq \answer[tolerance=.000002]{.000193}
\]


Thus, to six decimal places, $\sum_{k=1}^{\infty} \frac{1}{k^5} \leq \answer[tolerance=.000002]{1.036983}$

\begin{hint}
We can determine an upper bound for the error from the inequality $ r_n \leq \int_{n}^{\infty} \frac{1}{x^5} \d x$. 

\[
\int_n^\infty \frac{1}{x^5} \d x = \lim_{b \to \infty} \eval{\answer{-\frac{1}{4x^4}}}_n^b = \lim_{b \to \infty} \eval{-\frac{1}{4b^4}+\frac{1}{4n^4}} =\answer{\frac{1}{4n^4}}. 
\]

Since we want to approximate $\sum_{k=1}^{\infty} \frac{1}{k^5}$ by using $\sum_{k=1}^{6} \frac{1}{k^5}$, we should use $n=\answer{6}$, and doing so gives us the maximum possible error.
\end{hint}
\end{exercise}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{exercise}
Using the remainder results for the integral test, find the \emph{minimum} possible error made by using $\sum_{k=1}^{6} \frac{1}{k^5}$ to approximate $\sum_{k=1}^{\infty} \frac{1}{k^5}$.  Report your answer to six decimal places.

\[
\textrm{minimum possible error} \geq \answer[tolerance=.000002]{.000104}
\]

Thus, to six decimal places, $\sum_{k=1}^{\infty} \frac{1}{k^5} \geq \answer[tolerance=.000002]{ 1.036895}$

\begin{hint}
We can determine a lower bound for the error from the inequality $ r_n \geq \int_{n+1}^{\infty} \frac{1}{x^5} \d x$. 

\[
\int_{n+1}^\infty \frac{1}{x^5} \d x = \lim_{b \to \infty} \eval{\answer{-\frac{1}{4x^4}}}_{n+1}^b = \lim_{b \to \infty} \eval{-\frac{1}{4b^4}+\frac{1}{4n^4}} =\answer{\frac{1}{4(n+1)^4}}. 
\]

Since we want to approximate $\sum_{k=1}^{\infty} \frac{1}{k^5}$ by using $\sum_{k=1}^{6} \frac{1}{k^5}$, we should use $n=\answer{6}$, and doing so gives us the minimum possible error.

Since $\sum_{k=1}^{6} \frac{1}{k^5} = \answer[tolerance=.000002]{1.036794}$, we see that 

\begin{align*}
\sum_{k=1}^{\infty} \frac{1}{k^5} & \geq \sum_{k=1}^{6} \frac{1}{k^5} + \textrm{minimum possible error} \\
& \leq 1.036794 + .000104 \\
\end{align*}

\end{hint}
\end{exercise}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{exercise}
How many in the series $\sum_{k=1}^{\infty} \frac{1}{k^5}$ do we need to add to ensure that the approximation is valid to within $.001$ of the exact value of the series?  In other words, what is the smallest value of $N$ that should be chosen so the upper bound for the remainder $r_N$ is no more than $.001$?

\[
N=\answer{8}
\]
=======
\begin{selectAll}
\choice{$\sum_{k=1}^{\infty} 2^{1-k}$}
\choice{$\sum_{k=1}^{\infty} \frac{2^{2k}}{3^k}$}
\choice{$\sum_{k=1}^{\infty} \bigg(\sin(k)-\sin(k+1) \bigg)$}
\choice{$\sum_{k=1}^{\infty}  \left(\frac{1}{k^2} - \frac{1}{(k+1)^2}\right)$}
\end{selectAll}
>>>>>>> 6370795d971fad0695b8bbcf11626b151dd3c490

\begin{hint}
we want $r_N \leq .001$.  Since we have that $r_N \leq \int_{N}^{\infty} \frac{1}{x^5} \d x$, we can achieve this by setting $\int_{N}^{\infty} \frac{1}{x^5} \d x \leq .001$. 

\[
\int_N^\infty \frac{1}{x^5} \d x = \lim_{b \to \infty} \eval{\answer{-\frac{1}{4x^4}}}_N^b = \lim_{b \to \infty} \eval{-\frac{1}{4b^4}+\frac{1}{4N^4}} =\answer{\frac{1}{4N^4}}. 
\]


Thus, we set $\answer{\frac{1}{4N^4}} \leq .001$ and find $N \geq 7.95$, and we should thus choose $N = \answer{8}$.
\end{hint}
\end{exercise}
<<<<<<< HEAD





\end{exercise}
\end{document}
=======
\end{document}
>>>>>>> 6370795d971fad0695b8bbcf11626b151dd3c490
