\documentclass{ximera}

\input{../../preamble.tex}

\author{Jim Talamo}
\license{Creative Commons 3.0 By-NC}


\outcome{Understand the relationship between the sequence of remainders and the convergence of the series.}

\begin{document}

\begin{exercise}

A student claims that to within $.001$, $\sum_{k=1}^{\infty} \frac{4}{k^{3/2}} = 9.651$.  Is the student correct?

\begin{multipleChoice}
\choice{Yes.}
\choice{Not necessarily.}
\choice[correct]{No.}
\end{multipleChoice}

How would you establish this?

\begin{multipleChoice}
\choice{Approximate the series by a finite sum and find an upper bound for the error using the remainder results for the integral test.}
\choice[correct]{Approximate the series by a finite sum and find an lower bound for the error using the remainder results for the integral test.}
\choice{Calculate $\sum_{k=1}^{1000} \frac{4}{k^{3/2}}$ and check to see if it's within $.001$ of the student's answer.}
\end{multipleChoice}

\begin{feedback}
The student actually calculated $\sum_{k=1}^{100} \frac{4}{k^{3/2}}$ to obtain this approximation.  According to the integral test, what is the minimum possible error that the student should expect by using this approximation?

The minimum possible error according to the integral test is $\answer[tolerance=.002]{0.796}$ to three decimal places.  
\end{feedback}

\begin{selectAll}
\choice{$\sum_{k=1}^{\infty} 2^{1-k}$}
\choice{$\sum_{k=1}^{\infty} \frac{2^{2k}}{3^k}$}
\choice{$\sum_{k=1}^{\infty} \bigg(\sin(k)-\sin(k+1) \bigg)$}
\choice{$\sum_{k=1}^{\infty}  \left(\frac{1}{k^2} - \frac{1}{(k+1)^2}\right)$}
\end{selectAll}

\begin{hint}
Which inequality in the remainder results for the integral test establishes a lower bound for the error?
\end{hint}

\begin{feedback}
This means that the minimum possible value for $\sum_{k=1}^{\infty} \frac{4}{k^{3/2}}$ should be $\sum_{k=1}^{100} \frac{4}{k^{3/2}} +0.796 = 10.448$ to three decimal places.
\end{feedback}

\begin{exercise}
Find the value of $\sum_{k=1}^{\infty} \frac{4}{k^{3/2}}$ to within a precision of $.001$.  Report your answer to four decimal places.

\[
\sum_{k=1}^{\infty} \frac{4}{k^{3/2}} = \answer[tolerance=.001]{10.4495}
\]

\begin{hint}
One way to do this is to find $N$ so the upper bound for the error $r_N \leq .001$, then find the approximation by calculating $\sum_{k=1}^{N} \frac{4}{k^{3/2}}$.  However, this will produce a rather large value for $N$ (in fact, you should find that $N=4 \times 10^6$!), and it will take a computer a \emph{long} time to compute the approximation.  

Another way that is much more computationally efficient is to require that the difference between the upper estimate for the series and the lower estimate be no more than $.001$; that is note that for every $n$,

\[
\int_{n+1}^\infty \frac{4}{x^{3/2}} \d x  +s_n \leq \sum_{k=1}^{\infty} a_k \leq \int_{n}^\infty \frac{4}{x^{3/2}} \d x  +s_n.
\]
 
If the difference between the left and right sides of the inequality is less than $.001$, then the infinite series must be within $.001$ of its minimum possible value (the lefthand side) and its maximum possible value (the righthand side).
 
 Now, note that

\begin{align*}
RHS-LHS &\leq .001 \\
\left( \int_{n}^\infty \frac{4}{x^{3/2}} \d x  +s_n\right) - \left( \int_{n+1}^\infty \frac{4}{x^{3/2}} \d x  +s_n \right) &\leq .001 \\
 \int_{n}^\infty \frac{4}{x^{3/2}} \d x  - \int_{n+1}^\infty \frac{4}{x^{3/2}} \d x   &\leq .001 \\
  \int_{n}^\infty \frac{4}{x^{3/2}} \d x  +  \int_{\infty}^{n+1} \frac{4}{x^{3/2}} \d x   &\leq .001 \\
    \int_{n}^{n+1} \frac{4}{x^{3/2}}  &\leq .001 \\
\end{align*} 

Noting that $\int_n^{n+1}  \frac{4}{x^{3/2}} \d x = \answer{\frac{8}{\sqrt{n}} - \frac{8}{\sqrt{n+1}}}$, we can find that the smallest integer value for $N$ for which $\int_n^{n+1}  \frac{4}{x^{3/2}} \d x \leq .001$ is $N= \answer{252}$.  This is much better than the previous value found!
 
 We now know that the actual value for the series must be between $s_{252} + \int_{253}^\infty \frac{4}{x^{3/2}} \d x = \answer[tolerance=.002]{10.449}$ and $s_{252} + \int_{252}^\infty \frac{4}{x^{3/2}} \d x = \answer[tolerance=.002]{10.450}$
\end{hint}

\end{exercise}
\end{exercise}
\end{document}
