\documentclass{ximera}

\input{../../preamble.tex}

%\outcome{Find tangent lines to parametric curves}
\author{Jim Talamo}

\begin{document}
\begin{exercise}
Let $R(a)$ be the region bounded by the $x$-axis, the curve $C$ (defined by $x(t) = t^2, y(t)=t$), and the tangent line to the graph of $C$ at $t=a$.  The next part of the problem asks questions about this area and how it changes as $t$ increases.

In order to visualize this, it will be useful to make a Desmos Worksheet to animate the curve.  To see this curve, do the following.

\begin{itemize}
\item[1.] Go to https://www.desmos.com/ and click ``Start Graphing".
\item[2.] Change your viewing window using ``Graph Settings''.  This is done by clicking on the little wrench that appears in the upper righthand corner.  Use the window $-20\leq x \leq 20$ and $-3 \leq y \leq 8$.
\item[3.] Type the following expressions \emph{exactly} as written below on the indicated lines.
\begin{itemize}
\item In Line 1, type: \verb|X(t) = t^2|  
\item In Line 2, type: \verb|Y(t) = t|
\item Then, uncheck the boxes left of the expressions to make the graphs disappear.
\item In Line 3, type \verb|(X(t),Y(t))| and select $0\leq t \leq 5$ when the bounds for $t$ arise.  
\end{itemize}
What kind of curve do you see?

\begin{multipleChoice}
\choice{An entire parabola.}
\choice{The right half of a parabola that opens up.}
\choice{The left half of a parabola that opens up.}
\choice[correct]{The upper half of a parabola that opens right.}
\choice{The lower half of a parabola that opens right.}
\end{multipleChoice}

\item[3.] Now, uncheck the box next to: \verb|(X(t),Y(t))|.
\item[4.] In Line 4, type \verb|(X(a),Y(a))|.  When ``add slider'' pops up as an option, click on it, and choose the same range of values for $a$ as you did for $t$.  That is, select $0 \leq a \leq 5$.
\item[5.] Go back to Line 3 and add ``$\{t<a\}$" right after $Y(t)$ (before the final parenthesis).  

Line 3 should now read $\verb|(X(t),Y(t)\{t<a\} )|$.

\item[6.] Now click on the Play button next to $a$.  You should see an animation for the curve.  
\begin{itemize}
\item Experiment with the speed of the animation by by clicking on the expression in the ``$<< \quad >>$'' box (which appears only while the curve is being animated)
\item Experiment with the direction in which the curve is traced out by clicking on the arrows that appear to the left of the ``$<< \quad >>$'' box.  When both arrow point right, this shows the \emph{positive orientation} (the direction in which the curve is traversed as $t$ increases).
\end{itemize}

\item[7.] In Line 6, graph the tangent line at $t=a$ by typing \verb|x-2ay=-a^2|.
\end{itemize}

Check ``yes'' once you've done all of these steps to continue.

\begin{multipleChoice}
\choice[correct]{Yes.}
\choice{No.}
\end{multipleChoice}

\begin{exercise}
In order to graph the area, we need to find suitable inequalities.  Eventually, we will want to find the area for any $a$-value, and to do this, we will have to integrate.

\begin{itemize}
\item If we integrate with respect to $x$, we will need at least $\answer{2}$ integral(s).
\item If we integrate with respect to $y$, we will need at least $\answer{1}$ integral(s).
\end{itemize}

It is therefore more efficient to integrate with respect to $\answer{y}$.

\begin{exercise}
From the Desmos picture, the left and right curves are given by:

\[
x_{left} = \answer{-a^2+2a \cdot y} \qquad \textrm{ and } \qquad x_{right} = \answer{y^2}
\]
(type your answers in terms of $a$ and $y$).

Now, go to your Desmos sheet and type 

\[
\verb|x_{left}<x<x_{right}\left\{y>0\right\}| 
\]

into Line 7.  Use the expressions you just found for $x_{left}$ and $x_{right}$ (i.e. don't just type ``$x_{left}$'' and ``$x_{right}$'' into Desmos).

Now, go back to Line 5 and set $a=0$.  Click on the ``Play'' icon.   Notice that the area of $R(a)$ \wordChoice{\choice[correct]{increases}\choice{is constant}\choice{decreases}} as $a$ increases.

\end{exercise}



\end{exercise}

 
\end{exercise}
\end{document}
