\documentclass{ximera}

\input{../preamble.tex}

\title[Dig-In:]{Basic rules of differentiation}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle


\section{Basic derivative rules}


It is tedious to compute a limit every time we need to know the
derivative of a function.  Fortunately, we can develop a small
collection of examples and rules that allow us to compute the
derivative of almost any function we are likely to encounter.  We will
start simply and build-up to more complicated examples.


\subsection*{The constant rule}

The simplest function is a constant function.  Recall that derivatives
measure the rate of change of a function at a given point. Hence, the
derivative of a constant function is zero. For example:
\begin{itemize}
\item The constant function plots a horizontal line---so the slope of
  the tangent line is $0$.
\item If $s(t)$ represents the position of an object with respect to
  time and $s(t)$ is constant, then the object is not moving, so its
  velocity is zero. Hence $\dd{t} s(t) = 0$.
\item If $v(t)$ represents the velocity of an object with respect to
  time and $v(t)$ is constant, then the object's acceleration is
  zero. Hence $\dd{t} v(t) = 0$.
\end{itemize}
The examples above lead us to our next theorem.


\todo{To gain intuition, you should compute the derivative of
  $f(x) = 6$ using the limit definition of the derivative.}

\begin{theorem}[The Constant Rule]\index{derivative rules!constant}\index{constant rule}
Given a constant $c$,
\[
\ddx c = 0.
\]
\end{theorem}

\begin{proof}
From the limit definition of the derivative, write
\begin{align*}
\ddx c &= \lim_{h\to 0}\frac{c-c}{h}\\
&= \lim_{h\to 0} \frac{0}{h}\\
&= \lim_{h\to 0} 0 = 0.
\end{align*}
\end{proof}


\subsection*{The power rule}

Now let's examine derivatives of powers of a single variable.  Here we
have a nice rule.

\todo{To gain intuition, you should compute the derivative of
  $f(x) = x^3$ using the limit definition of the derivative.}

\begin{theorem}[The Power Rule]\index{derivative rules!power}\index{power rule}\label{T:powerrule}
For any real number $n$, 
\[
\ddx x^n = n x^{n-1}.
\]
\end{theorem}
\todo{Recall, the \textbf{Binomial Theorem}\index{Binomial Theorem} states that if $n$
  is a nonnegative integer, then
\[
(a+b)^n = a^nb^0 + \binom{n}{1} a^{n-1}b^1 + \dots + \binom{n}{n-1} a^{1}b^{n-1} +  a^{0}b^n   
\]
where
\[
\binom{n}{k} = \frac{n!}{k!(n-k)!}.
\]
}
\begin{proof}
At this point we will only prove this theorem for $n$ being a positive
integer. Later in Section~\ref{S:logdiff}, we will give the complete proof. From the limit
definition of the derivative, write
\[
\ddx x^n = \lim_{ h\to0} \frac{(x+ h)^n-x^n}{h}.
\]
Start by expanding the term $(x+h)^n$
\[
\ddx x^n=\lim_{h\to0} \frac{x^n + \binom{n}{1}x^{n-1} h + \binom{n}{2}x^{n-2} h^2+\cdots+\binom{n}{n-1}x h^{n-1} +  h^n-x^n}{h}
\]
Note, by the Binomial Theorem, we write $\binom{n}{k}$ for the
coefficients. Canceling the terms $x^n$ and $-x^n$, and noting
$\binom{n}{1}= \binom{n}{n-1}= n$, write
\begin{align*}
\ddx x^n&=\lim_{h\to0} \frac{nx^{n-1} h + \binom{n}{2}x^{n-2} h^2+\cdots+\binom{n}{n-1}x h^{n-1} +  h^n}{h} \\
&=\lim_{h\to0} nx^{n-1} + \binom{n}{2}x^{n-2} h+\cdots+\binom{n}{n-1}x h^{n-2} +  h^{n-1}.
\end{align*}
Since every term but the first has a factor of $h$, we see
\[
\ddx x^n = \lim_{ h\to0}\frac{(x+ h)^n-x^n}{h} = nx^{n-1}.
\]
\end{proof}

Now we will show you several examples. We begin with something basic.

\begin{example}
Compute
\[
\ddx x^{13}.
\]

Applying the power rule, we write
\[
\ddx x^{13} = 13x^{12}.
\]
\end{example}

Sometimes, it is not as obvious that one should apply the power rule.

\begin{example}
Compute
\[
\ddx \frac{1}{x^4}.
\]

Applying the power rule, we write
\[
\ddx \frac{1}{x^4} = \ddx x^{-4} = -4x^{-5}.
\]
\end{example}

The power rule also applies to radicals once we rewrite them as exponents.

\begin{example}
Compute
\[
\ddx \sqrt[5]{x}.
\]

Applying the power rule, we write
\[
\ddx \sqrt[5]{x} = \ddx x^{1/5} = \frac{x^{-4/5}}{5}.
\]
\end{example}






\subsection*{The sum rule}

We want to be able to take derivatives of functions ``one piece at a
time.'' The \textit{sum rule} allows us to do this. The sum rule says
that we can add the rates of change of two functions to obtain the
rate of change of the sum of both functions. For example, viewing the
derivative as the velocity of an object, the sum rule states that the
velocity of the person walking on a moving bus is the sum of the
velocity of the bus and the walking person.


\begin{theorem}[The Sum Rule]\index{derivative rules!sum}\index{sum rule}\label{theorem:sum rule}
If $f(x)$ and $g(x)$ are differentiable and $c$ is a constant, then 
\begin{enumerate}
\item\label{SR:1} $\ddx \big( f(x) + g(x)\big) = f'(x) + g'(x)$,
\item $\ddx \big( f(x) - g(x)\big) = f'(x) - g'(x)$,
\item $\ddx \big(c\cdot f(x)\big) = c\cdot f'(x)$.
\end{enumerate}
\end{theorem}

\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            clip=false,
            domain=0:5, 
            ymin=0, ymax=5,
            xlabel=$x$, ylabel=$y$,
            axis lines=center,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={3},ytickmin=1,ytickmax=0,
            xticklabels={$a$},
            axis on top,
          ]          
                    
          \addplot [penColor5,very thick] plot coordinates {(4,3/2.5) (4,4/2.5)};
          \addplot [dashed, very thick, textColor] plot coordinates {(3,3/2.5) (4,3/2.5)};

          \addplot [penColor4,very thick] plot coordinates {(4,3/1.5+.5) (4,4/1.5+.5)};
          \addplot [dashed, very thick, textColor] plot coordinates {(3,3/1.5+.5) (4,3/1.5+.5)};

          \addplot [very thick, penColor5] plot coordinates {(4,3/2.5+3/1.5+.5) (4,3/1.5+.5 + 4/2.5)};
          \addplot [very thick, penColor4] plot coordinates {(4,3/1.5+.5 + 4/2.5) (4,4/2.5 + 4/1.5 +.5)};
          \addplot [dashed, very thick, textColor] plot coordinates {(3,3/2.5+3/1.5+.5) (4,3/2.5+3/1.5+.5)};

          \addplot [dashed, textColor] plot coordinates {(3,0) (3,3/2.5+3/1.5+.5)};
        
          \addplot [very thick,penColor,smooth] {x/2.5};
          \addplot [very thick,penColor2,smooth] {x/1.5 +.5};         
          \addplot [very thick,penColor3,smooth,domain=(0:4.5)] {x/2.5+x/1.5 +.5};         

          \node at (axis cs:1.7,1) [anchor=west,penColor] {$f(x)$};
          \node at (axis cs:2.2,2.4) [anchor=west,penColor2] {$g(x)$};
          \node at (axis cs:1.2,3) [anchor=west,penColor3] {$f(x)+g(x)$};

          \node at (axis cs:4,1.4) [anchor=west,penColor5] {$f'(a)h$};
          \node at (axis cs:4,2.9) [anchor=west,penColor4] {$g'(a)h$};
          \node at (axis cs:4,4.3) [anchor=west] {${\color{penColor5}f'(a)h}+{\color{penColor4}g'(a)h}$};
          \node at (axis cs:3.5,1.2) [anchor=north] {$\underbrace{\hspace{.50in}}_{h}$};

          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(3,3/2.5)};  %% closed hole          
          \addplot[color=penColor2,fill=penColor2,only marks,mark=*] coordinates{(3,3/1.5+.5)};  %% closed hole          
          \addplot[color=penColor3,fill=penColor3,only marks,mark=*] coordinates{(3,3/2.5+3/1.5+.5)};  %% closed hole          
        \end{axis}
\end{tikzpicture}
%% \label{figure:sum-rule}
%% \caption{A geometric interpretation of the sum rule. Since every
%%   point on $f(x)+g(x)$ is the sum of the corresponding points on
%%   $f(x)$ and $g(x)$, increasing $a$ by a ``small amount'' $h$,
%%   increases $f(a)+g(a)$ by the sum of $f'(a)h$ and
%%   $g'(a)h$. Hence, 
%% \[
%% \frac{\Delta y}{\Delta x}\approx \frac{f'(a)h+g'(a)h}{h} = f'(a) + g'(a).
%% \]}
\end{image}

\begin{proof}
We will only prove part~\ref{SR:1} above, the rest are similar. Write
\begin{align*}
\ddx\big(f(x)+g(x)\big) &= \lim_{ h\to 0} {f(x+h)+g(x+ h) - (f(x)+g(x))\over  h}  \\
&= \lim_{ h\to 0} {f(x+h)+g(x+ h) - f(x)-g(x)\over  h}  \\
&= \lim_{ h\to 0} {f(x+h)-f(x) +g(x+ h) -g(x)\over  h}  \\
&= \lim_{ h\to 0} \left({f(x+h)-f(x)\over  h}  +{g(x+ h) -g(x)\over  h}\right)  \\
&= \lim_{ h\to 0} {f(x+h)-f(x)\over  h}  +
\lim_{ h\to 0} {g(x+ h) -g(x)\over  h}  \\
&=f'(x)+g'(x).
\end{align*}
\end{proof}



\begin{example}
Compute
\[
\ddx \left( x^5+\frac{1}{x}\right).
\] 

Write
\begin{align*}
\ddx \left(x^5+\frac{1}{x}\right) &= \ddx x^5 + \ddx x^{-1} \\
&=5x^4 - x^{-2}.
\end{align*}
\end{example}

\begin{example}
Compute
\[
\ddx \left(\frac{3}{\sqrt[3]{x}}-2\sqrt{x}+\frac{1}{x^7}\right).
\]

Write
\begin{align*}
\ddx \left(\frac{3}{\sqrt[3]{x}}-2\sqrt{x}+\frac{1}{x^7}\right) &= 3\ddx x^{-1/3} -2\ddx x^{1/2}+\ddx x^{-7}\\
&=-x^{-4/3} - x^{-1/2}-7x^{-8}.
\end{align*}
\end{example}



\end{document}
