\documentclass{ximera}
\input{../../preamble.tex}
\author{Steven Gubkin}
\license{Creative Commons 3.0 By-NC}
\outcome{Identify word problems as related rates problems.}
\outcome{Solve related rates word problems.}
\outcome{Translate word problems into mathematical expressions.}
\begin{document}


\begin{exercise}


A $12$ foot ladder is leaning against a wall.  At the instant that the foot of the ladder is $3$ feet away from the wall, the foot of the ladder is moving away from the wall at a rate of $5 \frac{\textrm{ft}}{\textrm{s}}$.  At what rate is the top of the ladder falling down the wall at this time?

\begin{hint}
We can model this in the following way.  Let $x(t)$ be the position of the foot of the ladder at time $t$ on the, and $y(t)$ be the position of the top of the ladder at time $t$. 
\end{hint}

\begin{hint}
	$x(t)$ and $y(t)$ must satisfy the relation

\[
\left( x(t) \right)^2 + \left( y(t)\right)^2 = 12^2
\]

by the Pythagorean theorem.
\end{hint}

\begin{hint}
	Differentiating with respect to time, we obtain

\[
2x(t)x'(t)+2y(t)y'(t) = 0.
\]
\end{hint}

\begin{hint}
	At the time (lets call it $t_0$) under consideration, we know $x(t_0)$, $x'(t)$, and we are trying to find $y'(t_0)$.  We can find $y(t_0)$ using the relation  

\begin{align*}
	\left( x(t_0) \right)^2 + \left( y(t_0)\right)^2 &= 12^2\\
	\left( 3 \right)^2 + \left( y(t_0)\right)^2 &= 12^2\\
	y(t_0) = \sqrt{144-9}\\
	y(t_0) = \sqrt{135}.
\end{align*}
\end{hint}

\begin{hint}
	So, going back to the relation between the rates, we have

\begin{align*}
2x(t_0)x'(t_0)+2y(t_0)y'(t_0) &= 0\\
2(3)(5)+2\sqrt{135}y'(t_0) &=0\\
y'(t_0) = -\frac{15}{\sqrt{135}}\frac{\textrm{ft}}{\textrm{s}}.
\end{align*}

So the top of the ladder is falling at  a rate of $\frac{15}{\sqrt{135}}$ feet per second.

\end{hint}

\begin{prompt}
	The top of the ladder is falling at a rate of $\answer{\frac{15}{\sqrt{135}}}$ feet per second.
\end{prompt}

\end{exercise}

\end{document}
