\documentclass{ximera}

\input{../preamble.tex}

\title[Refresh:]{Conic sections and their algebraic descriptions}

\begin{document}
\begin{abstract}
  Conic sections are examples of famous curves in mathematics.
\end{abstract}
\maketitle

\section{Conics Sections}
The \dfn{conic sections} are circles, ellipses, parabolas, and
hyperbolas.  In this section you'll explore the origin of the name
``conic sections.''

The interactive figure shows a cone, formed by rotating a line in
space around some fixed axis, and a plane.  Use the sliders for
$\alpha$ and $\beta$ to produce each of the four conic sections:

If the embedded interactive below does not display correctly, click \link[here]{https://www.geogebra.org/m/xHXwN8wK}

\geogebra{xHXwN8wK}{1366}{480}

\begin{problem}
  Use the figure to help you select the true statements:
  \begin{selectAll}
    \choice{Any plane perpendicular to the cone's axis produces a circle.}
    \choice[correct]{Any plane parallel to the cone's axis and doesn't pass through the vertex produces an hyperbola.}
    \choice[correct]{A plane skew to the cone that doesn't pass through the vertex produces an ellipse.}
    \choice[correct]{A plane can intersect a cone in one point.}
  \end{selectAll}
\end{problem}

\section{Algebraic description of conic sections}
The general algebraic description of all the conic sections in cartesian coordinates is given by
\[
  Ax^2 + Bxy + Cy^2 + Dx + Ey + F = 0,
\]
for constants $A$, $B$, $C$, $D$, $E$, and $F$.
For most of our cases, the conic sections will be symmetric with respect to either the $x$- or $y$-axes.

Therefore we can take $B = 0$ in the above equation and further
simplifications are possible.

You can find the standard forms of circles, ellipses, and hyperbolas online. For instance, check \link[here]{https://www.varsitytutors.com/hotmath/hotmath_help/topics/conic-sections-and-standard-forms-of-equations}

\begin{problem}
  What is the standard from of an ellipse centered at the origin with
  vertices at $(\pm a,0)$ and foci at $(\pm c ,0)$?
  \begin{explanation}
    The standard form of an ellipse is
    \[
    \frac{x^2}{a^2} + \frac{y^2}{b^2}=1\qquad c^2 = a^2 - b^2
    \]
  \end{explanation}
\end{problem}



\begin{problem}
  What is the standard from of a hyperbola centered at the origin with
  vertices at $(0,\pm a)$ and foci at $(0,\pm c)$?
  \begin{explanation}
    The standard form of an ellipse is
    \[
    \frac{y^2}{a^2} - \frac{x^2}{b^2}=1\qquad c^2 = a^2 + b^2
    \]
  \end{explanation}
\end{problem}


\begin{problem}
  Consider
  \[
  y^2 = -36x
  \]
  This is the formula for a
  \wordChoice{\choice{ellipse}\choice{hyperbola}\choice[correct]{parabola}}.
  Its focus is at $\left(-9,0\right)$ and its directrix is the line $x
  = \answer{9}$. 
\end{problem}


\end{document}
