\documentclass{ximera}

\input{../preamble.tex}

\title[Break-Ground:]{Same or different?}

\begin{document}
\begin{abstract}
  Here we see two young mathematicians examining one (or two!)
  functions.
\end{abstract}
\maketitle

Check out this dialogue between two calculus students (based on a true
story):

\begin{dialogue}
\item[Devyn] Riley, I have a pressing question.
\item[Riley] Tell me. Tell me everything.
\item[Devyn] Think about the function
  \[
  f(x) = \frac{x^2 - 3x + 2}{x-2}.
  \]
\item[Riley] OK.
\item[Devyn] Is this function equal to $g(x) = x+1$?
\item[Riley] Well if I plot them with my calculator, they look the
  same.
\item[Devyn] I know!
\item[Riley] And I suppose if I write
  \begin{align*}
    f(x) &= \frac{x^2 - 3x + 2}{x-2} \\
    &= \frac{(x-1)(x-2)}{x-2} \\
    &= x-1 \\
    &= g(x).
  \end{align*}
\item[Devyn] Sure! But what about when $x=2$? In this case
  \[
  g(2) = 1\qquad\text{but}\qquad f(2) \text{ is undefined!}.
  \]
\item[Riley] Hmm. Now I see the problem. Yikes!
\end{dialogue}

\begin{problem}
  Can the same function be represented by different formulas?

  \begin{multipleChoice}
    \choice[correct]{Yes.}
    \choice{No.}
  \end{multipleChoice}

  \begin{problem}
    What about $f(x) = |x|$ and $g(x) = \sqrt{x^2}$?  Are those the same function?

    \begin{multipleChoice}
      \choice[correct]{These are the same function although they are represented by different formulas.}
      \choice{These are different functions because they have different formulas.}
    \end{multipleChoice}
  \end{problem}
\end{problem}

\begin{problem}
  Suppose $f$ and $g$ are functions but the domain of $f$ is different
  from the domain of $g$.  Could it be that $f$ and $g$ are actually
  the same function?

  \begin{multipleChoice}
    \choice{Yes.}
    \choice[correct]{No.}
  \end{multipleChoice}

  \begin{feedback}
    The domain of a function is part of the ``data'' of the function.
    A function is not a rule for transforming the input to the output,
    but rather the relationship between a specified collection of
    inputs (the domain) and possible outputs (the range).
  \end{feedback}
\end{problem}
\input{../leveledQuestions.tex}


\end{document}
