\documentclass{ximera}

\input{../preamble.tex}

\outcome{Find the equation of the tangent line for curves that are not
  plots of functions.}

\title[Break-Ground:]{Standard form}


\begin{document}
\begin{abstract}
  Two young mathematicians discuss the standard form of a line.
\end{abstract}
\maketitle

Check out this dialogue between two calculus students (based on a true
story):

\begin{dialogue}
\item[Devyn] Riley, I think we've been too explicit with each
  other. We should try to be more implicit.
\item[Riley] I. Um. Don't really\dots
\item[Devyn] I mean when plotting things!
\item[Riley] Okay, but I still have no idea what you are talking about.
\item[Devyn] Remember when we first learned the equation of a line, and the ``standard form'' was
  \[
  ax+by = c
  \]
  or something, which is totally useless for graphing. Also a circle is
  \[
  x^2 + y^2 = r^2
  \]
  or something, and here $y$ isn't even a function of
  $x$. 
\item[Riley] Ah, I'm starting to remember.  We can write the same thing in two ways.  For example, if you write
  \[
  y = mx + b, %\,\leftarrow \,\text{where $y$ is \textbf{explicitly} a function of $x$}
  \]
 then $y$ is {\bf explicity} a function of $x$ but if you write
    \[
  ax + by = c, %\,\leftarrow\, \text{here $y$ is \textbf{implicitly} a function of $x$.}
  \]
  then $y$ is {\bf implicitly} a function of $x$.
\item[Devyn] What I'm trying to say is that we need to learn how to
  work with these ``implicit'' functions. 
\end{dialogue}

\begin{problem}
  Consider the unit circle
  \[
  x^2 + y^2 = 1.
  \]
  The point $P=(0,1)$ is on this circle. Reason geometrically to
  determine the slope of the line tangent to $x^2 + y^2 = 1$ at
  $P$.
  \begin{hint}
    Draw a picture.
  \end{hint}
  \begin{prompt}
    The slope is $\answer{0}$.
  \end{prompt}
\end{problem}

\begin{problem}
  Consider the unit circle
  \[
  x^2 + y^2 = 1.
  \]
  The point
  \[
  P=\left(\frac{\sqrt{2}}{2},\frac{\sqrt{2}}{2}\right)
  \]
  is on this circle. Reason geometrically to determine the slope of
  the line tangent to $x^2 + y^2 = 1$ at $P$.
  \begin{hint}
    Draw a picture.
  \end{hint}
  \begin{prompt}
    The slope is $\answer{-1}$.
  \end{prompt}
\end{problem}




\input{../leveledQuestions.tex}


\end{document}
