\documentclass{ximera}

\input{../preamble.tex}

\outcome{}


\title[Break-Ground:]{Slope of a curve}

\begin{document}
\begin{abstract}
Two young mathematicians discuss the novel idea of the ``slope of a curve.''
\end{abstract}
\maketitle


Check out this dialogue between two calculus students (based on a true
story):

\begin{dialogue}
\item[Devyn] Riley, do you remember `slope?'
\item[Riley] Most definitely. ``Rise over run.''
\item[Devyn] You know it.
\item[Riley] ``Change in $y$ over change in $x$.''
\item[Devny] That's right.  
\item[Riley] ``m.''
\item[Devny] Enough! My important question, is could we define
  ``slope'' for a curve?
\item[Riley] Well maybe if we ``zoom in'' on a curve, it would look
  like a line, and then we could call it ``the slope at that point.''
\item[Devyn] Ah! And this ``zoom in'' idea sounds like a limit!
\item[Riley] This is so awesome. We just made math!
\end{dialogue}

The idea above, of the ``slope of a curve at a point,'' will be completely 


\begin{problem}
  Give an argument (similar to the one above) supporting the idea that
  \[
  \lim_{x\to a}\left(f(x) \cdot g(x)\right) = \left(\lim_{x\to a} f(x) \right)\cdot\left(\lim_{x\to a} g(x)\right).
  \]
  \begin{freeResponse}
  \end{freeResponse}
\end{problem}


For the next problems, suppose $L$ is a number near $1$ and that $M$
is a number near $0$.


\begin{problem}

\end{problem}

\begin{problem}
  Using the context above, 
  \[
  \frac{\text{large}}{\text{small}} = ?
  \]
  \begin{multipleChoice}
  \choice[correct]{``large''}
  \choice{``small''}
  \choice{impossible to say}
  \end{multipleChoice}
\end{problem}

\begin{problem}
  Using the context above, 
  \[
  \frac{\text{small}}{\text{small}} = ?
  \]
  \begin{multipleChoice}
  \choice{``large''}
  \choice{``small''}
  \choice[correct]{impossible to say}
  \end{multipleChoice}
\end{problem}

\input{../leveledQuestions.tex}


\end{document}
