\documentclass{ximera}

\input{../preamble.tex}

\title[Dig-In:]{Vertical asymptotes}

\outcome{Recognize when a limit is indicating there is a vertical asymptote.}
\outcome{Evaluate the limit as $x$ approaches a point where there is a vertical asymptote.}
\outcome{Match graphs of functions with their equations based on vertical asymptotes.}
\outcome{Discuss what it means for a limit to equal $\infty$.}
\outcome{Define a vertical asymptote.}
\outcome{Understand the relationship between limits and vertical asymptotes.}
\outcome{Find vertical asymptotes of famous functions.}
\outcome{Find vertical asymptotes by looking at a graph.}


\begin{document}
\begin{abstract}
We explore functions that ``shoot to infinity'' near certain points.

\end{abstract}
\maketitle

Consider the function
\[
f(x) = \frac{1}{(x+1)^2}.
\]
\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            domain=-2:1,
            ymax=100,
            samples=100,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west}
          ]
	  \addplot [very thick, penColor, smooth, domain=(-2:-1.1)] {1/(x+1)^2};
          \addplot [very thick, penColor, smooth, domain=(-.9:1)] {1/(x+1)^2};
          \addplot [textColor, dashed] plot coordinates {(-1,0) (-1,100)};
        \end{axis}
\end{tikzpicture}
%% \caption{A plot of $f(x)=\protect\frac{1}{(x+1)^2}$.}
%% \label{plot:1/(x+1)^2}
\end{image}
While the $\lim_{x\to -1} f(x)$ does not exist, something can still be
said.

\begin{definition}\label{def:inflimit}\index{infinite limit}
	If $f(x)$ grows arbitrarily large as $x$ approaches $a$, we write
	\[ \lim_{x\to a} f(x) = \infty \]
	and say that the limit of $f(x)$ is equal to \dfn{ infinity} as $x$ goes to $a$.

	If $|f(x)|$ grows arbitrarily large as $x$ approaches $a$ and $f(x)$ is
	negative near $a$, we write
	\[ \lim_{x\to a} f(x) = -\infty \]
	and say that the limit of $f(x)$ is equal to \dfn{ negative infinity} as $x$ goes to $a$.
\end{definition}

\begin{question}
  Which of the following are correct?
  \begin{selectAll}
    \choice[correct]{$\lim_{x\to -1} \frac{1}{(x+1)^2} = \infty$}
    \choice{$\lim_{x\to -1} \frac{1}{(x+1)^2} \to \infty$}
    \choice{$f(x) = \frac{1}{(x+1)^2}$, so $f(-1) = \infty$}
    \choice[correct]{$f(x) = \frac{1}{(x+1)^2}$, so as $x\to -1$,
      $f(x) \to \infty$}
  \end{selectAll}
\end{question}

On the other hand, consider the function 
\[
f(x) = \frac{1}{(x-1)}.
\]
\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            domain=-1:2,
            ymax=50,
            ymin=-50,
            samples=100,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west}
          ]
	  \addplot [very thick, penColor, smooth, domain=(1.02:2)] {1/(x-1)};
          \addplot [very thick, penColor, smooth, domain=(-1:.98)] {1/(x-1)};
          \addplot [textColor, dashed] plot coordinates {(1,-50) (1,50)};
        \end{axis}
\end{tikzpicture}
%% \caption{A plot of $f(x)=\protect\frac{1}{x-1}$.}
%% \label{plot:1/(x-1)}
\end{image}
While the two sides of the limit as $x$ approaches $1$ do not agree, we can still consider the one-sided
limits.  We see $\lim_{x\to 1^+} f(x) = \infty$ and $\lim_{x\to 1^-} f(x) =
-\infty$.


\begin{definition}\label{def:vert asymptote}\index{asymptote!vertical}\index{vertical asymptote}
	If at least one of the following hold:
	\begin{itemize}
		\item $\lim_{x\to a} f(x) = \pm\infty$,
		\item $\lim_{x\to a^+} f(x) = \pm\infty$,
		\item $\lim_{x\to a^-} f(x) = \pm\infty$,
	\end{itemize}
	then the line $x=a$ is a \dfn{vertical asymptote} of $f$.
\end{definition}

\begin{warning}
	A vertical asymptote is given by the equation of a vertical line.
\end{warning}

\begin{example}
	Find the vertical asymptotes of 
	\[ f(x) = \frac{x^2-9x+14}{x^2-5x+6}. \]

\begin{explanation}
Since $f$ is a rational function, it is continuous on its domain. So the only points where the function can possibly have a vertical asymptote are zeroes of the denominator. \\
Start by factoring both the numerator and the denominator:
\[
\frac{x^2-9x+14}{x^2-5x+6} = \frac{(x-2)(x-7)}{(x-2)(x-3)}
\]
Using limits, we must investigate what happens with $ f(x)$ when $x\to 2$ and $x\to 3$, since $2$ and $3$ are the only zeroes of the denominator. Write
\begin{align*}
	\lim_{x\to 2} \frac{\cancel{(x-2)}(x-7)}{\cancel{(x-2)}(x-3)} &= \lim_{x\to 2} \frac{(x-7)}{(x-3)}\\
		&= \frac{-5}{-1}\\
		&=5.
\end{align*}

Now write
\begin{align*}
	\lim_{x\to 3} \frac{\cancel{(x-2)}(x-7)}{\cancel{(x-2)}(x-3)} &= \lim_{x\to 3} \frac{(x-7)}{(x-3)}\\
		&= \lim_{x\to 3}\frac{-4}{x-3}.
\end{align*}
Consider the one-sided limits separately. 

When $x\to 3^+$, the quantity $(x-3)$ is positive (because $x>3$) and approaches $0$ and the numerator is negative. This is a \numOverZero form which is negative for all $x$ sufficiently close to, but greater than, $3$. Therefore, $\displaystyle \lim_{x\to 3^+} f(x) = -\infty$.
(\textbf{Note:} Computing this one-sided limit is enough to justify that $x=3$ is a vertical asymptote of the function $f$. We will proceed to calculate the other one-sided limit to see if anything can be said about the two-sided limit.)


When $x\to 3^-$, the quantity $(x-3)$ is negative (because $x<3$) and approaches $0$ and the numerator is negative. This is a \numOverZero form which is positive for all $x$ sufficiently close to, but less than, $3$. Therefore, $\displaystyle \lim_{x\to 3^-} f(x) = \infty$. Since the two one-sided limits did not agree, $\displaystyle \lim_{x\to 3}f(x)$ does not exist.

\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            domain=1:4,
            ymax=50,
            ymin=-50,
            samples=100,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west}
          ]
	  \addplot [very thick, penColor, smooth, domain=(3.02:4)] {(x-7)/(x-3)};
          \addplot [very thick, penColor, smooth, domain=(1:2.98)] {(x-7)/(x-3)};
          \addplot [textColor, dashed] plot coordinates {(3,-50) (3,50)};
          \addplot[color=penColor,fill=background,only marks,mark=*] coordinates{(2,5)};  %% open hole
        \end{axis}
\end{tikzpicture}
%% \caption{A plot of $f(x)=\protect\frac{x^2-9x+14}{x^2-5+6}$.}
%% \label{plot:(x^2-9x+14)/(x^2-5x+6)}
\end{image}
Hence we have a vertical asymptote at $x=3$. Since $\displaystyle \lim_{x\to 2} f(x) = 5$ but $2$ is not in the domain of $f$, the graph of $f$ will have a hole a that point. We know there are no other vertical asymptotes because the function $f$ is continuous on $(-\infty, 2)$, $(2, 3)$, and $(3,\infty)$.
\end{explanation}
\end{example}

Notice that in this previous example, there was a vertical asymptote at $x=3$, even though the number $3$ was not in the domain of the function $f$. 
Vertical asymptotes are defined in terms of limits. The important point is that we can \emph{approach} the number $3$ from within the domain of $f$.

\begin{example}
	Find the vertical asymptotes of 
	\[ g(x) = \frac{x^2\sin(x)}{e^{2x}\left(x^2-4x\right)}. \]

	\begin{explanation}
		This function looks a bit more complicated than in the previous example, but the work required is very similar. We'll start
		by factoring the numerator and denominator.
		\[ \frac{x^2\sin(x)}{e^{2x}\left(x^2-4x\right)} = \frac{x\cdot x\sin(x)}{e^{2x}x\left(x-4\right)}.\]
		
		The numerator $x^2\sin(x)$ and the denominator $e^{2x}x(x-4)$ are each continuous on $(-\infty, \infty)$. That means the only
		possibilities to check for vertical asymptotes are those numbers making the denominator zero.
		
		Using limits, we must investigate what happens with $g(x)$ when $x\to 0$ and $x\to 4$, since $0$ and $4$ are the only zeroes of the 				denominator (since $e^{2x}$ is never zero). We write
		\begin{align*}
			\lim_{x\to 0} \frac{\cancel{x}\cdot x\sin(x)}{\cancel{x}e^{2x}(x-4)} &= \lim_{x\to 0} \frac{x\sin(x)}{e^{2x}(x-4)}\\
				&= \frac{0}{1(-4)}\\
				&=0.
		\end{align*}
		Since $\displaystyle \lim_{x\to 0}g(x)=0$, there is no vertical asymptote at $x=0$. 
		
		We turn our attention to the limit as $x$ approaches $4$. Let us start with the limit from the right.		
		\[ \lim_{x\to 4^+} \frac{\cancel{x}\cdot x\sin(x)}{\cancel{x}e^{2x}(x-4)} = \lim_{x\to 4^+} \frac{x\sin(x)}{e^{2x}(x-4)} \]
		This is a one-sided limit with form \numOverZero. We will determine the sign of the fraction for $x$ sufficiently close to, 
		but greater than, $4$.
		
		For the numerator, notice that $\pi < 4 < \dfrac{3\pi}{2}$. That means $\sin(4)$ is the sine value of something in Quadrant III, 
		making it negative. That $x$ is near $4$ makes the $x$ factor in the numerator positive. The numerator is negative for all $x$ sufficiently
		close to, but greater than, $4$.
		  
		For the denominator, $e^{2x}$ is always positive. Since $x>4$, the factor $x-4$ is also positive. The denominator is positive for all 
		relevant values of $x$.
		
		That makes the fraction $\dfrac{x\sin(x)}{e^{2x}(x-4)}$ negative valued for all $x$ sufficiently close to, but greater than, $4$.
		This means $\displaystyle \lim_{x\to 4^+} g(x) = -\infty$.
		
		There is a vertical asymptote at $x=4$. There are no others, since the only other discontinuity is at $0$ but $\displaystyle \lim_{x\to 0}g(x)=0$.
		
		
	\end{explanation}

\end{example}

\end{document}
