\documentclass{ximera}

\input{../preamble.tex}

\title[Dig-In:]{Horizontal asymptotes}

\outcome{Find horizontal asymptotes using limits.}
\outcome{Recognize that a curve can cross a horizontal asymptote.}
\outcome{Calculate the limit as $x$ approaches $\pm \infty$ of common functions algebraically.}
\outcome{Find the limit as $x$ approaches $\pm \infty$ from a graph.}
\outcome{Define a horizontal asymptote.}
\outcome{Compute limits at infinity of famous functions.}
\outcome{Identify horizontal asymptotes by looking at a graph.}

\begin{document}
\begin{abstract}
We explore functions that behave like horizontal lines as the input
grows without bound.
\end{abstract}
\maketitle

Let's start with an example:
\begin{example}
Consider the function:
\[
g(x) = \frac{1}{x}
\]
with the table below
\[
\begin{array}{c|c}
x       & g(x)=\frac{1}{x} \\\hline
10      & \answer[given]{0.1} \\ 
100     & \answer[given]{0.01} \\
1000    & \answer[given]{0.001} \\
10000   & \answer[given]{0.0001} \\
100000  & \answer[given]{0.00001}%\\
%1000000 & \answer[given]{0.000001}
\end{array}
\]

  What does the table tell us about $g(x)$ as $x$ grows bigger and bigger?
  \begin{explanation}
    
As $x$ grows bigger and bigger, it seems that $g(x)$ approaches $0$. It seems that we should write
\[
 \lim_{x\to \infty}\frac{1}{x}=0 
 \]
  \end{explanation}
\end{example}

\begin{example}
  Now consider a different function:
  \[
  f(x) = \frac{6x-9}{x-1}
  \]
  \begin{image}
    \begin{tikzpicture}
      \begin{axis}[
          domain=1:4,
          ymax=20,
          ymin=-10,
          samples=100,
          axis lines =middle, xlabel=$x$, ylabel=$y$,
          every axis y label/.style={at=(current axis.above origin),anchor=south},
          every axis x label/.style={at=(current axis.right of origin),anchor=west}
        ]
	\addplot [very thick, penColor, smooth, domain=(0:.9)] {(6*x-9)/(x-1)};
        \addplot [very thick, penColor, smooth, domain=(1.1:3)] {(6*x-9)/(x-1)};
        \addplot [textColor, dashed] plot coordinates {(1,-10) (1,20)};
        \addplot [textColor, dashed] plot coordinates {(0,6) (3,6)};
      \end{axis}
    \end{tikzpicture}
%% \caption{A plot of $f(x)=\protect\frac{6x-9}{x-1}$.}
    %% \label{plot:(6x-9)/(x-1)}
  \end{image}
  Fill in the table below, rounding to $5$ decimal places.
  \[
  \begin{array}{c|c}
    x      & f(x)=\frac{6x-9}{x-1} \\ \hline
    10     & \answer[given]{5.66667}\\
    100    & \answer[given]{5.96970}\\
    1000   & \answer[given]{5.99700}\\
    10000  & \answer[given]{5.99970}\\
    100000 & \answer[given]{5.99997}\\
  \end{array}
  \]
  What does the table tell us about $f(x)$ as $x$ grows bigger and bigger?
  \begin{explanation}
    As $x$ grows bigger and bigger, it seems like $f(x)$ approaches $6$.
    It seems that we should write
    \[
    \lim_{x\to \infty}\frac{6x-9}{x-1}=6 
    \]
  \end{explanation}
\end{example}

\begin{definition}\label{def:limitAtInfty}
If $f(x)$ becomes arbitrarily close to a specific value $L$ by making
$x$ sufficiently large, we write
\[
\lim_{x\to \infty} f(x) = L
\]
and we say, the \dfn{limit at infinity} of $f(x)$ is $L$.  

If $f(x)$ becomes arbitrarily close to a specific value $L$ by making
$x$ sufficiently large and negative, we write
\[
\lim_{x\to -\infty} f(x) = L
\]
and we say, the \dfn{limit at negative infinity} of $f(x)$ is $L$.  
\end{definition}

\begin{example}  Compute the limit.
\[
\lim_{x\to\infty} \frac{6x-9}{x-1}.
\]
We will confirm what we guessed earlier by computing the limit at infinity.\\
 We can assume that all the Limit Laws also apply to limits at infinity.
\begin{explanation}
Write
\begin{align*}
\lim_{x\to\infty}\frac{6x-9}{x-1} &= \lim_{x\to\infty}\frac{6x-9}{x-1}\cdot \frac{1/x}{1/x}\\
&=\lim_{x\to\infty}\frac{\frac{6x}{x} - \frac{9}{x}}{\frac{x}{x} - \frac{1}{x}}\\
&=\frac{\lim_{x\to\infty}\Bigl(6- \frac{9}{x}\Bigr)}{\lim_{x\to\infty}\Bigl(1- \frac{1}{x}\Bigr)}\\
&=\frac{6-9\lim_{x\to\infty}  \frac{1}{x}}{1-\lim_{x\to\infty} \frac{1}{x}}\\
&=\frac{6-9(0)}{1-0}\\
&=  \frac{6}{1}\\
&= 6.
\end{align*}
\end{explanation}
\end{example}

Notice in the first step of this computation, we divided the numerator and denominator by $x$ (by multiplying them by $1/x$). This is a ``trick''
that frequently helps in the calculation of limits as $x\to \pm \infty$, when the function is a fraction. The important point of using $x$ here, is that the
denominator has degree $1$. When we divide the denominator by $x$, the denominator approaches a nonzero constant. 
(We're dividing the numerator and denominator by the \textbf{highest growth order from the denominator}.)

Sometimes one must be careful when attempting to apply this trick. Consider this example of a non-rational function.

\begin{example}

\[
\lim_{x\to -\infty} \frac{x^3+1}{\sqrt{x^6+5}}
\]
\begin{explanation}
Notice that the denominator is inside a radical. In order to determine the correct factor to divide by for our trick, we will need to rewrite the radical.

\begin{align*}
	\sqrt{x^6 + 5} &= \sqrt{ x^6\left(\dfrac{x^6}{x^6} + \dfrac{5}{x^6}\right)}\\
				&=	\sqrt{ x^6\left( 1 + \dfrac{5}{x^6}\right) }\\
				&= \sqrt{x^6} \sqrt{1 + \dfrac{5}{x^6} }\\
				&= \sqrt{(x^3)^2} \sqrt{1 + \dfrac{5}{x^6} }\\
				&= |x^3|\sqrt{1 + \dfrac{5}{x^6} }\\
\end{align*}
For this last step, remember that $\sqrt{a^2} = |a|$ when $a$ could possibly have a negative value. Since we're looking at $x\to -\infty$, 
we know that eventually $x$ will be negative and $x^3$ will be, too.

This simplification means the denominator will, in the limit as $x\to -\infty$, act like $-x^3\sqrt{1+\dfrac{5}{x^6}}$. Since the square root factor here
has limit $\displaystyle \lim_{x\to -\infty} \sqrt{1 + \dfrac{5}{x^6} } = \sqrt{1 + \lim_{x\to \infty} \dfrac{5}{x^6}} = \sqrt{1+0}=1$, the growth of the denominator is controlled by the $-x^3$ factor. In this case we divide the numerator and denominator by $x^3$,

\begin{align*}
\lim_{x\to -\infty} \dfrac{x^3+1}{\sqrt{x^6+5}} &= \lim_{x\to -\infty} \frac{x^3+1}{-x^3 \sqrt{1 + \dfrac{5}{x^6} }} \cdot \frac{1/x^3}{1/x^3}\\
	&= \lim_{x\to -\infty} \dfrac{1+\dfrac{1}{x^3}}{\dfrac{-x^3}{x^3}\cdot\sqrt{1+\dfrac{5}{x^6}}}\\
	&= \lim_{x\to -\infty} \dfrac{1+\dfrac{1}{x^3}}{-1\cdot \sqrt{1+\dfrac{5}{x^6}}} \\
	&= \dfrac{ \lim_{x\to -\infty}\left(1+\dfrac{1}{x^3}\right)}{\lim_{x\to -\infty}-1\cdot\sqrt{1+\dfrac{5}{x^6}}}\\
	&= \dfrac{1}{-1\cdot 1} = -1.
%	&= \frac{ -1-\lim_{x\to -\infty}(1/x)^3}{\sqrt{\lim_{x\to -\infty}\Bigl(1+5/x^6\Bigr)}}\\
%	&= \frac{ -1-(\lim_{x\to -\infty}1/x)^3}{\sqrt{1+5(\lim_{x\to -\infty}1/x)^6}}\\
%	&= \frac{ -1-(0)^3}{\sqrt{1+5(0)^6}}\\
%	&= \frac{ -1-0}{\sqrt{1+0}}\\
%	&= -1.
\end{align*}

\end{explanation}
\end{example}

In the previous example, we used the fact that $|x^3|= -x^3$ when $x$ is negative. If the limit were as $x\to \infty$, then $x$ and $x^3$ would each 
eventually be positive valued. This would have made $|x^3|=x^3$. Repeat the calculation from the previous example, finding the limit
$\displaystyle \lim_{x\to \infty} \frac{x^3+1}{\sqrt{x^6+5}}$ taking note to use $|x^3|=x^3$ in your simplification. 
You'll find that the two limits (as $x\to \infty$ and as $x \to -\infty$) have different values.






Notice that since
\[
\lim_{x\to \infty} f(x) = \lim_{x\to 0^+} f\left(\frac{1}{x}\right)
\]
and
\[
\lim_{x\to -\infty} f(x) = \lim_{x\to 0^-} f\left(\frac{1}{x}\right)
\]
we can also apply the Squeeze Theorem when taking limits at infinity.
Here is an example of a limit at infinity that uses the Squeeze
Theorem, and shows that functions can, in fact, cross their horizontal
asymptotes.


\begin{example}
Compute:
\[
\lim_{x\to \infty} \frac{\sin(7x)+4x}{x}
\]
\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            domain=2:20,
            ymax=5,
            ymin=3,
            samples=100,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west}
          ]
	  \addplot [very thick, penColor, smooth] {(1/x) * sin(deg(7*x))+4};
	     \addplot [textColor, dashed] plot coordinates {(0,4) (20,4)};
        \end{axis}
\end{tikzpicture}
%% \caption{A plot of $f(x)=\frac{\sin(7x)}{x}+4$.}
%% \label{plot:sin7x/x+4}
\end{image}

\begin{explanation}
We can bound our function
\[
\frac{-1+4x}{x} \le \frac{\sin(7x)+4x}{x} \le \frac{1+4x}{x}.
\]
Now write with me
\begin{align*}
\lim_{x\to\infty}\frac{-1+4x}{x} \cdot \frac{1/x}{1/x} &= \lim_{x\to\infty}\frac{-1/x+4}{1}\\
&=4.
\end{align*}
And we also have
\begin{align*}
\lim_{x\to\infty}\frac{1+4x}{x} \cdot \frac{1/x}{1/x} &= \lim_{x\to\infty}\frac{1/x+4}{1}\\
&=4.
\end{align*}
Since 
\[
\lim_{x\to \infty} \frac{-1+4x}{x}  = 4 = \lim_{x\to \infty}\frac{1+4x}{x} 
\] 
we conclude by the Squeeze Theorem,
$\lim_{x\to\infty}\frac{\sin(7x)}{x}+4 = 4$.
\end{explanation}
\end{example}






\begin{definition}\label{def:horiz asymptote}\index{asymptote!horizontal}\index{horizontal asymptote}
If  
\[
\lim_{x\to \infty} f(x) = L \qquad\text{or}\qquad \lim_{x\to -\infty} f(x) = L,
\]
then the line $y=L$ is a \dfn{horizontal asymptote} of $f(x)$.
\end{definition}

\begin{example} 
Give the horizontal asymptotes of
\[
f(x) = \frac{6x-9}{x-1}
\]
\begin{explanation}
From our previous work, we see that $\lim_{x\to \infty} f(x) = 6$, and
upon further inspection, we see that $\lim_{x\to -\infty} f(x) =
6$. Hence the horizontal asymptote of $f(x)$ is the line $y=6$.
\end{explanation}
\end{example}


It is a common misconception that a function cannot cross an
asymptote. As the next example shows, a function can cross a horizontal
asymptote, and in the example this occurs an infinite number of times!

\begin{example}
Give a horizontal asymptote of
\[
f(x) = \frac{\sin(7x)+4x}{x}.
\]
\begin{explanation}
Again from previous work, we see that $\lim_{x\to \infty} f(x) =
\answer[given]{4}$. Hence $y=\answer[given]{4}$ is a horizontal asymptote of $f(x)$.
\end{explanation}
\end{example}


We conclude with an infinite limit at infinity.

\begin{example}
Compute
\[
\lim_{x\to \infty} \ln(x)
\]
\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            domain=0:20,
            ymax=4,
            ymin=-1,
            samples=100,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west}
          ]
	  \addplot [very thick, penColor, smooth] {ln(x)};
        \end{axis}
\end{tikzpicture}
%% \caption{A plot of $f(x)=\ln(x)$.}
%% \label{plot:lnx}
\end{image}
\begin{explanation}
The function $\ln(x)$ grows very slowly, and seems like it may have a
horizontal asymptote, see the graph above. However, if we consider the
definition of the natural log as the inverse of the exponential
function
\begin{quote}%%BADBAD I would like word choice for "positive"
  $\ln(x) = y$ means that $e^y =x$ and that $x$ is positive.
\end{quote}
We see that we may raise $e$ to higher and higher values to obtain
larger numbers.  This means that $\ln(x)$ is unbounded, and hence
$\lim_{x\to\infty}\ln(x)=\infty$.
\end{explanation}
\end{example}


\end{document}
