\documentclass{ximera}

\input{../preamble.tex}

\title{Using limits to detect asymptotes}

\begin{document}

\begin{abstract}
%Stuff can go here later if we want!
\end{abstract}

\maketitle

\begin{sectionOutcomes}

After completing this section, students should be able to do the following.

\begin{itemize}
\item Recognize when a limit is indicating there is a vertical asymptote.
\item Evaluate the limit as $x$ approaches a point where there is a vertical asymptote.
\item Match graphs of functions with their equations based on vertical asymptotes.
\item Discuss what it means for a limit to equal $\infty$.
\item Define a vertical asymptote.
\item Find horizontal asymptotes using limits.
\item Produce a function with given asymptotic behavior.
\item Recognize that a curve can cross a horizontal asymptote.
\item Understand the relationship between limits and vertical asymptotes.
\item Calculate the limit as $x$ approaches $\pm \infty$ of common functions algebraically.
\item Find the limit as $x$ approaches $\pm \infty$ from a graph.
\item Define a horizontal asymptote.
%\item Define a slant asymptote.
%\item Approximate a slant asymptote from the graph of a
%%  function.
%\item Find slant asymptotes algebraically and graphically.
\item Compute limits at infinity of famous functions.
\item Find vertical asymptotes of famous functions.
\item Identify horizontal asymptotes by looking at a graph.
\item Identify vertical asymptotes by looking at a graph.
%\item Identify slant asymptotes by looking at a graph.
\end{itemize}
\end{sectionOutcomes}

\end{document}
