\documentclass{ximera}

\input{../preamble.tex}

\outcome{Compute limits of families of functions.} 
\outcome{Compute average velocity.}
\outcome{Approximate instantaneous velocity.}
\outcome{Compare average and instantaneous velocity.}
\outcome{Plot difference quotients for varying approximations of the instantaneous rate of change.}

\title[Dig-In:]{Instantaneous velocity}

\begin{document}
\begin{abstract}
We use limits to compute instantaneous velocity.
\end{abstract}
\maketitle

When one computes average velocity, we look at 
\[
\frac{\text{change in displacement}}{\text{change in time}}.
\]
To obtain the (instantaneous) velocity, we want the change in time to
``go to'' zero. By this point we should know that ``go to'' is a
buzz-word for a \textit{limit}. The change in time is often given as
an interval whose length goes to zero.  However, intervals must always
be written
\[
[a,b] \qquad\text{where $a < b$.}
\]
Given
\[
I = [a, a+h],
\]
we see that $h$ cannot be negative, or else it violates the notation
for intervals. Hence, if we want smaller, and smaller, intervals
around a point $a$, and we want $h$ to be able to be negative, we
write
\[
I_h = 
\begin{cases}
  [a+h,a]  & \text{if $h<0$}, \\ %% note this is MORE correct than std books
  [a,a+h]  & \text{if $0<h$}.     %% in the content section, we can explain this in detail
\end{cases}
\]
\begin{question}
  Let $a = 3$ and $h = 0.1$
  \[
  I_h = \left[\answer{3},\answer{3.1} \right]
  \]
  \begin{question}
    Let $a = 3$ and $h = -0.1$
  \[
  I_h = \left[\answer{2.9},\answer{3} \right]
  \]
  \end{question}
\end{question}

Regardless of the value of $h$, the average velocity on the interval
$I_h$ is computed by
\[
\frac{\text{change in displacement}}{\text{change in time}} =
\frac{s(t+h)-s(t)}{h}.
\]
We will be most interested in this ratio when $h$ approaches zero.
Let's put all of this together by working an example.

\begin{example}
A group of young mathematicians recently took a road trip from
Columbus Ohio to Urbana-Champaign Illinois. The position (west of
Columbus, Ohio) of van they drove in is roughly modeled by
\[
s(t) = 36t^2 - 4.8t^3 \qquad\text{(miles West of Columbus)} %% note the model is wrong
\]
on the interval $[0,5]$, where $t$ is measured in hours. What is the
average velocity on the interval $[0,5]$?

Additionally, let
\[
I_h= \begin{cases}
  [1+h,1]  & \text{if $h<0$}, \\ %% note this is MORE correct than std books
  [1,1+h]  & \text{if $0<h$}.     %% in the content section, we can explain this in detail
\end{cases}
\]
What is the average velocity on $I_h$ when $h= 0.1$?
What is the average velocity on $I_h$ when $h= -0.1$?
\begin{explanation}
  The average velocity on the interval $[0,5]$ is
  \begin{align*}
  \frac{s(5)-s(0)}{5-0} &= \frac{36\cdot 5^2 - 4.8\cdot 5^3 - (36\cdot 0^2 - 4.8\cdot 0^3)}{5}\\
  &=\frac{\answer[given]{300}}{5}\\
  &=\answer[given]{60}\qquad\text{miles per hour.}
  \end{align*}
  On the other hand, consider the interval
  \[
  I_h = 
  \begin{cases}
    [1+h,1]  & \text{if $h<0$}, \\ %% note this is MORE correct than std books
    [1,1+h]  & \text{if $0<h$}.     %% in the content section, we can explain this in detail
  \end{cases}
  \]
  When $h = 0.1$, the average velocity is
  \begin{align*}
    \frac{s(1+0.1)-s(1)}{0.1} &= \frac{36\cdot (1.1)^2 - 4.8\cdot (1.1)^3 - (36\cdot 1^2 - 4.8 \cdot 1^3)}{0.1}\\
    &=\frac{\answer[given]{5.9712}}{0.1}\\
    &=\answer[given]{59.712}\qquad\text{miles per hour.}
  \end{align*}
  
  On the other hand, when $h=-0.1$, the average velocity is
  \begin{align*}
    \frac{s(1-0.1)-s(1)}{-0.1} &= \frac{36\cdot (0.9)^2 - 4.8\cdot (0.9)^3 - (36\cdot 1^2 - 4.8 \cdot 1^3)}{-0.1}\\
    &=\frac{\answer[given]{-5.5392}}{-0.1}\\
    &=\answer[given]{55.392}\qquad\text{miles per hour.}
  \end{align*}
\end{explanation}
\end{example}

In our previous example, we computed \textit{average velocity} on
three different intervals. If we let the size of the interval go to
zero, we get \dfn{instantaneous velocity}. Limits will allow us to
compute instantaneous velocity.  Let's use the same setting as before.

\begin{example}
  The position of van (west of Columbus, Ohio) our young
  mathematicians drove to Urbana-Champaign, Illinois is roughly
  modeled by
  \[
  s(t) = 36t^2 - 4.8t^3 \qquad \text{for $0\le t\le 5$,}
  \] 
  Find a formula for the (instantaneous) velocity of this van.
\begin{explanation}
  Again, we are working with the interval
  \[
  I_h= \begin{cases}
    [t+h,t]  & \text{if $h<0$}, \\ %% note this is MORE correct than std books
    [t,t+h]  & \text{if $0<h$}.     %% in the content section, we can explain this in detail
  \end{cases}
  \]
  To compute the average velocity, we write
  \[
    \frac{s(t+h)-s(t)}{h}
  \]
  but this time, we will let $h$ go to zero.
  Write
  \[
    \lim_{h\to 0} \frac{s(t+h)-s(t)}{h} = \lim_{h\to 0} \frac{\answer[given]{36(t+h)^2 - 4.8(t+h)^3} - \left(36t^2 - 4.8t^3\right)}{h}
  \]
  Now expand the numerator of the fraction.
    \[
    = \lim_{h\to 0}\frac{36t^2 + 72 t h + \answer[given]{36h^2} - 4.8 t^3 - 14.4t^2h - 14.4 t h^2 - 4.8h^3 - 36t^2 + 4.8t^3}{h}
    \]
  Now combine like-terms.
  \[
  = \lim_{h\to 0}\frac{72 t h + 36h^2 - 14.4 t^2 h - 14.4th^2 - \answer[given]{4.8h^3}}{h}
  \]
  Factor an $h$ from every term in the numerator.
  \[
  = \lim_{h\to 0}\frac{h\left(\answer[given]{72t + 36h - 14.4t^2 - 14.4th - 4.8h^2}\right)}{h}
  \]
  Cancel $h$ from the numerator and denominator.
  \[
  = \lim_{h\to 0}\left(\answer[given]{72t + 36h - 14.4t^2 - 14.4th - 4.8h^2}\right)
  \]
  Plug in $h=0$:
  \[
  = \answer[given]{72t-14.4t^2}
  \]
  This gives us a formula for our instantaneous velocity, $v(t) =
  \answer[given]{72t-14.4t^2}$.  For your viewing enjoyment, check out graphs of both
  $y=s(t)$ and $y=v(t)$:
  \begin{image}
    \begin{tikzpicture}
      \begin{groupplot}[group style={group size=2 by 1,horizontal sep=2cm}, width=0.5\textwidth]
        \nextgroupplot[
          clip=false,
          domain=0:5,
          xmax=5,
        xmin=0,
        ymax=300,
        ymin=0,
        axis lines =middle, 
        y label style={at={(axis description cs:-0.1,0.5)},rotate=90,anchor=south},
        x label style={at={(axis description cs:0.5,-.2)}},
        xlabel=hours into road-trip, ylabel=distance traveled in miles,
        ]
        \addplot [very thick, penColor, smooth] {36*x^2 - 4.8*x^3};
          \node at (axis cs:3,230) {\color{penColor}$s(t)$}; 
          \nextgroupplot[
            clip=false,
            domain=0:5,
            xmax=5,
            xmin=0,
            ymax=100,
            ymin=0,
            axis lines =middle, 
            y label style={at={(axis description cs:-0.1,0.5)},rotate=90,anchor=south},
            x label style={at={(axis description cs:0.5,-.2)}},
            xlabel=hours into road-trip, ylabel=velocity in miles per hour]
          \addplot [very thick, penColor2, smooth] {72*x-14.4*x^2};
          \node at (axis cs:3,95) {\color{penColor2}$v(t)$};
      \end{groupplot}
    \end{tikzpicture}
  \end{image}
\end{explanation}
\end{example}
\end{document}
