\documentclass{ximera}

\input{../preamble.tex}

\outcome{}


\title[Break-Ground:]{Roxy and Yuri like food}

\begin{document}
\begin{abstract}
Here we see a dialogue where two young mathematicians discuss the eating
habits of their cats.
\end{abstract}
\maketitle

Check out this dialogue between two calculus students (based on a true
story):

\begin{itemize}
  \item[\textbf{Devyn}] Yo Riley, I was watching my two cats
    \textit{Roxy} and \textit{Yuri} eat their dry cat food last night.
\item[\textbf{Riley}] Cats love food!
\item[\textbf{Devyn}] I know! I noticed something kinda funny though:
  Both Roxy and Yuri start and finish eating at the same times; and
  while I gave Roxy a little more food than Yuri, less food was left
  in Roxy's bowl when they stopped eating.

  I wonder, is there is a point in time when Roxy and Yuri have the
  exact same amount of \textbf{dry cat food} in their bowls?
\item[\textbf{Riley}] Hmmmmm. Do Roxy and Yuri both start and finish
  drinking their water at the same times?  And does Roxy start with a
  little more water than Yuri, and finish with less water left than
  Yuri?
  \item[\textbf{Devyn}] Yes!
\item[\textbf{Riley}] Interesting. I wonder, is there is a point in
  time when Roxy and Yuri have the exact same amount of \textbf{water}
  in their bowls?
\end{itemize}

\begin{problem}
  Is there a time when Roxy and Yuri have the same amount of dry cat
  food in their bowls assuming:
  \begin{itemize}
  \item They start and finish eating at the same times.
  \item Roxy starts with more food than Yuri, and leaves less food uneaten than Yuri. 
  \end{itemize}
  \begin{multipleChoice}
    \choice{Yes.}
    \choice{No.}
    \choice[correct]{There is no way to tell.}
  \end{multipleChoice}
\end{problem}

\begin{problem}
  Is there a time when Roxy and Yuri have the same amount of water in
  their bowls assuming:
  \begin{itemize}
  \item They start and finish drinking at the same times.
  \item Roxy starts with more water than Yuri, and leaves less water
    left in her bowl than Yuri.
  \end{itemize}
    \begin{multipleChoice}
    \choice[correct]{Yes.}
    \choice{No.}
    \choice{There is no way to tell.}
  \end{multipleChoice}
\end{problem}



\begin{problem}
  Within the context of the two problems above, what is the difference
  between ``dry cat food'' and ``water?''
  \begin{freeResponse}
    Since the dry cat food is a collection of individual kibbles, the
    quantity left in the bowl, when modeled as a function of time is
    not continuous.

    However, the amount of water, when modeled as a function of time
    is continuous. 
  \end{freeResponse}
\end{problem}



\begin{xarmaBoost}
  Write down at least \textbf{five} questions for this lecture. After
  you have your questions, label them as ``Level 1,'' ``Level 2,'' or
  ``Level 3'' where:
\begin{description}
\item[Level 1] Means you know the answer, or know exactly how to do
  this problem.
\item[Level 2] Means you think you know how to do the problem.
\item[Level 3] Means you have no idea how to do the problem.
\end{description}
\begin{freeResponse}
\end{freeResponse}
\end{xarmaBoost}


\end{document}
