\documentclass{ximera}

\input{../preamble.tex}

\title{Inverses of functions}

\begin{document}
\begin{abstract}
	Functions are inverses if they compose to the identity function
\end{abstract}
\maketitle

\begin{question}
	Let $f(x) = x^3+1$. Let $a$ be the input into $f$ which generates an output of $28$.
	
	\begin{hint}
		 $f(a) = 28$
	\end{hint}
	\begin{hint}
		\begin{align*}
			a^3+1 &= 28\\
			a^3 &= 27\\
			a &= 3 
		\end{align*}
	\end{hint}
	  \[a = \answer{3}\] 
\end{question}

The situation in the question above is very common.  We often need to know which inputs for a function generate a given output.

\begin{question}
		Let $f(x) = x^3+1$ again. Let $a$ be the input into $f$ which generates an output of $b$.
	
	\begin{hint}
		 $f(a) = 28$
	\end{hint}
	\begin{hint}
		\begin{align*}
			a^3+1 &= b\\
			a^3 &= b-1\\
			a &= \sqrt[3]{b-1} 
		\end{align*}
	\end{hint}
	  \[a = \answer{(b-1)^(1/3)}\] 
\end{question}

In the last question, you found a way to find the input $a$ which generates any desired output $b$ for the function $f$.  This is what inverse functions are all about!

\begin{question}
	Let $f(x) = x^3+1$ and $g(x) = \sqrt[3]{x-1}$.  What is $f(g(x))$?  How about $g(f(x))$?
	
	\begin{question}
		\begin{hint}
			\begin{align*}
				f(g(x)) &= f(\sqrt[3]{x-1})\\
				&= (\sqrt[3]{x-1})^3+1\\
				&=(x-1)+1\\
				&=x
			\end{align*}
		\end{hint}
	\[f(g(x)) = \answer{x}\]
	\end{question}
	
	\begin{question}
	\begin{hint}
			\begin{align*}
				g(f(x)) &= g(x^3+1)\\
				&= \sqrt[3]{(x^3+1) - 1}\\
				&=\sqrt[3]{x^3}\\
				&=x
			\end{align*}
		\end{hint}
	\[g(f(x)) = \answer{x}\]
	\end{question}
\end{question}

In the example above, $f$ and $g$ exactly undo each other!  For example, $f(3) = 28$, and $g(28) = 3$.  So $g$ ``undoes" $f$, and $f$ also ``undoes'' $g$.  We say these functions are \textbf{inverses} of each other.

\begin{definition}
	Let $f$ be a function with domain $A$ and range $B$, and let $g$ be a function with domain $B$ and range $A$.  We say that $f$ and $g$ are \textbf{inverses} of each other if $f(g(b)) = b$ for all $b$ in $B$, and also $g(f(a)) = a$ for all $a$ in $A$. 
	
	Sometimes we write $g = f^{-1}$ in this case.  So we could rephrase the conditions as $f(f^{-1}(x)) = x$ and $f^{-1}(f(x)) = x$.  
\end{definition}

\begin{warning}
	\[f^{-1}(x) \neq \frac{1}{f(x)}\]
	
	This is an incredibly common mistake.  Admittedly, the notation is misleading, but that is the fault of tradition.  Get used to it!

\end{warning}

\begin{question}
	Suppose you know that $f(3) = 7$.
	
	\begin{hint}
		Intuitively, an inverse function should take an output and return the input.  So $f^{-1}(7) = 3$.  
		
		If we want to be more rigorous, we could work from the definition.  
		
		\[f^{-7}(7) = f^{-1}(f(3)) = 3\]
	\end{hint}
	
	\[f^{-1}(7) = \answer{3}\]
	
\end{question}

\begin{example}
	The formula for converting a temperature from Celsius to Fahrenheit is 
	
	\[F = \frac{9}{5} C + 32\]
	
	In the language of functions, we could say $f(t) = \frac{9}{5} t + 32$ is the function which takes Celsius measurements and converts them to Fahrenheit measurements of temperature.
	
	What if we had a Fahrenheit measurement, and wanted to find the corresponding Celsius measurement?  We would have to find the inverse function of $f$.   We can do so as follows:
	
	\begin{align*}
		f(f^{-1}(t)) &= t \text{ by the definition of inverse functions}\\
		\frac{9}{5} f^{-1}(t)+32 &= t \text{ by the  rule for $f$}\\
		\frac{9}{5} f^{-1}(t)&= t -32\\
		f^{-1}(t) = \frac{5}{9}(t - 32)
	\end{align*}
	
	So $f^{-1}(t) = \frac{5}{9}(t - 32)$ is the inverse function of $f$, which converts a Fahrenheit measurement back into a Celsius measurement.  For example,  $f(0) = 32$, and $f^{-1}(32) = 0$, which shows that $0$ Celsius  is $32$ Fahrenheight (and vica versa)!
\end{example}

\begin{question}
	Let $f(x) = 2x+7$.
	
	\begin{hint}
		$f(f^{-1}(x))=x$
	\end{hint}
	\begin{hint}
		\begin{align*}
			2f^{-1}(x)+7 &= x\\
			2f^{-1}(x) &= x-7\\
			f^{-1}(x) &= \frac{1}{2}(x-7)
		\end{align*}
	\end{hint}
	\[
	f^{-1}(x) = \answer{(1/2)(x-7)}
	\]
\end{question}

\begin{question}
	Let $f(x) = x^2$ with domain $\RR$ and $g(x = \sqrt{x})$ with domain $[0,\infty]$.  Are $f$ and $g$ inverses of each other?
	\begin{multipleChoice}
		\choice{Yes}
		\choice[correct]{No}
	\end{multipleChoice}
	\feedback{Note that $f$ has domain $\RR$ and range $[0,\infty]$, while $g$ has domain $[0,\infty]$ and range $[0,\infty]$.  Already this tells us that the functions are not inverses, since they do not conform to the domain and range criteria.  Even more devastatingly, although we have $f(g(x)) = (\sqrt{x})^2 = x$, we do not have $g(f(x)) = x$, since for example $g(f(-2)) = \sqrt{(-2)^2} = \sqrt{4} = 2 \neq -2$.
	
	If the domain of $f$ had been restricted to only the non-negative real numbers $[0,\infty]$, then $f$ would have been the inverse of $g$.
	}
\end{question}

\begin{question}
	Let $f$ be a function.  If the point $(1,9)$ is on the graph of $f$, what point must be the the graph of $f^{-1}$?
	
	\[\left( \answer{9}, \answer{1} \right)\]
	
	\feedback{Since $f(1) = 9$, we must have $f^{-1}(f(1)) = 1$, so $f^{-1}(9) = 1$.  Thus $(9,1)$ is on the graph of $f^{-1}$.  This is a general rule.  If $(a,b)$ is on the graph of $f$, and $(b,a)$ will be the graph of $f^{-1}$. }
\end{question}

\begin{question}
	Let $f$ be a function, and imagine that the points $(2,3)$ and $(7,3)$ are both on its graph.  Could $f$ have an inverse function?
	
	\begin{multipleChoice}
		\choice{Yes}
		\choice[correct]{No}
	\end{multipleChoice}
	
	\feedback{$f$ could \textbf{not} have an inverse function.  Imagine that it did.  Then $f^{-1}(f(2)) = 2$ and $f^{-1}(f(7)) = 7$.  Then we have both $f^{-1}(3) = 2$ and $f^{-1}(3) = 7$.  Since a \textbf{function} cannot send the same input to two different outputs, $f$ must not have an inverse function}
\end{question}

The last question highlights something interesting about inverse functions.  If two different inputs into a function have the same output, there is no hope  of that function having an inverse, since a function can only have one output.  This leads us to the following definition:

\begin{definition}
A function $f:A \to B$ is said to be one to one (a.k.a $1-1$, a.k.a. injective) if $f(a_1) = f(a_2)$ implies that $a_1 = a_2$.  In words, $f$ is $1-1$ if each output generated by a unique input.
\end{definition}

\begin{question}
Which of the following functions are one to one?  Select all that apply.

\begin{selectAll}
\choice[correct]{$f(x) = x$}
\choice{$f(x) = x^2$}
\choice{$f(x) = x^3 - 4x$}
\choice[correct]{$f(x) = x^3+4$}
\end{selectAll}

\end{question}

\begin{question}
\end{question}
\end{document}