\documentclass{ximera}

\input{../preamble.tex}

\outcome{Solve basic related rates word problems.}
\outcome{Understand the process of solving related rates problems.}
\outcome{Calculate derivatives of expressions with multiple variables implicitly.}

\title[Dig-In:]{More than one rate}

\begin{document}
\begin{abstract}
  Here we work abstract related rates problems.
\end{abstract}
\maketitle


Suppose we have two variables $x$ and $y$ which are both changing with
respect to time.  A \textit{related rates} problem is a problem where
we know one rate at a given instant, and wish to find the other.

Here the chain rule is key: If $y$ is written in terms of $x$, and we
are given $\dd[x]{t}$, then it is easy to find $\dd[y]{t}$ using the
chain rule:
\[
\dd[y]{t}=y'(x(t))\cdot x'(t).
\]
In many cases, particularly the interesting ones, our functions will
be related in some other way. Nevertheless, in each case we'll use the
power of the chain rule to help us find the desired rate. In this
section, we will work several abstract examples, so we can emphasize
the mathematical concepts involved. In each of the examples below, we will follow essentially the same plan of attack:

\begin{description}
\item[\textbf{Draw a picture.}] If possible, draw a schematic picture with all the relevant information. 
\item[\textbf{Find an equation.}] We want an equation that relates all
  relevant functions.
\item[\textbf{Differentiate the equation(s).}] Here we will often use
  implicit differentiation.
\item[\textbf{Evaluate the equation at the desired values.}] The known
  values should let you solve for the relevant rate.
\end{description}


\section{Formulas}

One way to combine several functions is with a known formula.

\begin{example}
  Imagine an expanding circle. If we know that the perimeteter is
  expanding at a rate of $4$ m/s, what rate is the area changing
  when the radius is $3$ meters?
  \begin{explanation}
    To start, we \textbf{draw a picture.}
    \begin{image}
      \begin{tikzpicture}
        \draw [penColor, very thick] (0,0) circle [radius=2];
        \draw [penColor] (0,0) -- (2,0);
        \node [below,penColor] at (1,0) {$r=3$ m};
        \node [penColor,left] at (-1.5,1.42) {$P'(t) = 4$ m/s};
        \node [penColor, right] at (1.5,-1.42) {$A = \pi\cdot r^2$};
      \end{tikzpicture}
    \end{image}
    We must \textbf{find an equation} that combines relevant
    functions. Here we use the common formulas for perimeter and area
    \[
    P = 2\cdot \pi \cdot r
    \qquad\text{and}\qquad
    A = \pi \cdot r^2.
    \]
    Next we imagine that $A$, $r$, and $P$ are functions of time
    \[
    P(t) = 2\cdot \pi \cdot r(t)
    \qquad\text{and}\qquad
    A(t) = \pi \cdot r(t)^2.
    \]
    and we \textbf{differentiate the equations} using implicit
    differentiation, treating all functions as functions of $t$
    \[
    P'(t) = 2\cdot \pi\cdot r'(t)
    \qquad\text{and}\qquad
    A'(t) = 2\cdot \pi\cdot r(t) \cdot r'(t).
    \]
    Now we \textbf{evaluate the equation at the desired values.} We
    know $P'(t) = 4$ and that $r(t) = 3$. Hence our equations become
    \[
    4 = 2\cdot \pi\cdot r'(t)
    \qquad\text{and}\qquad
    A'(t) = 2\cdot \pi\cdot 3 \cdot r'(t).
    \]
    We see that
    \begin{align*}
      4 &= 2\cdot \pi\cdot r'(t)\\
      2/\pi &= r'(t).
    \end{align*}
    and now that
    \begin{align*}
      A'(t) &= 2\cdot \pi\cdot 3 \cdot 2/\pi\\
      &=12.
    \end{align*}
    Hence the area is expanding at a rate of $12$ m/s.
  \end{explanation}
\end{example}


%%BADBAD
%% There are a number of common formulas that arise in related rates
%% problems.
%%
%% It might be nice to add a little list of common area/volume formulas.



\section{Right triangles}

A common way to combine functions is through facts related to right
triangles.


\begin{example}
  Imagine an expanding right triangle. If one leg has a fixed length
  of $3$ m, one leg is increasing with a rate of $2$ m/s, and the
  hypotenuse is expanding to accomidate the expanding leg, at what
  rate is the hypotenuse expanding when both legs are $3$ m long?
  \begin{explanation}
    To start, we \textbf{draw a picture.}
    \begin{image}
      \begin{tikzpicture}
        \coordinate (A) at (0,2);
        \coordinate (B) at (0,5);
        \coordinate (C) at (6.5,2);
        \tkzMarkRightAngle(C,A,B)
        \tkzDefMidPoint(A,B) \tkzGetPoint{a}
        \tkzDefMidPoint(A,C) \tkzGetPoint{b}
        \tkzDefMidPoint(B,C) \tkzGetPoint{c}
        \draw (A)--(B)--(C)--cycle;
        \tkzLabelPoints[above](c)
        \tkzLabelPoints[above](b)
        %\tkzLabelPoints[left](a)
        \node [left] at (a) {$a = 3$};
        \node [below] at (b) {$b'(t) = 2$};
      \end{tikzpicture}
    \end{image}

    We must \textbf{find an equation} that combines relevant
    functions. Here we use the Pythagorean Theorem.
    \[
    c^2 = a^2 + b^2
    \]
    Imagining $c$ and $b$ as being functions of time
    \[
    c(t)^2 = a^2 + b(t)^2
    \]
    we are now able to \textbf{differentiate the equation} using
    implicit differentiation, treating all functions as functions of
    $t$, note $a$ is constant,
    \[
    2\cdot c(t)\cdot c'(t) = 2\cdot b(t)\cdot b'(t).
    \]
    Now we \textbf{evaluate the equation at the desired values.} We
    know that $b'(t) = 2$ and that $b(t) = 3$
    \[
    2\cdot c(t)\cdot c'(t) = 12
    \]
    However, we still need to know $c(t)$ when $b(t) = 3$. Here we use
    the Pythagorean Theorem,
    \begin{align*}
    c(t)^2 &= 3^2 + 3^2\\
    &=18,
    \end{align*}
    and so we see that $c(t) = 3\sqrt{2}$. We may now write
    \begin{align*}
      6\sqrt{2}\cdot c'(t) &= 12 \\
      c'(t) &= \sqrt{2}.
    \end{align*}
    Hence $c(t)$ is growing at a rate of $\sqrt{2}$ m/s.
  \end{explanation}
\end{example}


\section{Angular rates}


We can also investigate problems involving angular rates.

\begin{example}
  Imagine an expanding right triangle. If one leg has a fixed length
  of $3$ m, one leg is increasing with a rate of $2$ m/s, and the
  hypotenuse is expanding to accomidate the expanding leg, at what
  rate is the angle attached to the fixed leg changing when both legs
  are $3$ m long?
  \begin{explanation}
    To start, we \textbf{draw a picture.}
    \begin{image}
      \begin{tikzpicture}
        \coordinate (A) at (0,2);
        \coordinate (B) at (0,5);
        \coordinate (C) at (6.5,2);
        \tkzMarkRightAngle(C,A,B)
        \tkzMarkAngle[size=1.2cm,thin](B,C,A)
        \tkzLabelAngle[pos = 1](B,C,A){$\theta$}
        \tkzDefMidPoint(A,B) \tkzGetPoint{a}
        \tkzDefMidPoint(A,C) \tkzGetPoint{b}
        \tkzDefMidPoint(B,C) \tkzGetPoint{c}
        \draw (A)--(B)--(C)--cycle;
        \tkzLabelPoints[above](c)
        \tkzLabelPoints[above](b)
        %\tkzLabelPoints[left](a)
        \node [left] at (a) {$a = 3$};
        \node [below] at (b) {$b'(t) = 2$};
      \end{tikzpicture}
    \end{image} 

    We must \textbf{find an equation} that combines relevant
    functions. Here we note that
    \[
    \tan(\theta) = \frac{a}{b}
    \]
    Imagining $\theta$ and $b$ as being functions of time
    \[
    \tan(\theta(t)) = \frac{a}{b(t)}
    \]
    we are now able to \textbf{differentiate the equation} using
    implicit differentiation, treating all functions as functions of
    $t$, note $a$ is constant,
    \[
    \sec^2(\theta(t))\theta'(t) = \frac{-a\cdot b'(t)}{b(t)^2}.
    \]
    Now we \textbf{evaluate the equation at the desired values.}  We
    know that $a=3$, $b'(t) = 2$, and that $b(t) = 3$
    \begin{align*}
    \sec^2(\theta(t))\cdot \theta'(t) &= \frac{-3\cdot 2}{3^2}\\
    &= \frac{-6}{9}\\
    &= \frac{-2}{3}.
    \end{align*}
    However, we still need to know $\sec^2(\theta)$. Here we use the
    Pythagorean Theorem,
    \begin{align*}
    c^2(t) &= 3^2 + 3^2\\
    &=18,
    \end{align*}
    and so we see that $c(t) = 3\sqrt{2}$. Now
    \begin{align*}
      \sec^2(\theta) &= \frac{\mathrm{hypotenuse}^2}{\mathrm{adjacent}^2}\\
      &= \frac{\left(3\sqrt{2}\right)^2}{3^2}\\
      &= 2.
    \end{align*}
    Hence
    \begin{align*}
      \sec^2(\theta(t))\cdot \theta'(t) &= \frac{-2}{3}\\
      2\cdot \theta'(t) &= \frac{-2}{3}\\
      \theta'(t) &= \frac{-1}{3}.
    \end{align*}
    So when $a=b=3$, the angle is changing at $-1/3$ radians per second.
  \end{explanation}
\end{example}



\section{Similar triangles}

Finally, facts about similar triangles are often useful when solving
related rates problems.

\begin{example}
  Imagine a triangle with ...
  \begin{image}
    \begin{tikzpicture}
      \coordinate (A) at (0,2);
      \coordinate (B) at (0,5);
      \coordinate (C) at (6,2);
      \coordinate (D) at (2,2);
      \coordinate (E) at (2,4);
      \tkzMarkRightAngle(C,A,B)
      \tkzDefMidPoint(A,B) \tkzGetPoint{a}
      \tkzDefMidPoint(A,C) \tkzGetPoint{b}
      \tkzDefMidPoint(B,C) \tkzGetPoint{c}
      \draw (A)--(B)--(C)--cycle;
      \tkzLabelPoints[above](c)
      \tkzLabelPoints[above](b)
      %\tkzLabelPoints[left](a)
      \node [left] at (a) {$a = 3$};
      \node [below] at (b) {$b'(t) = 2$};
    \end{tikzpicture}
  \end{image} 
  expanding right triangle. If one leg has a fixed length
  of $3$ m, one leg is increasing with a rate of $2$ m/s, and the
  hypotenuse is expanding to accomidate the expanding leg, at what
  rate is the hypotenuse expanding when both legs are $3$ m long? 
  \begin{explanation}
    To start, we \textbf{draw a picture.}
 

    We must \textbf{find an equation} that combines relevant
    functions.

    
    we are now able to \textbf{differentiate the equation} using
    implicit differentiation, treating all functions as functions of
    $t$,

   
    Now we \textbf{evaluate the equation at the desired values.}
    
  \end{explanation}
\end{example}



\begin{example} Water is poured into a conical container at the rate of 10
cm${}^3$/sec.  The cone points directly down, and it has a height of
30 cm and a base radius of 10 cm.  How fast is the water level rising
when the water is 4 cm deep?
\end{example}

\begin{solution}
To start, \textbf{draw a picture}.

\begin{tikzpicture}
\draw[penColor,very thick] (0,4) ellipse (4 and 1);
\draw[very thick,penColor!20!background] (2,2) arc (0:180:2 and .5);% top half of ellipse
\draw[very thick,penColor] (-2,2) arc (180:360:2 and .5);% bottom half of ellipse
\draw[penColor, very thick] (3.97,3.85) -- (0,0);
\draw[penColor, very thick] (-3.97,3.85) -- (0,0);
\draw[penColor, very thick] (0,4) -- (4,4);
\draw[penColor!50!background, very thick] (0,2) -- (2,2);
\draw[->,line width=0.4cm, penColor!20!background] (0,6) -- (0,4.25);
\draw[dashed, penColor2, very thick] (2.1,0) -- (2.1,2);
\draw[dashed, penColor, very thick] (-4.1,0) -- (-4.1,4);
\node[right, penColor] at (.4,5.6) {$\dd[V]{t} = 10$ cm$^3$/sec};
\node[below, penColor] at (2,4) {$10$ cm};
\node[above, penColor] at (1,2) {$r$ cm};
\node[right, penColor2] at (2.1,1) {$h(t) = 4$ cm};
\node[left, penColor] at (-4.1,2) {$30$ cm};
\end{tikzpicture}

Note, no attempt was made to draw this picture to scale, rather we
want all of the relevant information to be available to the
mathematician.

Now we need to \textbf{find an equation}. The formula for the volume of a cone tells us that 
\[
V = \frac{\pi}{3} r^2 h.
\]

Now we must \textbf{differentiate the equation}. We should use implicit differentiation, and treat each of the variables as functions of $t$. Write
\begin{equation}\label{equation:cone/water}
\dd[V]{t} = \frac{\pi}{3}\left(2rh \dd[r]{t} + r^2 \dd[h]{t}\right).
\end{equation}

At this point we \textbf{evaluate the equation at the desired values}.
At first something seems to be wrong, we do not know $\dd[r]{t}$.
However, the dimensions of the cone of water must have the same
proportions as those of the container.  That is, because of similar
triangles, 
\[
\frac{r}{h}=\frac{10}{30} \qquad\text{so}\qquad r={h/3}.
\]  
In particular, we see that when $h = 4$, $r=4/3$ and 
\[
\dd[r]{t} = \frac{1}{3}\cdot \dd[h]{t}.
\]
Now we can \textbf{evaluate the equation at the desired
  values}. Starting with Equation~\ref{equation:cone/water}, we plug
in $\dd[V]{t} = 10$, $r = 4/3$, $\dd[r]{t} = \frac{1}{3}\cdot \dd[h]{t}$
and $h=4$. Write
\begin{align*}
10 &= \frac{\pi}{3}\left(2\cdot \frac{4}{3}\cdot 4 \cdot\frac{1}{3}\cdot\dd[h]{t} + \left(\frac{4}{3}\right)^2 \dd[h]{t}\right)\\
10 &= \frac{\pi}{3}\left(\frac{32}{9}\dd[h]{t} + \frac{16}{9} \dd[h]{t}\right)\\
10 &= \frac{16\pi}{9}\dd[h]{t}\\
\frac{90}{16\pi} &= \dd[h]{t}.
\end{align*}
Thus, $\dd[h]{t}=\frac{90}{16\pi}$ cm/sec.
\end{solution}



\end{document}
