\documentclass{ximera}
\input{../../preamble.tex}
\author{Steven Gubkin}
% (su18:TK) answer prompt wording modified
\license{Creative Commons 3.0 By-NC}

\outcome{Use the first derivative to determine whether a function is increasing or decreasing.}
\outcome{Define higher order derivatives.}
\outcome{Compare differing notations for higher order derivatives.}
\outcome{Identify the relationships between the function and its first and second derivatives.}
\outcome{Sketch a graph of the second derivative, given the original function.}
\outcome{Sketch a graph of the original function, given the graph of its first and second derivatives.}
\outcome{Sketch a graph of a function satisfying certain constraints on its higher-order derivatives.}
\outcome{State the relationship between concavity and the second derivative.}
\outcome{Interpret the second derivative of a position function as acceleration.}
\outcome{Calculate higher order derivatives.}
\begin{document}

\begin{exercise}

  The image below shows the graph of $f$, $f'$, and $f''$.  Determine which graph corresponds to which function.

  \begin{image}
    \begin{tikzpicture}
      \begin{axis}[
        xmin=-2, xmax=3, ymin=-1,ymax=3,
        unit vector ratio*=1 1 1,
        axis lines =middle, xlabel=$x$, ylabel=$y$,
        every axis y label/.style={at=(current axis.above origin),anchor=south},
        every axis x label/.style={at=(current axis.right of origin),anchor=west},
        ticks=none,
        % xtick={-4,...,4}, ytick={-3,...,3},
        % xticklabels={-4,,-2,,0,,2,,4,,6}, yticklabels={,-2,,0,,2,,4,,6,,8,,10},
        % grid=major,width=4in,
        % grid style={dashed, gridColor},
        ]

        \addplot [very thick, penColor, samples=100,smooth, domain=(-6:6)] {x^2};
        \addplot [very thick, penColor2, samples=100,smooth, domain=(-6:6)] {2*x};
        \addplot [very thick, penColor3, samples=100,smooth, domain=(-6:6)] {2};

	\node at (axis cs: -1, 1 ) [penColor,anchor=east] {$B$};
        \node at (axis cs:-0.25, -0.5) [penColor2,anchor=east] {$A$};
        \node at (axis cs: 2, 1.8 ) [penColor3,anchor=west] {$C$};
      \end{axis}
    \end{tikzpicture}
  \end{image}

  % \begin{prompt}
  %   \begin{align*}
  %     f(x) &= \answer{B}\\
  %     f'(x) &= \answer{A}\\
  %     f''(x) &= \answer{C}\\
  %   \end{align*}
  % \end{prompt}

  % (su18:TK) modified phrasing
  \begin{prompt}
    \begin{enumerate}
    \item Graph $\answer[format=string]{B}$ corresponds to the graph of $f$.
    \item Graph $\answer[format=string]{A}$ corresponds to the graph of $f'$.
    \item Graph $\answer[format=string]{C}$ corresponds to the graph of $f''$.
    \end{enumerate}
  \end{prompt}

\end{exercise}

\end{document}
