\documentclass{ximera}

\input{../preamble.tex}

\outcome{Define a differential equation.}

\title[Break-Ground:]{Modeling the spread of infectious diseases}

\begin{document}
\begin{abstract}
  A dialogue where two young mathematicians discuss differential equations.
\end{abstract}
\maketitle

% Misconception: difference between equation versus function

% Find solutions to functional equations

% I'm thinking of a function whose derivative is three times that function.

% It must be zero!

% Well that wasn't what I was thinking of.

% e^(3x)

% Include more testimonials from people
Check out this dialogue between two calculus students (based on a true
story):

\begin{dialogue}
\item[Devyn] Riley, check out this book, it has some cool applications
  of calculus.
\begin{quote}
  The following differential equation can be used to model the spread
  of an infectious disease:
  \[
  \mathrm{infect}'(t) = k\cdot \mathrm{infect}(t)(P-\mathrm{infect}(t))
  \]
  Here $k$ is a constant, $\mathrm{infect}(t)$ is the number of people
  infected by the disease on day $t$, and $P$ is the size of the
  population vulnerable to the disease.
\end{quote}
\item[Riley] Woah. That's like a formula that not only has derivatives
  in it, but is itself a formula for a derivative. Wow. Much calculus. 
\item[Devyn] I wonder how you solve equations like this?
\item[Riley] I wonder if we can sometimes just use facts about the
  derivative to give us an approximation for the function that models the spread of infection?
\end{dialogue}



\begin{problem}
  Suppose your calculus class has had a freak outbreak of the
  \textit{math-philia}. Some facts: We have around $200$ students in
  our class, we are now on the $23$rd day of the outbreak, and currently $100$ students are infected. Using the differential
  equation
  \[
  \mathrm{infect}'(t) = k\cdot \mathrm{infect}(t)(P-\mathrm{infect}(t))
  \]
  we can model the spread of \textit{math-philia} by setting $k=0.001$.
  What is $\mathrm{infect}'(13)$?
  \begin{prompt}
    \[
    \mathrm{infect}'(13) = \answer{10}
    \]
  \end{prompt}
\end{problem}

\begin{problem}
  Do your best to explain why the equation
  \[
  \mathrm{infect}'(t) = k\cdot \mathrm{infect}(t)(P-\mathrm{infect}(t))
  \]
  is reasonable.
  \begin{hint}
    Don't worry, just do your best.
  \end{hint}
  \begin{freeResponse}
    Answers will vary.
  \end{freeResponse}
\end{problem}

%% \begin{enumerate}
%% \item Explain why the two formulas above are reasonable, though you
%%   don't need to worry about the precise values of the coefficients.
%% \item Discuss the outlook on day $13$ if you know
%%   $\mathrm{suscept}(13) = 5028$ and $\mathrm{infect}(13) = 4012$. In
%%   particular, you should talk about both $\mathrm{infect}'(x)$ and
%%   $\mathrm{suscept}'(x)$.
%% \item Compute $\mathrm{infect}''(x)$ on day 13. What does this say
%%   about the outlook?
%% \end{enumerate}




\input{../leveledQuestions.tex}

\end{document}
