\documentclass{ximera}

\input{../preamble.tex}

\outcome{Mechanical process of substitution.}


\title[Break-Ground:]{Working with substitution}

\begin{document}
\begin{abstract}
Two students discuss how to do integration by substitution.
\end{abstract}
\maketitle

Check out this dialogue between two calculus students (based on a true
story):

\begin{dialogue}
\item[Devyn] Yo Riley, is it just me, or are integrals kind of fun?
\item[Riley] I always feel accomplished when I finish one.
\item[Devyn] I know! Also, even though antiderivatives are difficult,
  we can always check our work by taking the derivative.
\item[Riley] So awesome!
\item[Devyn] But something is bothering me, when doing substitution,
  we have to find $f$ and $u$ such that
  \[
  \int f(g(x)) \cdot g'(x) \d x = \int f(u) \d u\qquad\text{where}\qquad u = g(x),
  \]
  but how do we choose $f$ and $g$?
\item[Riley] Well, never ever pick $g(x) = x$, this doesn't change
    anything!
\item[Devyn] And never ever pick $g(x)$ to be the entire integrand,
  this doesn't help either.
\item[Riley] Somehow we must ``see'' one function ``nested'' inside of
  another.
\item[Devyn] I'm not sure there's an easy path to doing, this, I think
  it's gonna take practice.
\end{dialogue}

%% Maybe we pick several problems and make the students identify f and
%% u based soley on u or f being given... This should help with the
%% skill of finding f and u.

In the problems that follow, we will be using the substitution formula
\[
   \int f(g(x)) \cdot g'(x) \d x = \int f(u) \d
   u\qquad\text{where}\qquad u = g(x).
\]
While you may be using a slightly different method, to compute your
integrals, the skills developed by answering the problems below will
help you in your quest to conquer calculus.

\begin{problem}
  Consider
  \[
  \int \sin^5(3x) \cos(3x) \d x = \int f(g(x)) \cdot g'(x) \d x
  \]
  if $u = g(x) = 3x$, and 
  \[
  \int f(g(x)) \cdot g'(x) \d x = \int f(u) \d
  u\qquad\text{where}\qquad u = g(x).
  \]
  what is $f(u)$?
  \begin{prompt}
    \[
    f(u) = \answer{
      \frac{\sin^5(u) \cos(u)}{3}}
    \]
  \end{prompt}
\end{problem}

\begin{problem}
  Consider
  \[
  \int \sin^5(3x) \cos(3x) \d x = \int f(g(x)) \cdot g'(x) \d x
  \]
  if $f(u) = \frac{u^5}{3}$, and 
  \[
  \int f(g(x)) \cdot g'(x) \d x = \int f(u) \d
  u\qquad\text{where}\qquad u = g(x).
  \]
  what is $u=g(x)$?
  \begin{prompt}
    \[
    u=g(x) = \answer{\sin(3 x)}
    \]
  \end{prompt}
\end{problem}

\input{../leveledQuestions.tex}


\end{document}
