\documentclass{ximera}

\input{../../preamble.tex}

\begin{document}
\outcome{Recognize and distinguish between secant and tangent lines.}
\outcome{Understand the definition of the derivative at a point.}
An oil tank is to be drained for cleaning. There are $V(t)$ gallons of oil left in the tank $t$ minutes after the draining began, where $V(t)=45(60-t)^2$.
\begin{exercise}
Find the \textbf{average rate} at which oil drains during the first $15$ minutes. 
\[
\answer{-4725}\unit{gal/m}
\]
\begin{exercise}
Find the \textbf{average rate} at which oil drains during the time interval $[10,15]$. 
\[
\answer{-4275}\unit{gal/m}
\]
\begin{exercise}
Find the \textbf{rate} at which oil is flowing out the tank $15$ minutes after the draining began.
\[
\answer{-4050}\unit{gal/m}
\]
\begin{exercise}
Find the \textbf{average rate}, $AR(t)$, at which oil drains during the time interval $[15+ t,15]$ if $-1< t<0$. 
\[
AR(t)=\answer{45(t-90)}\unit{gal/m}
\]
Now find the \textbf{average rate}, $AR( t)$, at which oil drains during the time interval $[15,15+ t]$ if $0< t<1$. 
\[
AR(t)=\answer{45(t-90)}\unit{gal/m}
\]
\begin{exercise}
Using your result, compute the limit
\[
\lim_{t\to 0}AR(t)=\answer{-4050}\unit{gal/m}
\]
\begin{exercise}
What does this limit $\lim_{t\to 0}AR(t)$ represent?
\begin{multipleChoice}
\choice{The average rate of change at $t=0$.}
\choice{The quantity $V(15)$.}
\choice[correct]{The instantaneous rate of change of $V(t)$ at $t=15$.}
\choice{The derivative of $V(t)$ at $t=0$.}
\end{multipleChoice}
\begin{exercise}
Suppose the oil tank has been completely drained and is now being refilled. Let $F(t)$ denote the gallons of oil in the tank $t$ minutes after the refilling process began. Suppose we are told that the function $F$ is a \textbf{differentiable function} and that the tank is so large that it will take a while to fill. 

Find an expression for the \textbf{average rate}, $\bar{r}(t)$, at which oil flows into the tank in the first $t$ minutes for $t>0$. Express $\bar{r}$ \textbf{symbolically} in terms of the function $F(t)$, the variable $t$ and the quantity $F(0)$.
\begin{multipleChoice}
\choice{$\bar{r}(t)=\frac{F(0)-F(t)}{t}\unit{gal/m}$}
\choice{$\bar{r}(t)=F(0)-F(t)\unit{gal/m}$}
\choice{$\bar{r}(t)=\frac{F(0)}{t}\unit{gal/m}$}
\choice{$\bar{r}(t)=\frac{F(t)}{t}\unit{gal/m}$}
\choice[correct]{$\bar{r}(t)=\frac{F(t)-F(0)}{t}\unit{gal/m}$}
\choice{$\bar{r}(t)=\frac{F(t)}{t}\unit{gal/m}$}
\end{multipleChoice}
%% \begin{hint}
%% While the parameters of the problem have changed, the crux of the exercise has not. Refer to your thinking in the first to come up with a solution. 
%% \end{hint}
\begin{exercise}
Now let $\bar{r}_{\Delta}(t,\tau)$ be the function denoting the \textbf{average rate} at which oil drains in the time interval $[t,\tau]$ where $t$ is greater than $0$ and less than tau\textemdash that is, where $0<t<\tau$. Express $\bar{r}_{\Delta}$ \textbf{symbolically} in terms of the function $F$ and the variables $t_0$ and $t_f$.
\begin{multipleChoice}
\choice{$\bar{r}(t,\tau)=\frac{F(t)-F(\tau)}{\tau-t}\unit{gal/m}$}
\choice{$\bar{r}(t,\tau)=F(t)-F(\tau)\unit{gal/m}$}
\choice{$\bar{r}(t,\tau)=\frac{F(t)}{\tau}\unit{gal/m}$}
\choice{$\bar{r}(t,\tau)=\frac{F(\tau)}{t}\unit{gal/m}$}
\choice[correct]{$\bar{r}(t,\tau)=\frac{F(\tau)-F(t)}{\tau-t}\unit{gal/m}$}
\choice{$\bar{r}(t,\tau)=\frac{F(\tau)}{\tau}\unit{gal/m}$}
\end{multipleChoice}
\begin{exercise}
Find the \textbf{rate} at which oil is flowing out the tank $15$ minutes after the draining began. Use ``prime'' notation (i.e., Lagrange's notation) for the derivative when entering your answer.
\[
\answer{V'(15)}\unit{gal/m}
\]
\end{exercise}
\end{exercise}
\end{exercise}
\end{exercise}
\end{exercise}
\end{exercise}
\end{exercise}
\end{exercise}
\end{exercise}
\end{document}
