\documentclass{ximera}

\input{../preamble.tex}

\outcome{Identify word problems as related rates problems.}
\outcome{Translate word problems into mathematical expressions.}


\title[Break-Ground:]{Pizza and calculus, so cheesy}

\begin{document}
\begin{abstract}
Two young mathematicians discuss tossing pizza dough. 
\end{abstract}
\maketitle

Check out this dialogue between two calculus students (based on a true
story):

%% Right Triangle
%% Oil Spill

\begin{dialogue}
\item[Devyn] Hey Riley, do you know what I love?
\item[Riley] Calculus? 
\item[Devyn] And pizza! Last night I made my own pizza crust from
  scratch!
\item[Riley] Mmmmmmm. Calculus. 
\item[Devyn] I know! The best part of making the crust is tossing it.
  But during my pizza tossing, I noticed something.  The
  dough is basically a \textbf{cylinder} with a \textbf{radius that is
    expanding}, and a \textbf{height that is getting smaller}. Then I started
    wondering: how are those two rates related?
\item[Riley] Pizza. So delicious. So cheesy. So calculus.
\end{dialogue}


The problem above is an example of a related rates word problem.  To solve 
it, we will need the tools in the Dig-In.  For now, let's see if we can reason 
through some questions related to this problem.



\begin{problem}
  In the context above, which of the following should be assumed to be
  functions that vary?
  \begin{selectAll}
  \choice[correct]{The radius of the cylinder.}
  \choice[correct]{The height of the cylinder.}
  \choice[correct]{The surface area of the cylinder.}
  \choice{The volume of the cylinder.}
  \end{selectAll}
\end{problem}

\begin{problem}
  In the context above, which of the following should be assumed to be
  constant?
  \begin{selectAll}
  \choice{The radius of the cylinder.}
  \choice{The height of the cylinder.}
  \choice{The surface area of the cylinder.}
  \choice[correct]{The volume of the cylinder.}
  \end{selectAll}
\end{problem}


%\input{../leveledQuestions.tex}


\end{document}
