\documentclass{ximera}

\outcome{Calculate limits of the form number over zero.}
\outcome{Identify determinate and indeterminate forms.}
\outcome{Distinguish between determinate and indeterminate forms.}

\input{../preamble.tex}

\title[Dig-In:]{Limits of the form nonzero over zero}

\begin{document}
\begin{abstract}
  What can be said about limits that have the form nonzero over zero?
\end{abstract}

\maketitle

Let's cut to the chase:

\begin{definition}
  A limit
  \[
  \lim_{x\to a} \frac{f(x)}{g(x)}
  \]
  is said to be of the form \numOverZero\ if
  \[
  \lim_{x\to a} f(x) = k\qquad\text{and}\qquad \lim_{x\to a} g(x) =0.
  \]
  where $k$ is some nonzero constant.
\end{definition}

\begin{question}
  Which of the following limits are of the form \numOverZero?
  \begin{selectAll}
    \choice[correct]{$\lim_{x\to -1} \frac{1}{(x+1)^2}$}
    \choice{$\lim_{x\to 2}\frac{x^2-3x+2}{x-2}$}
    \choice{$\lim_{x\to 0}\frac{\sin(x)}{x}$}
    \choice[correct]{$\lim_{x\to 2}\frac{x^2-3x-2}{x-2}$}
    \choice[correct]{$\lim_{x\to 1}\frac{e^x}{\ln(x)}$}
  \end{selectAll}
\end{question}

Let's see what is going on with limits of the
form \numOverZero.

\begin{example}
Consider the function
  \[
  f(x) = \frac{1}{(x+1)^2}.
  \]
  Use a table of values to investigate $\lim_{x\to -1} \frac{1}{(x+1)^2}$.
  \begin{explanation}
    Fill in the table below:
    \[
    \begin{array}{c|c|c}
      x & (x+1)^{2} & f(x) = \frac{1}{(x+1)^2}\\ \hline
      -1.1    & 0.01       & 100 \\
      -1.01   & 0.0001     & 10000 \\
      -1.001  & 0.000001   & \answer{1000000} \\
      -1.0001 & 0.00000001 & \answer{100000000} \\
      -0.9    & 0.01       & \answer{100} \\
      -0.99   & 0.0001     & \answer{10000} \\
      -0.999  & 0.000001   & \answer{1000000} \\
      -0.9999 & 0.00000001 & \answer{100000000} \\
    \end{array}
    \]
    What does the table tell us about 
    \[
    \lim_{x\to -1} \frac{1}{(x+1)^2}?
    \]
    It appears that the limit  does not exist, since the expression
    \[
    \frac{1}{(x+1)^2}
    \]
    becomes larger and larger as  $x$ approaches $-1$ . 
        Since
    \[
    \lim_{x\to -1} 1 = 1 \qquad\text{and}\qquad \lim_{x\to -1}(x+1)^2 = 0
    \]
	we have that
    \[
    \lim_{x\to -1} \frac{1}{(x+1)^2}\qquad\text{is of the form }\numOverZero
    \]

    Moreover, as $x$ approaches $-1$:
    \begin{itemize}
    \item The numerator is positive.
    \item The denominator approaches zero and is positive.
    \end{itemize}
    Hence, the expression
    \[
    \frac{1}{(x+1)^2}
    \]
    will become arbitrarily large as $x$ approaches $-1$.  We can see this
    in the graph of $f$.
    \begin{image}
      \begin{tikzpicture}
	\begin{axis}[
            domain=-2:1,
            ymax=100,
            samples=100,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west}
          ]
	  \addplot [very thick, penColor, smooth, domain=(-2:-1.1)] {1/(x+1)^2};
          \addplot [very thick, penColor, smooth, domain=(-.9:1)] {1/(x+1)^2};
          \addplot [textColor, dashed] plot coordinates {(-1,0) (-1,100)};
        \end{axis}
      \end{tikzpicture}
    \end{image}
  \end{explanation}
\end{example}

We are now ready for our next definition.

\begin{definition}
If $f(x)$ grows arbitrarily large as $x$ approaches $a$, we write
\[
	\lim_{x\to a} f(x) = \infty
\]
and say that the limit of $f(x)$ is  \dfn{infinity} as $x$
goes to $a$.\\


If $|f(x)|$ grows arbitrarily large as $x$ approaches $a$ and $f(x)$ is
negative, we write
\[
	\lim_{x\to a} f(x) = -\infty
\]
and say that the limit of $f(x)$ is  \textbf{ negative infinity}
as $x$ goes to $a$.
\end{definition}

Note: Saying "the limit is equal to infinity"  does not mean that the limit exists.  
It merely allows us to describe more precisely the behavior of the function $f$ near $a$, than saying "the limit does not exist".\\


Let's consider a few more examples.

\begin{example}
  Compute:
  \[
  \lim_{x\to -2} \frac{e^x}{(x+2)^4}
  \]
  \begin{explanation}
    First let's look at the form of this limit. We do this by taking the limits of both the numerator and denominator separately:
    \[
    \lim_{x\to -2} e^x = \answer[given]{\frac{1}{e^2}}\qquad\text{and}\lim_{x\to-2}\left((x+2)^4\right) = 0.
    \]
    So, this limit is of the form \numOverZero. This form is \textbf{determinate}, since it implies that  the limit does not exist.\\
    But, we can do better than that!
     As $x$ approaches $-2$:
    \begin{itemize}
    \item The numerator is a \wordChoice{\choice[correct]{positive}\choice{negative}} number. 
    \item The denominator is \wordChoice{\choice[correct]{positive}\choice{negative}} and is approaching zero.
    \end{itemize}
    This means that 
    \[
    \lim_{x\to -2} \frac{e^x}{(x+2)^4} = \infty.
    \]
  
  \end{explanation}
\end{example}
In this example, when we say ``As $x$ approaches $-2$'' we don't mean that we're checking a single value of $x$. We mean ALL values of $x$ that are close enough to $-2$. Checking a single value is not enough to justify whether the numerator will be positive or negative for all $x$ values sufficiently close to $-2$. Constructing a sign-chart or graphing the numerator could be helpful with this determination.

\begin{example}
  Compute:
  \[
  \lim_{x\to 3^+} \frac{x^2-9x+14}{x^2-5x+6}
  \]
  \begin{explanation}
    First let's look at the form of this limit, which we do by taking the limits of both the numerator and denominator separately.
    \[
    \lim_{x\to 3^+} \left(x^2-9x+14\right) = \answer[given]{-4}\qquad\text{and}\lim_{x\to3^+}\left(x^2-5x+6\right) = 0
    \]
    This limit is of the form \numOverZero. Next, we should factor the numerator and denominator to see if we can simplify the problem at all. 
    \begin{align*}
      \lim_{x\to 3^+}\frac{x^2-9x+14}{x^2-5x+6} &= \lim_{x\to 3^+}\frac{\cancel{(x-2)}(x-7)}{\cancel{(x-2)}(x-3)}\\
      &= \lim_{x\to 3^+}\frac{x-7}{x-3}
    \end{align*}
    Canceling a factor of $x-2$ in the numerator and denominator
    means we can more easily check the behavior of this limit.  As $x$
    approaches $3$ from the right:
    \begin{itemize}
    \item The numerator is a \wordChoice{\choice{positive}\choice[correct]{negative}} number. 
    \item The denominator is \wordChoice{\choice[correct]{positive}\choice{negative}} and approaching zero.
    \end{itemize}
    This means that
    \[
    \lim_{x\to 3^+} \frac{x^2-9x+14}{x^2-5x+6} = -\infty.
    \]
   \end{explanation}
\end{example}
Again, to determine if the numerator will be positive or negative for all $x$ sufficiently close to, but greater than, $3$, we cannot plug in a single value of $x$. (Even plugging in $x=3.0000001'$ would not be enough to justify our resullt.) We have to verify the sign for ALL values of $x$ that are sufficiently close to, but greater than $3$.

Here is our final example.

\begin{example}
  Compute:
  \[
  \lim_{x\to 3} \frac{x^2-9x+14}{x^2-5x+6}
  \]
  \begin{explanation}
    We've already considered part of this example, but now we consider the two-sided limit. We already know that
    \[
    \lim_{x\to 3} \frac{x^2-9x+14}{x^2-5x+6} = \lim_{x\to
      3}\frac{x-7}{x-3},
    \]
    and that this limit is of the form \numOverZero.
    We also know that as $x$ approaches $3$ from the right,
    \begin{itemize}
    \item The numerator is a negative number. 
    \item The denominator is positive and approaching zero.
    \end{itemize}
    Hence our function is approaching $-\infty$ from the right.
    
    As $x$ approaches $3$ from the left,
    \begin{itemize}
    \item The numerator is negative.
    \item The denominator is negative and approaching zero.
    \end{itemize}
    Hence our function is approaching $\infty$ from the left.
    This means
    \[
    \lim_{x\to 3} \frac{x^2-9x+14}{x^2-5x+6} = \answer[format=string,given]{DNE}.
    \]
    \begin{onlineOnly}
     We can confirm our results of the previous two examples by looking at the graph of $y=\frac{x^2-9x+14}{x^2-5x+6}$:
     \[
     \graph[xmin=-10,xmax=15,ymin=-10,ymax=10]{\frac{x^2-9x+14}{x^2-5x+6}}
     \]
   \end{onlineOnly}
  \end{explanation}
\end{example}

Some people worry that the mathematicians are passing into mysticism
when we talk about infinity and negative infinity. However, when we write
\[
\lim_{x\to a} f(x) = \infty \qquad\text{and}\qquad \lim_{x\to a} g(x) = -\infty
\]
all we mean is that as $x$ approaches $a$, $f(x)$ becomes arbitrarily
large and $|g(x)|$ becomes arbitrarily large, with $g(x)$ taking
negative values. We're using $\pm \infty$ to indicate directions, rather than a specific value.
\end{document}
