\documentclass{ximera}

\input{../preamble.tex}

\outcome{Understand what is meant by the form of a limit.}
\outcome{Calculate limits of the form zero over zero.}
\outcome{Identify determinate and indeterminate forms.}
\outcome{Distinguish between determinate and indeterminate forms.}
\outcome{Discuss why infinity is not a number.}

\title[Dig-In:]{Limits of the form zero over zero}

\begin{document}
\begin{abstract}
  We want to evaluate limits for which the Limit Laws do not apply. 
\end{abstract}

\maketitle

In the last section we computed  limits 
using continuity and the limit laws. What about limits that cannot be
directly computed using these methods?  Consider the following limit,
\[
\lim_{x\to 2}\frac{x^2-3x+2}{x-2}.
\]
Here 
\[
\lim_{x\to 2}\left(x^2-3x+2\right) = 0\qquad\text{and}\qquad \lim_{x\to
  2}\left(x-2\right) = 0
\]
in light of this, you may think that the limit is one or
zero. \textbf{Not so fast}. This limit is of an \textit{indeterminate
  form}. What does this mean? Read on, young mathematician.

\begin{definition}
  A limit
  \[
  \lim_{x\to a} \frac{f(x)}{g(x)}
  \]
  is said to be of the form \zeroOverZero\ if
  \[
  \lim_{x\to a} f(x) = 0\quad\text{and}\quad \lim_{x\to a} g(x) =0.
  \]
\end{definition}

\begin{question}
  Which of the following limits are of the form \zeroOverZero?
  \begin{selectAll}
    \choice[correct]{$\lim_{x\to 0}\frac{\sin(x)}{x}$}
    \choice{$\lim_{x\to 0}\frac{\cos(x)}{x}$}
    \choice{$\lim_{x\to 0}\frac{x^2-3x+2}{x-2}$}
    \choice[correct]{$\lim_{x\to 2}\frac{x^2-3x+2}{x-2}$}
    \choice{$\lim_{x\to 3}\frac{x^2-3x+2}{x-3}$}
  \end{selectAll}
\end{question}

\begin{warning}
  The symbol \zeroOverZero\ is \textbf{not} the number $0$ divided by
  $0$. It is simply a short-hand notation.  It means the limit $\lim_{x\to a} \frac{f(x)}{g(x)}$ has the property that
  \[
  \lim_{x\to a} f(x) = 0\quad\text{and}\quad \lim_{x\to a} g(x) =0.
  \]
  When writing a form of a limit, we NEVER write $\lim_{x \to a}\dfrac{f(x)}{g(x)}=\frac{0}{0}$.  It is the $=$ that makes this a mathematically incorrect statement.  Instead, write $\lim_{x \to a}\dfrac{f(x)}{g(x)}$ is of the form $\frac{0}{0}$
\end{warning}

To find the form of $\lim_{x \to a}\dfrac{f(x)}{g(x)}$, take the limits of the numerator and denominator SEPARATELY. Do not just ``plug in $x=a$''. ``Plugging in'' can not occur until after the function is noticed as continuous.

Let's finish the example with the function above.

\begin{example}
  Compute:
  \[
  \lim_{x\to 2}\frac{x^2-3x+2}{x-2}
  \]
  \begin{explanation}
	To find the form, notice that $\lim_{x\to 2}(x^2-3x+2)=0$ and $\lim_{x\to 2}(x-2)=0$. This means 
  	our limit has the form \zeroOverZero. 
  	However, note that if we assume $x\ne 2$, then we can write
    \[
    \frac{x^2-3x+2}{x-2} = \frac{(x-2)(\answer[given]{x-1})}{(x-2)} = \answer[given]{x-1}.
    \]
  
    \begin{image}
      \begin{tikzpicture}
        \begin{axis}[
            domain=-2:4,
            width=2.5in,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={-2,...,4},
            ytick={-3,...,3},
          ]
	  \addplot [very thick, penColor, smooth] {x-1};
        \end{axis}
        \node [penColor] at (2,-.75) {$y= x-1$};
      \end{tikzpicture}
      \quad
      \begin{tikzpicture}
	\begin{axis}[
            domain=-2:4,
            width=2.5in,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={-2,...,4},
            ytick={-3,...,3},
          ]
	  \addplot [very thick, penColor, smooth] {x-1};
          \addplot[color=penColor,fill=background,only marks,mark=*] coordinates{(2,1)};  %% open hole
        \end{axis}
        \node [penColor] at (2,-.5) {$y=\frac{x^2-3x+2}{x-2}$};
      \end{tikzpicture}
    \end{image}
      So, ``our function" is equal to the polynomial $x-1$ everywhere, except at $x=2$. Therefore, for all values of $x$ near $2$, ``our function" is equal to the polynomial $x-1$!
    This means that
    \[
    \lim_{x\to 2}\frac{x^2-3x+2}{x-2} = \lim_{x\to 2} (x-1).
    \]
    But now, we have to take the limit of a polynomial,
    $\lim_{x\to 2} (x-1) =\answer[given]{1}$.
    
     Hence
    \[
    \lim_{x\to 2}\frac{x^2-3x+2}{x-2} = \answer[given]{1}.
    \]
  \end{explanation}
\end{example} 


Let's consider a few more examples of the form \zeroOverZero.

\begin{example}
  Compute:
  \[
  \lim_{x\to1}\frac{x-1}{x^2+2x-3}.
  \]
  \begin{explanation}
    First note that
    \[
    \lim_{x\to1}\left(x-1\right)=0 \quad\text{and}\quad  \lim_{x\to1}\left(x^2+2x-3\right) = 0.
    \]
    Hence, this limit is of the form \zeroOverZero. Again, we cannot apply the Quotient Law or any other Limit Law. We cannot use continuity, either. Namely, ``our function" is not continuous at $x=1$, since it is not defined 
    at $x=1$. 
    
    What can be done? We can hope to be able to  cancel a factor going to $0$ out of the numerator and
    denominator.  Since $\answer[given]{(x-1)}$ is a factor going to $0$ in the
    numerator, let's see if we can factor a $\answer[given]{(x-1)}$ out of the
    denominator as well.
    \begin{align*}
      \lim_{x\to1}\frac{x-1}{x^2+2x-3}&=\lim_{x\to1}\frac{\cancel{x-1}}{\cancel{(x-1)}\answer[given]{(x+3)}} \\
      &=\lim_{x\to1}\frac{1}{\answer[given]{x+3}}\\
      &=\frac{1}{4}.
    \end{align*}
  \end{explanation}
\end{example}


\begin{example}
  Compute:
  \[
  \lim_{x\to 1} \frac{\frac{1}{x+1}-\frac{3}{x+5}}{x-1}.
  \]
\begin{explanation}
  We find the form of this limit by looking at the limits of the
  numerator and denominator separately. The limit of the numerator is:
  \begin{align*}
    \lim_{x\to 1}\left(\frac{1}{x+1}-\frac{3}{x+5}\right)&=  \lim_{x\to 1}\frac{x+5-3(x+1)}{(x+1)(x+5)}\\
    &= \lim_{x\to 1}\frac{-2x+2}{(x+1)(x+5)}\\
    &= \frac{0}{12}\\
    &=0
  \end{align*}
  The limit of the denominator is:
  \[
  \lim_{x\to 1}\left(x-1\right)=0
  \]
  Our limit is therefore of the form \zeroOverZero\ and we can
  probably factor a term going to $0$ out of both the numerator and
  denominator.
  \[
  \lim_{x\to 1} \frac{\frac{1}{x+1}-\frac{3}{x+5}}{x-1}
  \]
  When looking at the denominator, we hope that this
  term is $(x-1)$.  Unfortunately, it is not immediately obvious how to
  factor an $(x-1)$ out of the numerator. So, we should first simplify the complex fraction
  by multiplying it by
  \[
  1 = \frac{(x+1)(x+5)}{(x+1)(x+5)}
  \]
  this will allow us to cancel immediately
\begin{align*}
  \lim_{x\to 1}& \frac{\frac{1}{x+1}-\frac{3}{x+5}}{x-1}  \cdot \frac{(x+1)(x+5)}{(x+1)(x+5)} \\
  &= \lim_{x\to 1}\frac{(x+5)-3(x+1)}{(x+1)(x+5)(x-1)}.
\end{align*}

Now we will multiply out the numerator.  Note that we do not want to
multiply out the denominator because we already have an $(x-1)$
factored out of the denominator and that was the goal.
\[
\lim_{x\to 1}\frac{(x+5)-3(x+1)}{(x+1)(x+5)(x-1)}
\]
\begin{align*}
  &= \lim_{x\to 1}\frac{x+5-3x-3}{(x+1)(x+5)(x-1)} \\
  &= \lim_{x\to 1}\frac{-2x+2}{(x+1)(x+5)(x-1)}\\
  &= \lim_{x\to 1}\frac{-2\cancel{(x-1)}}{(x+1)(x+5)\cancel{(x-1)}}\\
  &= \lim_{x\to 1}\frac{-2}{(x+1)(x+5)}.
\end{align*}

Now, we can see that the limit of ``our function" is equal to the limit of a rational function $\frac{-2}{(x+1)(x+5)}$. This rational function is continuous on its domain, and, therefore, at $x=1$.  Hence
\begin{align*}
\lim_{x\to 1} \frac{\frac{1}{x+1}-\frac{3}{x+5}}{x-1}&=\lim_{x\to
  1}\frac{-2}{(x+1)(x+5)}\\
&= \frac{-2}{(1+1)(1+5)} \\
&=\answer[given]{\frac{-1}{6}}.
\end{align*}
\end{explanation}
\end{example}

We will  look at one more example.

\begin{example}
  Compute:
  \[
  \lim_{x\to-1} \frac{\sqrt{x+5}-2}{x+1}.
  \]

\begin{explanation} 
  Note that 
  \[
  \lim_{x\to-1} \left(\sqrt{x+5}-2\right)=0\quad\text{and}\quad\lim_{x\to -1} \left(x+1\right) =0.
  \]
  Our limit is, again, of the form \zeroOverZero\ and we
  can probably factor a term going to $0$ out of both the numerator
  and denominator.  We suspect from looking at the denominator that
  this term is $(x+1)$.  Unfortunately, it is not immediately obvious
  how to factor an $(x+1)$ out of the numerator.
 
  We will use an algebraic technique called \dfn{multiplying by the
    conjugate}.  This technique is useful when you are trying to
  simplify an expression that looks like
  \[
  \sqrt{\text{something}} \pm \text{something else}.
  \]
  It takes advantage of the difference of squares rule 
  \[
  a^2-b^2=(a-b)(a+b).
  \]
  In our case, we will use $a=\sqrt{x+5}$ and $b=2$.  Write
  \[
  \lim_{x\to-1} \frac{\sqrt{x+5}-2}{x+1}
  \]
  \begin{align*}
    &= \lim_{x\to-1} \frac{\left(\sqrt{x+5}-2\right)}{(x+1)} \cdot \frac{\left(\sqrt{x+5}+2\right)}{\left(\sqrt{x+5}+2\right)} \\
&=\lim_{x\to-1} \frac{\answer[given]{\left(\sqrt{x+5}\right)^2-2^2}}{(x+1)\left(\sqrt{x+5}+2\right)} \\
&=\lim_{x\to-1} \frac{x+5-4}{(x+1)\left(\sqrt{x+5}+2\right)} \\
&=\lim_{x\to-1} \frac{\cancel{(x+1)}}{\cancel{(x+1)}\left(\sqrt{x+5}+2\right)} \\
&=\lim_{x\to-1} \frac{1}{\sqrt{x+5}+2}\\
&= \frac{1}{\sqrt{-1+5}+2}\\
&=\answer[given]{\frac{1}{4}}.
\end{align*}
\end{explanation}
\end{example}

All of the examples in this section are limits of the form \zeroOverZero.
When you come across a limit of the form \zeroOverZero, you should try
to use algebraic techniques to come up with a continuous
function whose limit you can evaluate.

Notice that we solved multiple examples of limits of the form
\zeroOverZero\ and we got different answers each time.  This tells us
that just knowing that the form of the limit is \zeroOverZero\ is not enough
to compute the limit. The moral of the story is
\begin{center}
  \textbf{Limits of the form \zeroOverZero\ can take any value.}
\end{center}

\begin{definition}
A form that gives us no information about whether the limit exists or not, and if the limit exists, no information about the value of the limit, is
called an \dfn{indeterminate form}.

A form that gives information about whether the limit exists or not, and if it exists gives information about the value of the limit, is called a
\dfn{determinate form}.
\end{definition}  

Finally, you may find it distressing that we introduced a form, namely
\zeroOverZero, only to end up saying they give no information on the
value of the limit. But this is precisely what makes
indeterminate forms interesting\dots~they're a mystery!



\end{document}
