\documentclass{ximera}

\input{../../preamble.tex}

\author{Jim Talamo}
\license{Creative Commons 3.0 By-bC}


\outcome{}


\begin{document}
\begin{exercise}
Consider the series $\sum_{k=0}^{\infty} \frac{5}{k^2+4}$.  Which of the following should be done to check whether the series converges or diverges?
\begin{multipleChoice}
\choice{Recognize this as a geometric series then apply results for convergence of geometric series.}
\choice{Recognize this as a $p$-series then apply results for convergence of geometric series.}
\choice{Recognize that $\lim_{n \to \infty} \frac{5}{n^2+4} = 0$, hence the series converges.}
\choice[correct]{Apply the integral test.}
\end{multipleChoice}

\begin{exercise}
The integral test applies here since by letting $f(x) = \frac{5}{x^2+4}$:
\begin{selectAll}
\choice[correct]{$f(x)$ is decreasing.}
\choice[correct]{$f(x) \geq 0$ for all $x \geq 0$.}
\choice[correct]{$f(x)$ is continuous for $x \geq 0$.}
\end{selectAll}

To use the integral test, we must determine if the improper integral:

\[
\int_0^{\infty} \frac{5}{x^2+4} \d x
\]
converges or diverges.

Thus, we must analyze $\lim_{b \to \infty} \left[\int_0^b \frac{5}{x^2+4} \d x\right]$.  We find:

\[
\int_0^b \frac{2x}{x^2+4} \d x = \answer{\frac{5}{2} \arctan\left(\frac{b}{2}\right)}
\]

\begin{exercise}
We can now compute the limit:

\[ \lim_{b \to \infty} \left[\int_1^b \frac{5}{x^2+4} \d x\right] = \answer{\frac{5}{4} \pi} \]
(Use $\infty$ or $-\infty$ where appropriate or write ``DNE" if the limit otherwise does not exist)

Hence, the series  $\sum_{k=1}^{\infty} \frac{5}{k^2+4}$:
\begin{multipleChoice}
\choice[correct]{converges by the integral test.}
\choice{converges to $\frac{5}{4}\pi$ by the integral test.}
\choice{diverges by the integral test.}
\end{multipleChoice}
\end{exercise}
\end{exercise}
\end{exercise}

\end{document}
