\documentclass{ximera}

\input{../../preamble.tex}

\author{Jim Talamo}
\license{Creative Commons 3.0 By-bC}


\outcome{}


\begin{document}
\begin{exercise}
Consider the series $\sum_{k=1}^{\infty} \frac{k!}{3^{k^2}}$.  Note that the $k^2$ is in the exponent, that is $3^{k^2} = 3^{(k^2)}$.

\begin{multipleChoice}
\choice[correct]{The Ratio Test applies to this series.}
\choice{The Ratio Test does not apply to this series.}
\end{multipleChoice}

Why is the Ratio Test preferable to the Root Test here?
\begin{multipleChoice}
\choice[correct]{The Root Test would require that we evaluate $\lim_{n \to \infty} \sqrt[n]{n!}$}
\choice{The assumptions for the Root Test are not met.}
\end{multipleChoice}

To use the Ratio Test, note that the sequence $\{a_n\}_{n=1}$ is given by the rule:

\[
a_n = \answer{\frac{n!}{3^{n^2}}}
\]

Thus, to use the Ratio Test, we must compute the limit:

\[
L = \lim_{n \to \infty} \answer{\frac{(n+1)!}{3^{(n+1)^2}}  \frac{3^{n^2}}{n!}}
\]

Evaluating this limit, we find:

\[
L = \answer{0}
\]

\begin{hint}
\begin{question}
We can rearrange this limit to group like terms:

\[
L = \lim_{n \to \infty} \frac{3^{n^2}}{3^{(n+1)^2}} \cdot  \frac{(n+1)!}{n!}
\]

Simplifying, we find:

\begin{align*}
 \frac{3^{n^2}}{3^{(n+1)^2}} &=  \frac{1}{3^{\answer{2n+1}}} \\
 \frac{(n+1)!}{n!} &= \answer{n+1}
\end{align*} 

\begin{question}
Hence, 
\[
L = \lim_{n \to \infty} \frac{n+1}{3^{2n+1}} = \answer{0} \textrm{ by growth rate results.}
\]
\end{question}
\end{question}
\end{hint}

Hence, the Ratio Test:
\begin{multipleChoice}
\choice[correct]{guarantees that the series converges.}
\choice{guarantees that the series converges to $0$.}
\choice{guarantees that the series diverges.}
\choice{is inconclusive.}
\end{multipleChoice}

\begin{exercise}
By growth rates, we have that $\lim_{n \to \infty} \frac{n!}{3^n} = \answer{\infty}$.

(Use $\infty$ if appropriate or write ``DNE if the limit does not exist otherwise)

However, since the \emph{series} $\sum_{k=1}^{\infty} \frac{k!}{3^{k^2}}$ converges, we can determine the limit of the \emph{sequence} whose terms we are summing and find:

\[
\lim_{n \to \infty} \frac{n!}{3^{n^2}} = \answer{0}
\]

\end{exercise}

\end{exercise}
\end{document}
