\documentclass{ximera}

\input{../preamble.tex}

\outcome{}

\title[Break-Ground:]{Idea of substitution}

\begin{document}
\begin{abstract}
Two students consider substitution geometrically.
\end{abstract}
\maketitle

% Start story with definite integral

% Something with constant jacobian like $e^{3x}$

% Do this calculation by thinking about definite integrals.

% How would you ever know that substitution worked for definite integrals without the fundamental theorem?

% Relate antiderivative and fundamental theorem of calculus

% Why are limits changing?

% What happens when the substituted function is not a bijection?

% Break down this into a "bonus" section?

\begin{dialogue}
\item[Devyn] You know, Riley, we should be able to figure some integrals geometrically using transformations of functions.
\item[Riley] That sounds like a cool idea.  One example would be, since we know the graph of $f(x) = \sqrt{1-x^2}$ is a semicircle, we get an ellipse defined on $[-2,2]$ just by stretching the graph of $f$ by a factor of $2$ horizontally.  The equation of this ellipse would be $g(x) = \answer{\sqrt{1-(\frac{x}{2})^2}}$
\item[Devyn] Exactly!  So since we know that $\int_{-1}^1 \sqrt{1-x^2} dx = \answer{\frac{\pi}{2}}$ geometrically...
\item[Riley] and we know that the area under $g$ from $[-2,2]$ is twice the under $f$...
\item[Devyn and Riley] we must have $\int_{-2}^2 \sqrt{1-(x/2)^2} dx =  \answer{\pi}$ !
\item[Devyn and Riley] Jinx!
\item[Devyn] It is kind of like we just stretched out our whole coordinate system, and that helped us solve an integral.
\item[Riley] In this case, everything got stretched out by a constant factor of $2$ in the horizontal direction.  I wonder if we could ever say anything useful about cases where we stretch the $x$ axis by a different amount at each point?
\item[Devyn] Whao, that is a wild thought.  That seems really hard.  Since derivatives measure how much a function stretches a little piece of the domain, maybe the derivative will come into play here?
\item[Riley] Ya, but I do not see exactly how.  Maybe we should ask our TA about this?
\end{dialogue}

\begin{question}
Say we know that $\int_1^4 f(x) dx = 5$.  Then, using this transformation idea, we can evaluate $\int_a^b f(3x+1)$ if $a= \answer{0}$ and $b=\answer{1}$.  The value of the integral on this interval is $\answer{5/3}$.
\end{question}

\input{../leveledQuestions.tex}


\end{document}
