\documentclass{ximera}

\input{../preamble}

\title[Dig-In:]{Integration by substitution}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle


Computing antiderivatives is not as easy as computing derivatives. One
issue is that the chain rule can be difficult to ``undo.'' Sometimes 
it is helpful to transform the integral in question via substitution. 


\begin{theorem}[Integral Substitution Formula] 
If $u(x)$ is differentiable on the interval $[a,b]$ and $f(x)$ is
differentiable on the interval $[u(a),u(b)]$, then
\[
\int_a^b f'(u(x)) u'(x) \d x =\int_{u(a)}^{u(b)} f'(u) \d u.
\]
\end{theorem}
\begin{proof} First we recognize the chain rule
\[
\int_a^b f'(u(x)) u'(x) \d x = \int_a^b (f\circ u)'(x) \d x.
\]
Next we apply the Fundamental Theorem of Calculus. 
\begin{align*} 
\int_a^b (f\circ u)'(x) \d x &= \eval{f(u(x))}_a^b \\
&= \eval{f(x)}_{u(a)}^{u(b)}\\ 
&= \int_{g(a)}^{g(b)} f'(u) \d u.
\end{align*}
\end{proof}


There are several different ways to think about substitution. The
first is directly using the formula
\[
\int_a^b f'(u(x)) u'(x) \d x = \int_{u(a)}^{u(b)} f'(u) \d u.
\]
\begin{example}
Compute
\[
\int_1^3 x\cos(x^2)\d x.
\]


A little thought reveals that if $x\cos(x^2)$ is the derivative of
some function, then it must have come from an application of the chain
rule. Here we have $x$ on the ``outside,'' which is the derivative of
$x^2$ on the ``inside,'' 
\[
\int \underbrace{x}_{\text{outside}}\cos(\underbrace{x^2}_{\text{inside}})\d x.
\]
Set $u(x) = x^2$ so $u'(x) = 2x$ and now it must be that $f(u) =
\frac{\cos(u)}{2}$. Now we see
\begin{align*}
\int_1^3 x\cos(x^2)\d x &= \int_1^9 \frac{\cos(u)}{2}\d u\\
&= \eval{\frac{\sin(u)}{2}}_1^9 \\
&= \frac{\sin(9) -\sin(1)}{2}.
\end{align*}
\end{example}

Sometimes we frame the solution in a different way. Let's do the same
example again, this time we'll think in terms of differentials.

\begin{example}
Compute
\[
\int_1^3 x\cos(x^2)\d x.
\]


Here we will set $u=x^2$. Now $du = 2x \d x$, we are thinking in terms
of differentials. Now we see
\[
\int_{u(1)}^{u(3)} \frac{\cos(u)}{2}\d u = \int_1^3\frac{\cos(x^2)}{2}2x\d x.
\]
At this point, we can continue as we did before and write
\[
\int_1^3 x\cos(x^2)\d x= \frac{\sin(9) -\sin(1)}{2}.
\]
\end{example}

Finally, sometimes we simply want to deal with the antiderivative on
its own, we'll repeat the example one more time demonstrating this.

\begin{example}
Compute
\[
\int_1^3 x\cos(x^2)\d x.
\]


Here we start as we did before, setting $u=x^2$. Now $du = 2x \d x$,
again thinking in terms of differentials. Now we see
\[
\int  \frac{\cos(u)}{2}\d u = \int \frac{\cos(x^2)}{2}2x\d x.
\]
Hence 
\[
\int x\cos(x^2)\d x = \frac{\sin(u)}{2} = \frac{\sin(x^2)}{2}.
\]
Now we see
\begin{align*}
\int_1^3 x\cos(x^2)\d x &=\eval{\frac{\sin(x^2)}{2}}_1^3\\
&= \frac{\sin(9) -\sin(1)}{2}.
\end{align*}
\end{example}

With some experience, it is not hard to see which function is $f(x)$
and which is $u(x)$, let's see another example.
\begin{example}
Compute
\[
\int x^4(x^5+1)^{99} \d x.
\]


Here we set $u = x^5+1$ so $du = 5x^4 \d x$, and $f(u) = \frac{u^{99}}{5}$. Now
\begin{align*}
\int x^4(x^5+1)^{99} \d x &= \int \frac{u^{99}}{5} \d u\\
&= \frac{u^{100}}{500}.
\end{align*}
Recalling that $u = x^5+1$, we have our final answer
\[
\int x^4(x^5+1)^{99} \d x= \frac{(x^5+1)^{100}}{500}+C.
\]
\end{example}


Our next example is a bit different.

\begin{example}
Compute
\[
\int_{2}^{3} \frac{1}{x\ln(x)} \d x.
\]


Let $u=\ln(x)$ so $du=\frac{1}{x}\d x$. Write
\begin{align*}
\int_{2}^{3} \frac{1}{x\ln(x)} \d x = \int_{\ln(2)}^{\ln(3)} \frac{1}{u} \d u\\
&= \eval{\ln(u)}_{\ln(2)}^{\ln(3)}\\
& = \ln(\ln(3)) - \ln(\ln(2)).
\end{align*}
\end{example}


On the other hand our next example is much harder.

\begin{example} Compute
\[
\int x^3\sqrt{1-x^2}\d x.
\]


Here it is not apparent that the chain rule is involved. However, if
it was involved, perhaps a good guess for $u$ would be
\[
u = 1-x^2
\]
in this case
\[
du = -2x \d x.
\]
Now consider our indefinite integral
\[
\int x^3\sqrt{1-x^2}\d x,
\]
immediately we can substitute. Write
\[
\int x^3\sqrt{1-x^2}\d x = \int -\frac{x^2\sqrt{u}}{2}\d u.
\]
However, we cannot continue until each $x$ is replaced. We know however that 
\begin{align*}
u &= 1-x^2 \\
u -1 &= -x^2\\
1- u &= x^2
\end{align*}
so now we may write
\[
\int x^3\sqrt{1-x^2}\d x = \int -\frac{(1-u)\sqrt{u}}{2}\d u.
\]
At this point, we are close to being done. Write
\begin{align*}
\int -\frac{(1-u)\sqrt{u}}{2}\d u &= \int \left(\frac{u\sqrt{u}}{2} - \frac{\sqrt{u}}{2}\right) \d u \\
&= \int \frac{u^{3/2}}{2} \d u - \int \frac{\sqrt{u}}{2} \d u \\
&= \frac{u^{5/2}}{5} - \frac{u^{3/2}}{3}.
\end{align*}
Now recall that $u = 1-x^2$. Hence our final answer is
\[
\int x^3\sqrt{1-x^2}\d x = \frac{(1-x^2)^{5/2}}{5} - \frac{(1-x^2)^{3/2}}{3}+C.
\]
\end{example}

To summarize, if we suspect that a given function is the derivative of
another via the chain rule, we let $u$ denote a likely candidate for
the inner function, then translate the given function so that it is
written entirely in terms of $u$, with no $x$ remaining in the
expression. If we can integrate this new function of $u$, then the
antiderivative of the original function is obtained by replacing $u$
by the equivalent expression in $x$.


\end{document}
