\documentclass{ximera}

\input{../../preamble.tex}


\author{Jim Talamo}
\license{Creative Commons 3.0 By-NC}
\outcome{}
%\newcommand\Mydiv[2]{%
%$\strut#1$\kern.25em\smash{\raise.3ex\hbox{$\big)$}$}\mkern-8mu
%\overline{\enspace\strut#2$}}

\begin{document}

\begin{exercise}
We define an \emph{irreducible quadratic polynomial} as follows.

Let $p(x) = ax^2+bx+c$.  Then, $p(x)$ is an \emph{irreducible quadratic} if it cannot be factored as a product of degree one polynomials with real coefficients.

\begin{example}
Suppose that $p(x) = x^2-3x+2$.  Notice that \[p(x) = x^2-3x+2 = (x-2)(x-1),\] so $p(x)$ is not an irreducible quadratic.
\end{example}

A theorem in algebra guarantees that any polynomial with real coefficients can be factored as a product of linear polynomials and irreducible quadratic polynomials with real coefficients.  

This theorem is pivotal to develop a theory of integrating rational functions, and a fundamental skill that becomes necessary to master is determining whether a given quadratic polynomial is irreducible or not.  Part of the difficulty in making this determination arises because there is a difference between not seeing how to factor a polynomial versus showing such a factorization is not possible.  Thankfully, there's a theorem for that.

\begin{theorem}
A quadratic polynomial $p(x) = ax^2+bx+c$ is irreducible if and only if the quadratic equation $ax^2+bx+c =0$ has no real roots.
\end{theorem}

A useful result in this vein is thus the quadratic formula, which states that the solutions to $ax^2+bx+c=0$ are:
\[
x = \frac{-b\pm\sqrt{b^2-4ac}}{2a}
\]

\begin{example}
Suppose that $p(x) = x^2-3x+1$.  Notice that it is difficult to factor this polynomial, but the roots of the equation $x^2-3x+1 = 0$ are

\[
x = \frac{3\pm\sqrt{(-3)^2-4(1)(1)}}{2(1)}=\frac{3 \pm \sqrt{5}}{2}
\]
The roots are \wordChoice{\choice[correct]{real}\choice{not real}}so $p(x)$ \wordChoice{\choice{is}\choice[correct]{is not}} an irreducible quadratic.
\end{example}

\begin{example}
Suppose that $p(x) = x^2-3x+4$.  Notice that it is difficult to factor this polynomial, but the roots of the equation $x^2-3x+1 = 0$ are

\[
x = \frac{3\pm\sqrt{(-3)^2-4(4)(1)}}{2(1)}=\frac{3 \pm \sqrt{-7}}{2}
\]
The roots are \wordChoice{\choice{real}\choice[correct]{not real}}so $p(x)$ \wordChoice{\choice[correct]{is}\choice{is not}} an irreducible quadratic.
\end{example}

Now, select all of the choices below that are irreducible quadratic polynomials.
\begin{selectAll}
\choice[correct]{$x^2+1$}
\choice{$x^2-10x+24$}
\choice[correct]{$2x^2+5x+4$}
\choice{$x^2+5x+2$}
\choice[correct]{$x^2-x+5$}
\end{selectAll}
\end{exercise}
\end{document}
