\documentclass{ximera}

\input{../../preamble.tex}

\author{Gregory Hartman \and Matthew Carr\and Nela Lakos}
\license{Creative Commons 3.0 By-NC}
\acknowledgement{https://github.com/APEXCalculus}

\outcome{Compute definite integrals using the properties of integrals.}
\outcome{Given a velocity function, calculate displacement and distance traveled.}
\outcome{Understand the relationship between position, velocity and acceleration.}
\outcome{Solve basic word problems involving maxima or minima.}
\outcome{Understand the difference between displacement and distance traveled.}
\outcome{Understand the relationship between indefinite and definite integrals.}

\begin{document}
\begin{exercise}

An object is thrown straight up  from a height of $48$ ft and its velocity, in ft/s, is given by
$v(t)=-32t+64$, for $t$ in time interval $I$, where $t$ is time in seconds.

\begin{enumerate}
\item What is the object's maximum velocity? \begin{prompt}\[v_{max}=\answer{64}\,ft/s\]\end{prompt}
\item Find the position of the object at any time $t$, for $t$ in $I$? 
\begin{hint}
You have to solve the Initial Value Problem:

$s'(t)=-32t+64$

$s(0)=48$
\end{hint}
\begin{prompt}\[s(t)=\answer{-16} t^2+\answer{64} t+\answer{48} \,ft/s\]\end{prompt}
\item When will the object reach the ground? 
\begin{prompt}\[t=\answer{2+\sqrt{7}}\,s\]\end{prompt}

\item What is the object's maximum displacement from its starting position? \begin{prompt}\[d=\answer{64}\,ft\]\end{prompt}
\item When does the maximum displacement occur? \begin{prompt}\[t=\answer{2}\,s\]\end{prompt}

\end{enumerate}

\end{exercise}
\end{document}
