\documentclass{ximera}

\input{../preamble.tex}

\outcome{Compute limits of vector-valued functions.}
\outcome{Determine continuity of vector-valued functions.}
\outcome{Compute the derivative of a vector-valued function.}
\outcome{Compute the tangent vector of a vector-valued function.}
\outcome{Find the unit tangent vector of a vector-valued function.}
\outcome{Compute integrals of vector-valued functions.}


\title[Dig-In:]{Calculus and vector-valued functions}

\begin{document}
\begin{abstract}
  With one input, and vector outputs, we work component-wise.
\end{abstract}
\maketitle

A question I've often asked myself is: ``How do you know when you are
doing a calculus problem?'' The answer, I think, is that you are doing
a calculus problem when you are computing: a limit, a derivative, or
an integral. Now we are going to do calculus with vector-valued
functions.  To build a theory of calculus for vector-valued functions,
we simply treat each component of a vector-valued function as a
regular, single-variable function.  Since we are currently thinking
about vector-valued functions that only have a single input, we can
work \dfn{component-wise}. Let's see this in action.



\section{Limits of vector-valued functions}

With a vector-valued function, you have something like
\begin{align*}
  \vec{f} &= \vector{x(t),y(t),z(t)}\\
  &= \veci x(t) + \vecj y(t)  + \veck z(t)
\end{align*}
where each component is completely independent of the other
components. When computing a limit, we can write:
\begin{align*}
  \lim_{t\to a} \vec{f}(t) &= \lim_{t\to a}\left(\veci x(t) + \vecj y(t)  + \veck z(t)\right)\\
  &= \lim_{t\to a}\veci x(t) + \lim_{t\to a}\vecj y(t)  + \lim_{t\to a}\veck z(t)\\
  &= \veci \lim_{t\to a} x(t) + \vecj\lim_{t\to a} y(t)  + \veck\lim_{t\to a} z(t)\\
  &= \vector{\lim_{t \to a}x(t),\lim_{t \to a}y(t),\lim_{t \to a}z(t)}.
\end{align*}

This is worth stating as a theorem.

\begin{theorem}\index{limit!of a vector-valued function}
  Let $\vec{f}:\R \to \R^3$ be a vector-valued function
  \[
  \vec{f}(t) = \vector{x(t),y(t),z(t)}
  \]
  then the vector-valued limit as $t$ goes to $a$ is given by 
  \[
  \lim_{t \to a} \vec{f}(t) = \vector{\lim_{t \to a}x(t),\lim_{t \to a}y(t),\lim_{t \to a}z(t)}.
  \]
  This limit exists if and only if each of the limits of the
  components exist.
\end{theorem}

We evaluate limits by just taking the limit of each component
separately.

\begin{question}
  Let $\vec{f}(t) = \vector{\sin(t),\cos(t),\frac{\sin(t)}{t}}$.
  Compute:
  \[
  \lim_{t \to 0} \vec{f}(t)
  \begin{prompt}
    =\vector{\answer{0},\answer{1},\answer{1}}
  \end{prompt}
  \]
  \begin{hint}
    Take the limit of each component separately.
  \end{hint}
\end{question}


Now that we have the notion of limits, we may also define the concept
of continuity of vector-valued functions:

\begin{definition}
  A vector-valued function $\vec{f}$ is \dfn{continuous} at $t= a$ if
  and only if
  \[
  \lim_{t \to a} \vec{f}(t)  = \vec{f}(a)
  \]
\end{definition}
Because of the component-wise nature of limits, we can see that a
function $\vec{f}$ is continuous if and only if each component
function $x(t)$, $y(t)$, $z(t)$ is also continuous at $t=a$.

\begin{question}
Which of the following vector-valued functions are continuous for all
real values of $t$?
\begin{selectAll}
  \choice{$\vector{t,t^2,\tan(t)}$}
  \choice{$\vector{\frac{2}{t-5},\sin(t),0}$}
  \choice[correct]{$\vector{\cos(5 t), t^2-3t+1, e^t}$}
  \choice{$\vector{e^{\sqrt{t}}, 4\sin(t), t^7}$}
  \choice{$\vector{\ln(t^2), \ln(t), t^{3/2}}$}
  \choice[correct]{$\vector{0, 0, 0}$}
\end{selectAll}
\begin{hint}
  Recall the following theorem:
  \begin{theorem}[Continuity of Famous Functions]\index{continuity of famous functions}\label{theorem:continuity}
    The following functions are continuous on their natural domains, for $k$ a real number and $b$ a positive real number:
    \begin{description}
    \item[Constant function]\index{constant function} $f(x) =k$
    \item[Identity function]\index{identity function} $f(x) = x$
    \item[Power function]\index{power function} $f(x)=x^b$
    \item[Exponential function]\index{exponential function} $f(x)=b^x$
    \item[Logarithmic function]\index{logarithmic function} $f(x)=\log_b(x)$
    \item[Sine and cosine] Both $\sin(x)$ and $\cos(x)$ \index{sine}\index{cosine}
    \end{description}
    In essence, we are saying that the functions listed above are
    continuous wherever they are defined.
  \end{theorem}
\end{hint}
\end{question}


\section{Derivatives}

Recall the limit definition of the derivative:
\[
\ddx f(x) = \lim_{h\to 0} \frac{f(x+h)-f(x)}{h}
\]
We have a similar limit definition of vector-valued functions:
\[
\dd{t} \vec{f}(t) = \lim_{h\to 0} \frac{\vec{f}(t+h)-\vec{f}(t)}{h}
\]
Since limits can be computed component-wise, this derivative can be computed component-wise.

\begin{theorem}
  Let $\vec{f}:\R \to \R^3$ be a vector-valued function
  \[
  \vec{f}(t) = \vector{x(t),y(t),z(t)},
  \]
  then the vector-valued \dfn{derivative} is given by
  \[
  \vec{f}'(t) = \vector{x'(t),y'(t),z'(t)}.
  \]
\end{theorem}

The derivative of a vector-valued function gives a vector that points
in the direction that the vector-valued function draws the curve.

\begin{onlineOnly}
  Below we see the derivative of the vector-valued function along with
  an approximation of the limit for small values of $h$:
  \begin{center}
    \geogebra{CmhkZhqk}{800}{600} %%https://ggbm.at/CmhkZhqk
  \end{center}
\end{onlineOnly}




\begin{question}
  Let $\vec{f}(t) = \vector{\arctan(t),7^t, \sec(t^2)}$.
  Compute:
  \[
  \dd{t} \vec{f}(t)
  \begin{prompt}
    =\vector{\answer{\frac{1}{1+t^2}},\answer{7^t \ln(7)},\answer{ \sec(t^2)\tan(t^2)2t}}
  \end{prompt}
  \]
\end{question}

We also have some (additional) derivative rules:
\begin{theorem}
  Let $\vec{f}$ and $\vec{g}$ be differentiable vector-valued
  functions, $s(t)$ be a differentiable scalar function, and $c$ be a
  constant. Then:
  \begin{itemize}
  \item $\dd{t} \left(\vec{f}(t) + \vec{g}(t) \right) = \vec{f}'(t) + \vec{g}'(t)$
  \item $\dd{t} c\vec{f}(t) = c\vec{f}'(t)$
  \item $\dd{t} s(t)\vec{f}(t) = s'(t)\vec{f}(t) + s(t)\vec{f}'(t)$
  \item $\dd{t} \left(\vec{f}(t)\dotp \vec{g}(t)\right) = \vec{f}'(t)\dotp \vec{g}(t) + \vec{f}(t)\dotp\vec{g}'(t)$
  \item $\dd{t} \left(\vec{f}(t)\cross \vec{g}(t)\right) = \vec{f}'(t)\cross \vec{g}(t) + \vec{f}(t)\cross\vec{g}'(t)$
  \item $\dd{t} \vec{f}(s(t)) = \vec{f}'(s(t))\cdot s'(t)$
  \end{itemize}
\end{theorem}


\begin{question}
  Suppose that $s:\R\to\R$, $\vec{f}:\R\to\R^3$, and $\vec{g}:\R\to\R^3$ where:
  \begin{itemize}
  \item $s(2) = 3$ and  $s'(2) =-1$.
  \item $\vec{f}(2) = \vector{1,-1,0}$ and $\vec{f}'(2) = \vector{-1,1,0}$.
  \item $\vec{g}(2) = \vector{0,1,2}$ and $\vec{g}'(2) = \vector{2,0,-1}$.
  \end{itemize}
    Compute: $\eval{\dd{t} s(t)\vec{f}(t)}_{t=2}$
    \begin{prompt}
      \[
      \eval{\dd{t} s(t)\vec{f}(t)}_{t=2} = \vector{\answer{-4},\answer{4},\answer{0}}
      \]
    \end{prompt}
    \begin{question}
      Compute: $\eval{\dd{t} \vec{f}(t)\dotp\vec{g}(t)}_{t=2}$
      \begin{prompt}
        \[
        \eval{\dd{t} \vec{f}(t)\dotp\vec{g}(t)}_{t=2} = \answer{3}
        \]
      \end{prompt}
      \begin{question}
        Compute: $\eval{\dd{t} \vec{f}(t)\cross\vec{g}(t)}_{t=2}$
        \begin{prompt}
          \[
          \eval{\dd{t} \vec{f}(t)\cross\vec{g}(t)}_{t=2} = \vector{\answer{3},\answer{3},\answer{1}}
          \]
        \end{prompt}
      \end{question}
    \end{question}
  \end{question}
  



The derivative of a vector-valued function gives a \textit{tangent
  vector}. A tangent vector is a vector that points in the direction
that the curve is drawn.

\begin{onlineOnly}
  Below we see the derivative of the vector-valued function along with
  an approximation of the limit for small values of $h$:
  \begin{center}
    \geogebra{TgtNuNhF}{800}{600} %%https://ggbm.at/TgtNuNhF
  \end{center}
\end{onlineOnly}

\begin{definition}
Let $\vec{f}(t)$ be a differentiable vector-valued function on an open
interval $I$ containing $c$, where $\vec{f}'(c)\neq \vec{0}$.
\begin{itemize}
\item A vector $\vec v$ is a \dfn{tangent vector} to the graph of
  $\vec{f}(t)$ at $t=c$, if $\vec v$ is parallel to $\vec{f}'(c)$.
\item The \dfn{tangent line} to the graph of $\vec f(t)$ at $t=c$
  is the line through $\vec f(c)$ in a direction parallel to
  $\vec{f}'(c)$. An equation of the tangent line is
  \[
 \vec{ \boldsymbol{\l}}(t) = \vec f(c) + t\vec{f}'(c).
  \]
\item The direction of $\vec{f}$ is given by the \dfn{unit tangent
  vector} of $\vec{f}$ is given by
  \[
  \utan(t)=\frac{\vec{f}'(t)}{|\vec{f}'(t)|} = \frac{\vec{f}'(t)}{\sqrt{\vec{f}'(t)\dotp \vec{f}'(t)}}.
  \]
\end{itemize}
\end{definition}

\begin{question}
  Let $\vec{f}(t) = \vector{t,t^2,t^3}$ on $[-1.5,1.5]$.  Find the
  vector equation of the line tangent to the graph of $\vec f$ at
  $t=-1$.
  \begin{prompt}
    \[
    \vec{\boldsymbol{\l}}(t) = \vector{\answer{-1},\answer{1},-1} + t\vector{\answer{1},-2,\answer{3}}
    \]
  \end{prompt}
\end{question}

We want to be able to predict how a curve drawn by a vector-valued
function behaves based on the function's derivative. However, if the
derivative is ever the zero vector, we loose all such information.
When this happens we cannot use the tools of calculus. Because of
this, we have a special name for functions where the tangent vector is
never the zero vector:

\begin{definition}
Let $\vec{f}(t)$ be a differentiable vector-valued function on an open
interval $I$. $\vec f(t)$ is \dfn{smooth} on $I$ if
\[
\vec{f}'(t)\neq\vec{0}
\]
for $t$ in $I$.
\end{definition}

\begin{question}
  Is the vector-valued function $\vec{f}(t) = \vector{t^2,t^2}$ smooth on
  $\R$?
  \begin{prompt}
    \begin{multipleChoice}
      \choice{yes}
      \choice[correct]{no}
    \end{multipleChoice}
    \begin{feedback}
      This function looks like a ``ray.'' Conceptually it isn't smooth
      because it goes to the origin and turns around ``abruptly.'' Check it out:
      \[
      \graph{(t^2,t^2)}
      \]
    \end{feedback}
  \end{prompt}
\end{question}


Finally, we point out that if a vector-valued function has constant
length, then there is a special relationship between the function and
its tangent vectors.


\begin{theorem}
  Let $\vec f(t)$ be a differentiable vector-valued function on an
  open interval $I$. $\vec{f}(t)$ is of constant length if and only if
  the function and its tangent vector are orthogonal.
  \begin{explanation}
    Suppose that $|\vec f(t)| = c$ for all $t$ in $I$. In particular,
    we see
    \[
    \vec{f}(t) \dotp \vec{f}(t) = c^2.
    \]
    Now take the derivative with respect to $t$ of both sides and
    apply the derivative rules from above. Write with me
    \begin{align*}
      \vec{f}'(t) \dotp \vec{f}(t) + \vec{f}(t)\dotp \vec{f}'(t) &= 0\\
      2\vec{f}'(t) \dotp \vec{f}(t) &= 0\\
      \vec{f}'(t) \dotp \vec{f}(t) &= 0
    \end{align*}
    and this means that the vectors $\vec{f}'(t)$ and $\vec{f}(t)$ are
    \wordChoice{\choice{parallel}\choice[correct]{perpendicular}}.

    Now suppose that for all $t$ in $I$, $\vec{f}'(t)$ and
    $\vec{f}(t)$ are orthogonal. This means
    \begin{align*}
      \vec{f}'(t) \dotp \vec{f}(t) &= 0\\
      2\vec{f}'(t) \dotp \vec{f}(t) &= 0\\
      \vec{f}'(t) \dotp \vec{f}(t) + \vec{f}(t)\dotp \vec{f}'(t) &= 0,
    \end{align*}
    but we know where the derivative on the left came from! So we may
    write
    \[
    \vec{f}(t) \dotp \vec{f}(t) = c^2
    \]
    for some constant $c$. Now we see that $\vec{f}$ is a
    vector-valued function of constant length on an open interval
    $I$.
  \end{explanation}
\end{theorem}
The theorem above says a lot in a very little. To start, it says that
if the magnitude of a vector-valued function is constant, then the
vector-value of the function and its tangent vector are
orthogonal. Moreover, this means that the curve looks locally like a
circle. However, it does not mean the curve looks like a circle globally.
\begin{onlineOnly}
  For example, consider this curve:

  \begin{center}
    \geogebra{BjaXggAp}{800}{600} %%https://ggbm.at/BjaXggAp
  \end{center}

  By moving the slider around you can see the vector of constant
  length that draws the curve and the tangent vector. Note that these
  vectors are orthogonal.
\end{onlineOnly}






\section{Integrals}
Since we took the derivative of vector-valued functions by
differentiating each component, we will also compute indefinite and
definite integrals by computing antiderivatives of each component.


\begin{theorem}
  Let $\vec{f}:\R \to \R^3$ be a continuous vector-valued function
  \[
  \vec{f}(t) = \vector{x(t),y(t),z(t)},
  \]
  then the vector-valued \dfn{integral} is given by
  \[
  \int \vec{f}(t) \d t= \vector{\int x(t) \d t,\int y(t)\d t,\int z(t)\d t}.
  \]
\end{theorem}

\begin{question}
  Let $\vec f(t) = \vector{e^{2t},\sin t}$. Compute:
  \[
  \int_0^1 \vec f(t) \d t
  \begin{prompt}
    = \vector{\answer{\frac{1}{2}(e^2-1)},\answer{-\cos(1)+1}}
  \end{prompt}
  \]
\end{question}

We can also solve initial value problems, check out our next example:

\begin{example}
  Let $\vec{f}''(t) = \vector{2, \cos(t), 12t}$. Find $\vec f(t)$ where:
  \begin{itemize}
  \item $\vec f(0) = \vector{-7,-1,2}$ and
  \item $\vec{f}'(0) = \vector{5,3,0}.$
  \end{itemize}
  \begin{explanation}
    Knowing $\vec{f}''(t) = \vector{2,\cos(t), 12t}$, we find
    $\vec{f}'(t)$ by evaluating the indefinite integral.
    \begin{align*}
      \int \vec{f}''(t)\d t &= \vector{\int \answer[given]{2}\d t, \int \answer[given]{\cos(t)}\d t , \int \answer[given]{12t}\d t} \\
      &= \vector{\answer[given]{2t}+C_1, \answer[given]{\sin(t)}+ C_2, \answer[given]{6t^2} + C_3} \\
      &= \vector{2t,\sin(t),6t^2} + \vector{C_1,C_2,C_3} \\
      &= \vector{2t,\sin(t),6t^2} + \vec{C}.
    \end{align*}
    Note how each indefinite integral creates its own constant which
    we collect as one constant vector $\vec C$. Knowing
    \[
    \vec{f}'(0) = \vector{5,3,0}
    \]
    allows us to solve for $\vec C$:
    \begin{align*}
      \vec{f}'(t) & = \vector{2t,\sin(t),6t^2} + \vec C\\
      \vec{f}'(0) &= \vector{\answer[given]{0},\answer[given]{0},\answer[given]{0}} + \vec C\\
      \vector{5,3,0} &= \vec C.
    \end{align*}
    So
    \begin{align*}
      \vec{f}'(t) &= \vector{2t,\sin(t),6t^2} + \vector{\answer[given]{5},\answer[given]{3},\answer[given]{0}}\\
      &=\vector{2t+5, \sin(t) + 3, 6t^2}.
    \end{align*}
    To find $\vec f(t)$, we integrate once more.
    \begin{align*}
      \int \vec{f}'(t)\d t &= \vector{\int\answer[given]{2t+5}\d t, \int \answer[given]{\sin(t) + 3}\d t, \int \answer[given]{6t^2}\d t }\\
      &= \vector{\answer[given]{t^2+5t}, \answer[given]{-\cos(t) + 3t}, \answer[given]{2t^3}} + \vec C
    \end{align*}
    With $\vec f(0) = \vector{-7,-1,2}$, we solve for $\vec C$:
    \begin{align*}
      \vec f(t) &= \vector{\answer[given]{t^2+5t}, \answer[given]{-\cos(t) + 3t}, \answer[given]{2t^3}} + \vec C\\
      \vec f(0) &= \vector{\answer[given]{0},\answer[given]{-1},\answer[given]{0}} + \vec C\\
      \vector{-7,-1,2} &= \vector{0,-1,0} + \vec C.
    \end{align*}
    Hence:
    \[
    \vec{C}=\vector{\answer[given]{-7},\answer[given]{0},\answer[given]{2}}
    \]
    So
    \begin{align*}
      \vec f(t) &= \vector{t^2+5t, -\cos(t) + 3t, 2t^3} + \vector{-7,0,2}\\
      &= \vector{\answer[given]{t^2+5t-7},\answer[given]{-\cos(t)+3t},\answer[given]{2t^3+2}}.
    \end{align*}
  \end{explanation}
\end{example}

We now leave you with a question: What does  integration of a
vector-valued function \textit{mean}?  There are many applications,
but none as direct as ``the area under the curve'' that we used in
understanding the integral of a real-valued function. We will explore
this later in our study of calculus.

\end{document}
