\documentclass{ximera}
\input{../../preamble.tex}
\author{Bart Snapp\and Nela Lakos}
\license{Creative Commons 3.0 By-NC}
\acknowledgement{https://www.whitman.edu/mathematics/calculus/}
  \outcome{Interpret an optimization problem as the procedure used to make a system or design as effective or functional as possible.}
  \outcome{Set up an optimization problem by identifying the objective function and appropriate constraints.}
  \outcome{Solve optimization problems by finding the appropriate absolute extremum.}
  \outcome{Solve basic word problems involving maxima or minima.}
\begin{document}
\begin{exercise}

  You have $100$ feet of fence to make a rectangular play area
  alongside the wall of your house.  The wall of the house bounds one
  side.  What is the largest size possible (in square feet) for the
  play area?
  \begin{hint}
 \begin{image}
\begin{tikzpicture}
	\begin{axis}[
            domain=-3:3, ymin=0, ymax=9, xmin=-3, xmax=3,
                     ticks=none,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top,
          ]
            \addplot [very thick,penColor5,domain=(-3:3)] {7};
          \addplot [draw=none, fill=fill2, domain=(-1.5:1.5)] {7} \closedcycle;
          \addplot [draw=none, fill=background, domain=(-1.5:1.5)] {2.25} \closedcycle;
          \addplot [very thick, penColor] plot coordinates {(1.5,2.25) (-1.5,2.25)};
          \addplot [very thick, penColor] plot coordinates {(1.5,2.25) (1.5,7)};
          \addplot [very thick, penColor] plot coordinates {(-1.5,7) (-1.5,2.25)};
        \node at (axis cs:-2,4.625) [penColor] {$x$};
 \node at (axis cs:0,1.625) [penColor] {$y$};
          \node at (axis cs:0,4.625) [penColor] {$A=x\cdot y$};
  \node at (axis cs:-2,7.625) [penColor] {$Wall$};
      
        \end{axis}
\end{tikzpicture}
\end{image}
\end{hint}
\begin{hint}
We have to maximize the area, $A$.
Since, $100=2x+y$, it follows that

$y=100-2x$.
\end{hint}
\begin{hint}
We now express the area as the function of $x$.

$A(x)=\answer{x}\cdot(100-2x)$
\end{hint}
\begin{hint}
So, we have to find the global maximum of the function $A$, given by

$A(x)=100\cdot\answer{x}-2x^2$

on its domain $[0,50]$.
\end{hint}
\begin{hint}
We have to find all the critical points of $A$ on $[0,50]$.
Since

$A'(x)=\answer{100}-4x$,

it follows that  the function $A$ has its only critical point at $x=\answer{25}$.

\end{hint}
\begin{hint}
We have to evaluate $A$ at the end points and the critical point.

$A(0)=\answer{0}$

$A(50)=\answer{0}$

$A(25)=\answer{1250}$

Now we compare these three values, and determine the global maximum of $A$.
\end{hint}
  \begin{prompt}
  \[
  \answer{1250} ft^2
  \]
  \end{prompt}
\end{exercise}
\end{document}
