\documentclass[handout,hints]{ximera}

\input{../../../preamble.tex}

\usepackage{paralist}

\author{Tom Needham}

\outcome{Recall basic rules and techniques of integration.}

\title[]{Review of Integration Techniques}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

\vspace{-0.9in}

\section{Discussion Questions}

\begin{problem}
Classify each of the following expressions as a number, a function of $t$, a function of $x$ or none of the above.

\begin{center}
\begin{inparaenum}
\item $\frac{\d}{\d t} \int_0^t e^{x^2} \d x$ \hspace{0.2in}
\item $\int e^{x^2} \d x$ \hspace{0.2in}
\item $\int_0^2 e^{x^2} \d x$ \hspace{0.2in}
\item $\frac{d}{dx} \int_0^9 e^{x^2} \d x$
\end{inparaenum}
\end{center}
\end{problem}

\begin{problem}
Compute the following indefinite integrals via basic antiderivative rules, without explicitly using $u$-substitution.

\begin{center}
\begin{inparaenum}
\item $\int \cos(2x) \d x$ \hspace{0.4in}
\item $\int e^{x/2} \d x$
\end{inparaenum}
\end{center}

\end{problem}

\begin{problem}
Consider the indefinite integral $\int \frac{1}{3x} \d x$. Student A calculates the integral using basic antiderivative rules as
$$
\int \frac{1}{3x} \d x = \frac{1}{3} \ln |3x| + C.
$$
Student B calculates the integral by moving the constant to the front as 
$$
\int \frac{1}{3x} \d x = \frac{1}{3} \int \frac{1}{x} \d x = \frac{1}{3} \ln |x| + C.
$$
Which student is correct?
\end{problem}


\section{Group Work}


\begin{problem}
Calculate the following integrals.


\begin{inparaenum}
\item $\int \left(\sqrt{y}+1\right)^2 \d y$ \hspace{.05in}
\item $\int_0^{\sqrt{\ln(2)}} x e^{x^2} \d x$ \hspace{.05in}
\item $\int \frac{2x+1}{x^2+4} \d x$ \hspace{.05in}
\item $\int_{-\pi/8}^{\pi/8} \frac{\cos^2(2t) + 1}{\cos^2(2t)} \d t$ 
\end{inparaenum}

\end{problem}



\begin{problem}
Find a function $f(x)$ satisfying each of the given condition(s).

$$
\begin{array}{lll}
\mathrm{(a)} \; \int f(x) \d x = 2x \sqrt{1+ x^2} + C & \hspace{.2in} & \mathrm{(b)} \; \int f(x) \d x = x e^x + C \\
\mathrm{(c)} \; \frac{\d}{\d t} \int_0^t f(x) \d x = t e^t & \hspace{.2in} & \mathrm{(d)} \; f'(x) = 3x^2 \mbox{ and } f(0)=2
\end{array}
$$
\end{problem}



\begin{problem}
A student calculates an indefinite integral as follows:
$$
\int \frac{1}{3x^2 + 5} \d x = \ln (3x^2 + 5) + C.
$$
Is the student correct? First explain your answer without actually calculating the antiderivative. Then, if the student is incorrect, calculate the correct antiderivative.
\end{problem}

\begin{problem}
The figure below shows a plot of a function $y = f(x)$. On the same axes, draw representations of 

\begin{center}
\begin{inparaenum}
\item $\int f(x) \d x$ \hspace{0.4in}
\item $\int_{-2}^2 f(x) \d x$
\end{inparaenum}
\end{center}

\begin{image}
          \begin{tikzpicture}
	    \begin{axis}[
            domain=-3:3,
            xmin=-3, xmax=3,
            ymin=-2, ymax=6
         ,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
          ]
	  \addplot [very thick, penColor, smooth] {x^2};
        
        \end{axis}
\end{tikzpicture}
%% \caption{A plot of $f(x)=x^2$ and $f^{-1}(x) = \sqrt{x}$. While
%%   $f(x)=x^2$ is not one-to-one on $\RR$, it is one-to-one on
%%   $[0,\infty)$.}
%% \label{plot:fxn and inverse x^2}
\end{image}
\end{problem}


\end{document}
