\documentclass[handout,nooutcomes,noauthor]{ximera}

\input{../../../preamble.tex}

\author{Bart Snapp}

\title[Collaborate:]{Distance between points and\dots}

\begin{document}
\begin{abstract}
  Investigate the distance between points and lines or planes. 
\end{abstract}
\maketitle

\textbf{Work in groups of 3--4, writing your answers on a separate
  sheet of paper.}

Consider a line and a point.
\begin{image}
  \begin{tikzpicture}
    \begin{axis}[
        xmin=-1,xmax=5,ymin=-1,ymax=3,
        clip=true,
        axis lines=center,
        %ticks=none,
        unit vector ratio*=1 1 1,
        xlabel=$x$, ylabel=$y$,
        %ytick={-2,-1,...,7},
	%xtick={-2,-1,...,10},
	grid = major,
        every axis y label/.style={at=(current axis.above origin),anchor=south},
        every axis x label/.style={at=(current axis.right of origin),anchor=west},
      ]
      \addplot[ultra thick,penColor] {-x/2+2};
      \filldraw [penColor] (axis cs:3,2) circle (3pt);     

    \end{axis}
  \end{tikzpicture}
\end{image}
We wish to find the (minimum) distance between the point and the
line. Let's solve this problem in as many ways as possible.

\begin{problem}
  Draw the line segment from the point to the line of minimum
  length. Compare your drawing with others, make sure you all agree.
\end{problem}


\begin{problem}
  Solve this problem as you might in Calculus I: Find an explicit
  formula for a line, $y = mx+b$, then minimize the distance (or the
  distance squared!) between the point and the points on the line.
\end{problem}

\begin{problem}
  Use facts about vectors and projection, $\proj_{\vec{w}}(\vec{v})$,
  to find the minimum distance between the point and the line.
\end{problem}

\begin{problem}
  Use facts about the cross product to find the minimum distance
  between the point and the line.
\end{problem}

\begin{problem}
  Let:
  \[
  S(x,y) = (x-a)^2 + (y-b)^2
  \]
  for appropriate values of $a$ and $b$, and let $\vec{\vecl}(t)$ be a
  vector-valued function that parameterizes the line. Use $S$ and
  $\vecl$ to minimize the distance between the point and points on the
  line.
\end{problem}

\begin{problem}
  Let:
  \[
  S(x,y) = (x-a)^2 + (y-b)^2
  \]
  for appropriate values of $a$ and $b$. Use Lagrange multipliers to
  to find the minimum distance between the point and the line.
\end{problem}

\begin{problem}
  Now consider the point $(1,2,3)$ and the plane
  \[
  x + y+z = 0.
  \]
  Use the method of your choosing to find the minimum distance between
  the point and the plane. Share your method with a friend.
\end{problem}


\begin{problem}
  Now consider the point $(1,2,3)$ and the sphere
  \[
  x^2 + y^2+z^2 = 1.
  \]
  Use the method of your choosing find the minimum distance between
  the point and the sphere.  Share your method with a friend.
\end{problem}


\end{document}
