\documentclass{ximera}

\author{Jim Talamo}

\input{../../preamble.tex}

\outcome{Evaluate limits of functions of several variables.}

\begin{document}
\begin{exercise}

Several students are asked to determine if $\lim_{(x,y) \to (3,1)} \frac{x-2y}{3x+4y}$ exists and provide the responses below.

Student $1$ explains:

\begin{explanation}
We will analyze the function along different paths.

\begin{itemize}
\item If $x=0$, then $F(x,y) = F(0,y) = \frac{-2y}{4y} = -\frac{1}{2}$.
\item If $y=0$, then $F(x,y) = F(x,0) = \frac{x}{3x} = \frac{1}{3}$.
\end{itemize}

Since the function approaches different values along different paths, the limit does not exist.
\end{explanation}

Student $2$ explains:

\begin{explanation}
We will analyze the function along paths of the form $y=mx$.  Along $y=mx$, we have

\[
F(x,y)=F(x,mx) = \frac{x-2mx}{3x-4mx} = \frac{1-2m}{3-4m}.
\]

Since the value the function approaches depends on $m$, the limit does not exist.
\end{explanation}

Which student, if either, is correct?

\begin{multipleChoice}
\choice{Student $1$ is correct, but Student $2$ is wrong.}
\choice{Student $1$ is wrong, but Student $2$ is correct.}
\choice{Both students are correct.}
\choice{Both students are wrong.}
\end{multipleChoice}

\begin{exercise}
Let's explore what went wrong.  In order to determine if the given limit exists, we need to find a path along which $\point{x,y} \to \point{\answer{3},\answer{1}}$.

\begin{itemize}
\item Let's consider Student $1$ first.

In their argument, student $1$ attempted to use a path along which $x=0$.  Can $\point{x,y}$ approach $\point{3,1}$ along this path? \wordChoice{\choice{Yes}\choice[correct]{No}}

Student $1$ also attempted to use a path along which $y=0$.  Can $\point{x,y}$ approach $\point{3,1}$ along this path? \wordChoice{\choice{Yes}\choice[correct]{No}}

\item Now let's look at Student $2$'s argument.  In the argument, Student $2$ tried to analyze the function the most general straight line path.  However, note that $y=mx$ is the most general straight line path through $(0,0)$; there's only one value for $m$ for which $y=mx$ passes through $(3,1)$, and that is $m= \answer{3}$.

The most general straight line path through $(3,1)$ is found using 

\[
y-y_0 = m(x-x_0),
\] 
which in this case is $y-\answer{1} = m\left(x-\answer{3}\right)$.
\end{itemize}

That being said, how should we attempt to determine if $\Lim{x,y}{3,1} \frac{x-2y}{3x+4y}$ exists? If we simply try direct substitution, we find that the limit actually \emph{does} exist and

\[
\Lim{x,y}{3,1} \frac{x-2y}{3x+4y} = \answer{\frac{1}{13}}.
\]

\begin{feedback}[correct]
There are many techniques that can be used to determine whether limits exist or not, but remember to take a step back, take a deep breath, and think about limits before you start trying to evaluate them.

\end{feedback}
 

\end{exercise}

 \end{exercise}
\end{document}
