\documentclass{ximera}

\input{../preamble.tex}

\outcome{Understand what is meant by the form of a limit.}
\outcome{Calculate limits of the form \zeroOverZero.}
\outcome{Identify determinate and indeterminate forms.}
\outcome{Distinguish between determinate and indeterminate forms.}
\outcome{Discuss why infinity is not a number.}

\title[Dig-In:]{Limits of the form zero over zero}

\begin{document}
\begin{abstract}
  We want to solve limits that cannot be solved directly using the
  Limit Laws.
\end{abstract}

\maketitle

In the last section, we were interested in the limits we could compute
using continuity and the limit laws. What about limits that cannot be
directly computed using these methods? Let's think about an example. Consider
\[
\lim_{x\to 2}\frac{x^2-3x+2}{x-2}.
\]
Here 
\[
\lim_{x\to 2}\left(x^2-3x+2\right) = 0\qquad\text{and}\qquad \lim_{x\to
  2}\left(x-2\right) = 0
\]
in light of this, you may think that the limit is zero. \textbf{Not so
  fast}. This limit is of an \textit{indeterminate form}. What does
this mean? Read on, young mathematician.

\begin{definition}
  A limit
  \[
  \lim_{x\to a} \frac{f(x)}{g(x)}
  \]
  is said to be of the form \zeroOverZero\ if
  \[
  \lim_{x\to a} f(x) = 0\qquad\text{and}\qquad \lim_{x\to a} g(x) =0.
  \]
\end{definition}

\begin{question}
  Which of the following limits are of the form \zeroOverZero?
  \begin{selectAll}
    \choice[correct]{$\lim_{x\to 0}\frac{\sin(x)}{x}$}
    \choice{$\lim_{x\to 0}\frac{\cos(x)}{x}$}
    \choice{$\lim_{x\to 0}\frac{x^2-3x+2}{x-2}$}
    \choice[correct]{$\lim_{x\to 2}\frac{x^2-3x+2}{x-2}$}
    \choice{$\lim_{x\to 3}\frac{x^2-3x+2}{x-3}$}
  \end{selectAll}
\end{question}

\begin{warning}
  The symbol \zeroOverZero\ is \textbf{not} the number $0$ divided by
  $0$. It is simply short-hand and means that a limit $\lim_{x\to a}
  \frac{f(x)}{g(x)}$ has the property that
  \[
  \lim_{x\to a} f(x) = 0\qquad\text{and}\qquad \lim_{x\to a} g(x) =0.
  \]
\end{warning}


Let's consider an example that we looked at in the previous section:

\begin{example}
  Compute:
  \[
  \lim_{x\to 2}\frac{x^2-3x+2}{x-2}
  \]
  \begin{explanation}
  This limit is of the form \zeroOverZero. However, note that if we
  assume $x\ne 2$, then we can write
    \[
    \frac{x^2-3x+2}{x-2} = \frac{(x-2)(x-1)}{(x-2)} = x-1.
    \]
    \begin{image}
      \begin{tikzpicture}
        \begin{axis}[
            domain=-2:4,
            width=2.5in,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={-2,...,4},
            ytick={-3,...,3},
          ]
	  \addplot [very thick, penColor, smooth] {x-1};
        \end{axis}
        \node [penColor] at (2,-.75) {$y= x-1$};
      \end{tikzpicture}
      \qquad
      \begin{tikzpicture}
	\begin{axis}[
            domain=-2:4,
            width=2.5in,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={-2,...,4},
            ytick={-3,...,3},
          ]
	  \addplot [very thick, penColor, smooth] {x-1};
          \addplot[color=penColor,fill=background,only marks,mark=*] coordinates{(2,1)};  %% open hole
        \end{axis}
        \node [penColor] at (2,-.5) {$y=\frac{x^2-3x+2}{x-2}$};
      \end{tikzpicture}
    \end{image}
    This means
    \[
    \lim_{x\to 2}\frac{x^2-3x+2}{x-2} = \lim_{x\to 2} (x-1).
    \]
    Ah! $\lim_{x\to 2} (x-1) =1$. Hence
    \[
    \lim_{x\to 2}\frac{x^2-3x+2}{x-2} = \lim_{x\to 2} (x-1) = 1.
    \]
  \end{explanation}
\end{example} 



%% \begin{definition}
%%   Consider
%%   \[
%%   \lim_{x\to a} f(x).
%%   \]
%%   The \dfn{form} of this limit is a symbol which is not a number that
%%   results when $a$ is plugged directly into $f(x)$ and the result is
%%   simplified as much as possible.  Because a form is not a number, we
%%   will not say that a limit equals its form.  We will use the symbol
%%   $\sim$ to say that a limit has a certain form.
%% \end{definition} 

Knowing this form is informative.  Since we know that both the
numerator and the denominator are approaching $0$ as $x$ approaches
$2$, we suspect that there might be a common factor in both the top
and the bottom which are going to $0$.  If we could find that factor
and cancel it, we would have a another (easier to work with) function
whose limit is equal to the limit of the original function.  In fact,
this is exactly what happened in this example when we factored out the
$(x-2)$ term.

For now, lets consider some more examples of the form \zeroOverZero.

\begin{example}
  Compute:
  \[
  \lim_{x\to1}\frac{x-1}{x^2+2x-3}
  \]
  \begin{explanation}
    First note that
    \[
    \lim_{x\to1}\left(x^2+2x-3\right)=0 \qquad\text{and}\qquad  \lim_{x\to1}\left(x^2+2x-3\right) = 0
    \]
    Hence this limit is of the form \zeroOverZero.  This tells us we
    can likely cancel a factor going to $0$ out of the numerator and
    denominator.  Since $(x-1)$ is a factor going to $0$ in the
    numerator, let's see if we can factor a $(x-1)$ out of the
    denominator as well.
    \begin{align*}
      \lim_{x\to1}\frac{x-1}{x^2+2x-3}&=\lim_{x\to1}\frac{x-1}{(x-1)(x+3)} \\
      &=\lim_{x\to1}\frac{1}{x+3)}\\
      &=\frac{1}{4}.
    \end{align*}
  \end{explanation}
\end{example}


\begin{example}
  Compute:
  \[
  \lim_{x\to 1} \frac{\frac{1}{x+1}-\frac{3}{x+5}}{x-1}
  \]
\begin{explanation}
  We begin by finding the form of this limit by looking at the limits
  of the numerator and denominator separately.
  \[
  \lim_{x\to 1}\left(\frac{1}{x+1}-\frac{3}{x+5}\right)=0\qquad\text{and}\qquad\lim_{x\to 1}\left(x-1\right)=0.
  \]
This tells us that this limit is of the form \zeroOverZero\ and we can
probably factor a term going to $0$ out of both the numerator and
denominator.  We suspect from looking at the denominator that this
term is $(x-1)$.  Unfortunately, it is not immediately obvious how to
factor an $(x-1)$ out of the numerator.  In order to simplify the
numerator, we will find a common denominator for the fractions in the
numerator.
\begin{align*}
\lim_{x\to 1} \frac{\frac{1}{x+1}-\frac{3}{x+5}}{x-1}  &= \lim_{x\to 1} \frac{\frac{1}{x+1} \cdot \frac{x+5}{x+5}-\frac{3}{x+5} \cdot \frac{x+1}{x+1}}{x-1}\\
&= \lim_{x\to 1} \frac{\frac{(x+5)-3(x+1)}{(x+1)(x+5)}}{x-1}\\
&= \lim_{x\to 1}\frac{(x+5)-3(x+1)}{(x+1)(x+5)} \cdot \frac{1}{x-1}\\
&= \lim_{x\to 1}\frac{(x+5)-3(x+1)}{(x+1)(x+5)(x-1)}
\end{align*}

Now we will multiply out the numerator.  Note that we do not want to
multiply out the denominator because we already have an $(x-1)$
factored out of the denominator and that was the goal.

\begin{align*}
\lim_{x\to 1}\frac{(x+5)-3(x+1)}{(x+1)(x+5)(x-1)} &= \lim_{x\to 1}\frac{x+5-3x-3}{(x+1)(x+5)(x-1)} \\
&= \lim_{x\to 1}\frac{-2x+2}{(x+1)(x+5)(x-1)}\\
&= \lim_{x\to 1}\frac{-2(x-1)}{(x+1)(x+5)(x-1)}\\
&= \lim_{x\to 1}\frac{-2}{(x+1)(x+5)}.
\end{align*}
  
Hence

\[
\lim_{x\to 1} \frac{\frac{1}{x+1}-\frac{3}{x+5}}{x-1}=\lim_{x\to 1}\frac{-2}{(x+1)(x+5)} = \frac{-2}{((1)+1)((1)+5)} = \frac{-1}{6}
\]
\end{explanation}
\end{example}

And one more example:

\begin{example}
  Compute:
  \[
  \lim_{x\to-1} \frac{\sqrt{x+5}-2}{x+1}
  \]

\begin{explanation} 
  Note that 
  \[
  \lim_{x\to-1} \left(\sqrt{x+5}-2\right)=0\qquad\text{and}\qquad\lim_{x\to -1} \left(x+1\right) =0.
  \]
  This tells us this limit is of the form \zeroOverZero\ and that we
  can probably factor a term going to $0$ out of both the numerator
  and denominator.  We suspect from looking at the denominator that
  this term is $(x+1)$.  Unfortunately, it is not immediately obvious
  how to factor an $(x+1)$ out of the numerator.
 
  We will use an algebraic technique called \dfn{multiplying by the
    conjugate}.  This technique is useful when you are trying to
  simplify an expression that looks like
  \[
  \sqrt{\text{something} \pm \text{something else}}.
  \]
  It takes advantage of the difference of squares rule that
  \[
  a^2-b^2=(a-b)(a+b).
  \]
  In our case, we will use $a=\sqrt{x+5}$ and $b=2$.  Then, we have
  the $(a-b)$ term.  If we introduce the $(a+b)$ term by multiplying
  both the numerator and denominator by $(\sqrt{x+5}+2)$, we will be
  able to simplify our result using the differences of squares.
 
\begin{align*}
\lim_{x\to-1} \frac{\sqrt{x+5}-2}{x+1}&=
\lim_{x\to-1} \frac{(\sqrt{x+5}-2)}{(x+1)} \cdot \frac{(\sqrt{x+5}+2)}{(\sqrt{x+5}+2)} \\
&=\lim_{x\to-1} \frac{(\sqrt{x+5})^2-(2)^2}{(x+1)(\sqrt{x+5}+2)} \\
&=\lim_{x\to-1} \frac{x+5-4}{(x+1)(\sqrt{x+5}+2)} \\
&=\lim_{x\to-1} \frac{(x+1)}{(x+1)(\sqrt{x+5}+2)} \\
&=\lim_{x\to-1} \frac{1}{\sqrt{x+5}+2}\\
&= \frac{1}{\sqrt{-1+5}+2}\\
&=\frac{1}{4}.
\end{align*}
\end{explanation}
\end{example}

All the examples in this section are limits of the form \zeroOverZero.
When you come across a limit of the form \zeroOverZero, you should try
to use various algebraic techniques to come up with a continuous
function whose limit you can evaluate.

Notice that we solved multiple examples of limits of the form
\zeroOverZero\ and we got different answers each time.  This tells us
that just knowing the form of the limit is \zeroOverZero\ is not enough
to compute the limit. The moral of the story is
\begin{center}
  \textbf{Limits of the form \zeroOverZero\ can take any value.}
\end{center}

\begin{definition}
Forms that give us no information about the value of the limit are
called \dfn{indeterminate forms}.

Forms that give information about the value of the limit are called
\dfn{determinate forms}.
\end{definition}  

Finally, you may find it distressing that we introduced a form, namely
\zeroOverZero, only to end up saying they give no information on the
value of the limit. But you see, this is precisely what makes
indeterminate forms interesting\dots they're a mystery!



\end{document}
