\documentclass{ximera}

\input{../preamble.tex}

\title{The Almost Equal Theorem}

\begin{document}
\begin{abstract}
  We want to solve limits that cannot be solved directly using the
  Limit Laws.
\end{abstract}

\maketitle

%This chapter will consist of three dig in's 1. The Almost Equal Theorem 2. Limits of the form non-zero over zero 3. Determinate and indeterminate forms 

In the last section, we were interested in the limits we could compute
using continuity and the limit laws. What about the limits that cannot
be directly computed using these methods?  Is it possible to determine
these limits?  Often times, it is.  We will use two important tools to
find these limits.  The first is the following theorem:

\begin{theorem}[Almost Equal Theorem]
Let $f$ and $g$ be two functions such that $f(x)=g(x)$ for every $x$
in some open interval $I$ except for at a finite set of points in $I$.
Then,
\[
\lim_{x\to a} f(x)= \lim_{x\to a} g(x),
\]
provided at least one of these limits exists.  We call $f$ and $g$
\dfn{almost equal functions} in the context of considering the limit
as $x$ approaches $a$.
\end{theorem}

Here's the upshot: If two functions agree at every point near $a$,
then they have the same limit at $a$.
\begin{image}
      \begin{tikzpicture}
	\begin{axis}[
            domain=-2:4,
            ymin=-3,
            ymax=3,
            width=2.5in,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={-2,...,4},
            ytick={-3,...,3},
          ]
	  \addplot [very thick, penColor, smooth] {-.5*x^3 + 1.6*x^2-.2*x -1};
          \addplot[color=penColor,fill=background,only marks,mark=*] coordinates{(2,1)};  %% open hole
          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(2,-1.5)};  %% closed hole
        \end{axis}
      \end{tikzpicture}\qquad
      \begin{tikzpicture}
	\begin{axis}[
            domain=-2:4,
            ymin=-3,
            ymax=3,
            width=2.5in,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={-2,...,4},
            ytick={-3,...,3},
          ]
	  \addplot [very thick, penColor, smooth] {-.5*x^3 + 1.6*x^2-.2*x -1};
          \addplot[color=penColor,fill=background,only marks,mark=*] coordinates{(.4,-.86)};  %% open hole
          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(.4,2)};  %% closed hole
        \end{axis}
      \end{tikzpicture}
\end{image}
\begin{question}
  Are the two functions that produce the graphs above almost equal on
  the interval $(-1,3)$?
    \begin{multipleChoice}
      \choice[correct]{Yes}
      \choice{No}
    \end{multipleChoice}
  \begin{feedback}
    Since these functions are equal on the open interval $(-1,3)$,
    except for at two disconnected points, they are almost equal on
    the interval.
  \end{feedback}
\end{question}
This should make sense. We already know that the limit does not
``see'' what is happening exactly at $a$, it only sees what is
happening nearer and nearer to $a$.

Let's consider an example that we looked at in the previous section:

\begin{example}
  Compute:
  \[
  \lim_{x\to 2}\frac{x^2-3x+2}{x-2}
  \]
  \begin{explanation}
    In the previous section, we decided that we could not find this
    limit using the Quotient Law because the denominator is equal to 0
    at $x=2$.  Can we use the Almost Equal Thoerem?

    Well note that if we assume $x\ne 2$, then we can write
    \[
    \frac{x^2-3x+2}{x-2} = \frac{(x-2)(x-1)}{(x-2)} = x-1.
    \]
    Since this relation is true provided $x\ne 2$, we have that
    \[
    \frac{x^2-3x+2}{x-2} \text{ is almost equal to }x-1.
    \]
    \begin{image}
      \begin{tikzpicture}
        \begin{axis}[
            domain=-2:4,
            width=2.5in,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={-2,...,4},
            ytick={-3,...,3},
          ]
	  \addplot [very thick, penColor, smooth] {x-1};
        \end{axis}
        \node [penColor] at (2,-.75) {$y= x-1$};
      \end{tikzpicture}
      \qquad
      \begin{tikzpicture}
	\begin{axis}[
            domain=-2:4,
            width=2.5in,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={-2,...,4},
            ytick={-3,...,3},
          ]
	  \addplot [very thick, penColor, smooth] {x-1};
          \addplot[color=penColor,fill=background,only marks,mark=*] coordinates{(2,1)};  %% open hole
        \end{axis}
        \node [penColor] at (2,-.5) {$y=\frac{x^2-3x+2}{x-2}$};
      \end{tikzpicture}
    \end{image}
    This means
    \[
    \lim_{x\to 2}\frac{x^2-3x+2}{x-2} = \lim_{x\to 2} (x-1).
    \]
    Ah! $\lim_{x\to 2} (x-1) =1$. Hence
    \[
    \lim_{x\to 2}\frac{x^2-3x+2}{x-2} = \lim_{x\to 2} (x-1) = 1.
    \]
  \end{explanation}
\end{example} 

This trick does not always work.  For example, consider
\[
\lim_{x\to 3}\frac{x^2-3x+2}{x-3}.
\]
First note that if we plug $3$ into the denominator of this fraction,
the denominator will equal $0$.  Therefore, we cannot use the Quotient
Law to solve this problem.  Factoring will give us
$\frac{x^2-3x+2}{x-3}=\frac{(x-2)(x-1)}{x-3}$ and nothing cancels.  We
seem to be stuck.  We need to develop some new tools and come back to
this example.

In order to give us a concrete way of talking about the differences
between these two examples, we are going to introduce a new idea, the
\textit{form} of a limit.

\begin{definition}
  A limit
  \[
  \lim_{x\to a} \frac{f(x)}{g(x)}
  \]
  is said to be of the form \zeroOverZero\ if
  \[
  \lim_{x\to a} f(x) = 0\qquad\text{and}\qquad \lim_{x\to a} g(x) =0.
  \]
\end{definition}

\begin{question}
  Which of the following limits are of the form \zeroOverZero?
  \begin{multipleChoice}
    \choice[correct]{$\lim_{x\to 0}\frac{\sin(x)}{x}$}
    \choice{$\lim_{x\to 0}\frac{\cos(x)}{x}$}
    \choice{$\lim_{x\to 0}\frac{x^2-3x+2}{x-2}$}
    \choice[correct]{$\lim_{x\to 2}\frac{x^2-3x+2}{x-2}$}
    \choice{$\lim_{x\to 3}\frac{x^2-3x+2}{x-3}$}
  \end{multipleChoice}
\end{question}


%% \begin{definition}
%%   Consider
%%   \[
%%   \lim_{x\to a} f(x).
%%   \]
%%   The \dfn{form} of this limit is a symbol which is not a number that
%%   results when $a$ is plugged directly into $f(x)$ and the result is
%%   simplified as much as possible.  Because a form is not a number, we
%%   will not say that a limit equals its form.  We will use the symbol
%%   $\sim$ to say that a limit has a certain form.
%% \end{definition} 

Consider the first example.
\[
\lim_{x\to 2}\frac{x^2-3x+2}{x-2}.
\]
Since
\[
\lim_{x\to 2}\left(x^2-3x+2\right) = 0\qquad\text{and}\qquad \lim_{x\to
  2}\left(x-2\right) = 0
\]
we see that $\lim_{x\to 2}\frac{x^2-3x+2}{x-2}$ is of the form
\zeroOverZero.

Knowing this form is informative.  Since we know that both the
numerator and the denominator are approaching $0$ as $x$ approaches
$2$, we suspect that there might be a common factor in both the top
and the bottom which are going to $0$.  If we could find that factor
and cancel it, we would have a function which is almost equal to the
original function and whose limit we may be able to find using the
Limit Laws.  In fact, this is exactly what happened in this example
when we factored out the $(x-2)$ term.

For now, lets consider some more examples of the form $\frac{0}{0}$.

\begin{example}
  Compute:
  \[
  \lim_{x\to1}\frac{x^2+2x-3}{x-1}
  \]
  \begin{explanation}
    We begin by finding the form of this limit.
    \[
    \lim_{x\to1}\frac{x^2+2x-3}{x-1} \sim \frac{(1)^2+2(1)-3}{(1)-1} \sim
    \frac{0}{0}.
    \]
    This tells us we can likely find an almost equal function if we can
    cancel a term going to $0$ out of the numerator and denominator.  Since
    $(x-1)$ is a term going to $0$ in the denominator, let's see if we can
    factor a $(x-1)$ out of the numerator as well.
    \begin{align*}
      \lim_{x\to1}{x^2+2x-3\over x-1}&=\lim_{x\to1}{(x-1)(x+3)\over x-1} \\
      &=\lim_{x\to1}(x+3)=4
    \end{align*}
  \end{explanation}
\end{example}



Now consider again
\[
\lim_{x\to 3}\frac{x^2-3x+2}{x-3} \sim \frac{(3)^2-3(3)+2}{(3)-3} \sim \frac{5}{0}.
\]
This limit has the form $\frac{5}{0}$, which immediately tell us that
there cannot be a factor $(x-3)$ which can be factored out of the
denominator.  In fact, it tells us more than that.  We will look at
how to solve limits with a form $\frac{\text{non-zero}}{\text{zero}}$
in the next dig in.


\begin{example}
  Compute:
  \[
  \lim_{x\to1}\frac{x^2+2x-4}{x-2}
  \]
\begin{explanation}
  We begin by finding the form of this
  limit.
  \[
  \lim_{x\to1}\frac{x^2+2x-4}{x-2} \sim\frac{(1)^2+2(1)-4}{(1)-2} \sim \frac{-1}{-1}.
  \]
  This is a numeber, not an indeterminate form! We have discovered
  that this is a limit for which we could have and should have used
  the continuity of rational functions on their domains to compute
  this limit.  All we have to do is change our $\sim$ symbols to $=$
  symbols to complete the problem.
  \[
  \lim_{x\to1}\frac{x^2+2x-4}{x-2} = \frac{(1)^2+2(1)-4}{(1)-2} =
  \frac{-1}{-1} =1.
  \]
\end{explanation}
\end{example}

\begin{example}
  Compute:
  \[
  \lim_{x\to 1} \frac{\frac{1}{x+1}-\frac{3}{x+5}}{x-1}
  \]
\begin{explanation}
  We begin by finding the form of this limit.
  \[
  \lim_{x\to 1} \frac{\frac{1}{x+1}-\frac{3}{x+5}}{x-1} \sim
  \frac{\frac{1}{(1)+1}-\frac{3}{(1)+5}}{(1)-1} \sim
  \frac{\frac{1}{2}-\frac{1}{2}}{0} \sim \frac{0}{0}
  \]
This tells us that we can probably factor a term going to $0$ out of
both the numerator and denominator.  We suspect from looking at the
denominator that this term is $(x-1)$.  Unfortunately, it is not
immediately obvious how to factor an $(x-1)$ out of the numerator.  In
order to simplify the numerator, we will find a common denominator for
the fractions in the numerator.
\begin{align*}
\lim_{x\to 1} \frac{\frac{1}{x+1}-\frac{3}{x+5}}{x-1}  &= \lim_{x\to 1} \frac{\frac{1}{x+1} \cdot \frac{x+5}{x+5}-\frac{3}{x+5} \cdot \frac{x+1}{x+1}}{x-1}\\
&= \lim_{x\to 1} \frac{\frac{(x+5)-3(x+1)}{(x+1)(x+5)}}{x-1}\\
&= \lim_{x\to 1}\frac{(x+5)-3(x+1)}{(x+1)(x+5)} \cdot \frac{1}{x-1}\\
&= \lim_{x\to 1}\frac{(x+5)-3(x+1)}{(x+1)(x+5)(x-1)}
\end{align*}

Now we will multiply out the numerator.  Note that we do not want to
multiply out the denominator because we already have an $(x-1)$
factored out of the denominator and that was the goal.

\begin{align*}
\lim_{x\to 1}\frac{(x+5)-3(x+1)}{(x+1)(x+5)(x-1)} &= \lim_{x\to 1}\frac{x+5-3x-3}{(x+1)(x+5)(x-1)} \\
&= \lim_{x\to 1}\frac{-2x+2}{(x+1)(x+5)(x-1)}\\
&= \lim_{x\to 1}\frac{-2(x-1)}{(x+1)(x+5)(x-1)}\\
&= \lim_{x\to 1}\frac{-2\cancel{(x-1)}}{(x+1)(x+5)\cancel{(x-1)}}\\
&= \lim_{x\to 1}\frac{-2}{(x+1)(x+5)}.
\end{align*}
  
Now we finally have an almost equal function that is continuous at
$x=1$.

\[
\lim_{x\to 1}\frac{-2}{(x+1)(x+5)} = \frac{-2}{((1)+1)((1)+5)} = \frac{-2}{12} = \frac{-1}{6}
\]
\end{explanation}
\end{example}

\begin{example}
  Compute:
  \[
  \lim_{x\to-1} \frac{\sqrt{x+5}-2}{x+1}
  \]

\begin{explanation} 
  We begin by finding the form of this limit.
  \[
  \lim_{x\to-1} \frac{\sqrt{x+5}-2}{x+1} \sim \frac{\sqrt{(-1)+5}-2}{(-1)+1} \sim \frac{\sqrt{4}-2}{0} \sim \frac{0}{0}.
  \]
  This tells us that we can probably factor a term going to $0$ out of
  both the numerator and denominator.  We suspect from looking at the
  denominator that this term is $(x+1)$.  Unfortunately, it is not
  immediately obvious how to factor an $(x+1)$ out of the numerator.
 
  We will use an algebraic technique called \textbf{multiplying by the
    conjugate}.  This technique is useful when you are trying to
  simplify an expression that looks like
  \[
  \sqrt{\text{something} \pm \text{something else}}.
  \]
  It takes advantage of the difference of squares rule that
  \[
  a^2-b^2=(a-b)(a+b).
  \]
  In our case, we will use $a=\sqrt{x+5}$ and $b=2$.  Then, we have
  the $(a-b)$ term.  If we introduce the $(a+b)$ term by multiplying
  both the numerator and denonimator by $(\sqrt{x+5}+2)$, we will be
  able to simpify our result using the differences of squares.
 
\begin{align*}
\lim_{x\to-1} \frac{\sqrt{x+5}-2}{x+1}&=
\lim_{x\to-1} \frac{(\sqrt{x+5}-2)}{(x+1)} \cdot \frac{(\sqrt{x+5}+2)}{(\sqrt{x+5}+2)} \\
&=\lim_{x\to-1} \frac{(\sqrt{x+5})^2-(2)^2}{(x+1)(\sqrt{x+5}+2)} \\
&=\lim_{x\to-1} \frac{x+5-4}{(x+1)(\sqrt{x+5}+2)} \\
&=\lim_{x\to-1} \frac{(x+1)}{(x+1)(\sqrt{x+5}+2)} \\
&=\lim_{x\to-1} \frac{\cancel{(x+1)}}{\cancel{(x+1)}(\sqrt{x+5}+2)} \\
&=\lim_{x\to-1} \frac{1}{\sqrt{x+5}+2}\\
&= \frac{1}{\sqrt{-1+5}+2}\\
&=\frac{1}{4}.
\end{align*}
\end{explanation}
\end{example}

All the examples in this section use the Almost Equal Theorem to
compute limits of the form $\frac{0}{0}$.  When you come across a
limit of the form $\frac{0}{0}$, you should try to use various algebra
techniques to come up with an almost equal function whose limit you
can evaluate.

Notice that we solved multiple examples of limits of the form
$\frac{0}{0}$ and we got different answers each time.  This tells us
that just knowing the form of the limit is $\frac{0}{0}$ is not enough
to compute the limit.

\begin{definition}
Forms which do not give us enough information to know the answer to
the limit from just looking at the form are called \dfn{indeterminate
  forms}.  There are some forms which we will see in future sections
which will be able to tell us the answer to the limit from just
looking at the form. These forms are called \dfn{determinate forms}.
\end{definition}  

The form $\frac{0}{0}$ is an indeterminate form.




\end{document}
