\documentclass{ximera}

\outcome{Define the average value of a function.}
\outcome{Find the average value of a function.}
\outcome{State the Mean Value Theorem for integrals.}
\outcome{Use the Mean Value Theorem for integrals.}

\input{../preamble.tex}

\title[Dig-In:]{Applications of integrals}
\begin{document}
\begin{abstract}
We give more contexts to understand integrals.
\end{abstract}
\maketitle



\section{Velocity and displacement, speed and distance}

Some values include ``direction'' that is relative to some fixed point.

\begin{definition}\hfill
  \begin{itemize}
  \item $v(t)$ is the \dfn{velocity} of an object at time $t$. This
    represents the ``change in position'' at time $t$.
  \item $s(t)$ is the \dfn{position} of an object at time $t$. This
    gives location with respect to the origin. If we can assume that
    $s(a) = 0$, then
    \[
    s(t) = \int_a^t v(x) \d x.
    \]
  \item $s(b) -s(a)$ is the \dfn{displacement}, the distance between the
    starting and finishing locations.
  \end{itemize}
\end{definition}

On the other hand \textit{speed} and \textit{distance} are values
without ``direction.''

\begin{definition}\hfil
  \begin{itemize}
  \item $|v(t)|$ is the \dfn{speed}.
  \item $\int_a^b |v(t)| \d t$ is the \dfn{distance} traveled.
  \end{itemize}
\end{definition}

\begin{question}
  Consider a particle whose velocity at time $t$ is given by $v(t) = \sin(t)$.

\begin{image}
  \begin{tikzpicture}[
      declare function = {f(\x) = sin(deg(\x));}]
    \begin{axis}[  
        domain=-.2:7, xmin =-.2,xmax=7,ymax=1.2,ymin=-1.2,
        width=6in,
        height=3in,xtick={3.141,6.282},
        xticklabels={$\pi$, $2\pi$},
        ytick style={draw=none},
        yticklabels={},
        axis lines=center, xlabel=$x$, ylabel=$y$,
        every axis y label/.style={at=(current axis.above origin),anchor=south},
        every axis x label/.style={at=(current axis.right of origin),anchor=west},
        axis on top,
      ]
               \addplot [very thick,penColor, smooth] {f(x)};
    \end{axis}
  \end{tikzpicture}
\end{image}
What is the displacement of the particle from $t=0$ to $t=\pi$? That
is, compute:
\[
\int_0^\pi \sin(t) \d t \begin{prompt}= \answer[given]{2}\end{prompt}
\]
What is the displacement of the particle from $t=0$ to $t=2\pi$? That is, compute:
\[
\int_0^{2\pi} \sin(t) \d t \begin{prompt}= \answer[given]{0}\end{prompt}
\]
What is the distance traveled by the particle from $t=0$ to $t=\pi$?
That is, compute:
\[
\int_0^\pi  \left| \sin(t) \right|  \d t \begin{prompt}= \answer[given]{2}\end{prompt}
\]
What is the distance traveled by the particle from $t=0$ to $t=2\pi$?
That is, compute:
\[
\int_0^{2\pi} \left| \sin(t) \right| \d t \begin{prompt}=\answer[given]{4}\end{prompt}
\]
\end{question}



\section{Average value}

Conceptualizing definite integrals as ``signed area'' works great as
long as one can actually visualize the ``area.'' In some cases, a
better metaphor for integrals comes from the idea of \textit{average
  value}.  Looking back to your days as an even younger mathematician,
you may recall the idea of an \textit{average}:
\[
\frac{f_1+f_2+\dots+f_n}{n} = \frac{1}{n}\sum_{k=1}^n f_i
\]
If we want to know the average value of a function, a naive approach
might be to partition the interval $[a,b]$ into $n$ equally spaced
subintervals, 
\[
a=x_0 < x_1 < \dots < x_{n}=b
\]
and choose any $x_k^*$ in $[x_i,x_{i+1}]$. The average of $f(x_1^*)$,
$f(x_2^*)$, \dots, $f(x_n^*)$ is:
\[
\frac{f(x_1^*) + f(x_2^*) + \dots + f(x_n^*)}{n} = \frac1n\sum_{k=1}^n f(x_k^*).
\]
Multiply this last expression by $1 = \frac{(b-a)}{(b-a)}$:
\begin{align*}
  \frac1n\sum_{k=1}^n f(x_k^*) &= \sum_{k=1}^n f(x_k^*)\frac1n \\
  &= \sum_{k=1}^n f(x_k^*)\frac1n \frac{(b-a)}{(b-a)} \\
  &= \frac{1}{b-a} \sum_{k=1}^n f(x_k^*)\frac{b-a}n  \\
  &=\frac{1}{b-a} \sum_{k=1}^n f(x_k^*)\Delta x
\end{align*}
where $\Delta x = (b-a)/n$.  Ah! On the right we have a Riemann Sum!
Now take the limit as $n\to\infty$:
\[
\lim_{n\to\infty} \frac{1}{b-a} \sum_{k=1}^n f(x_k^*)\Delta x = \frac{1}{b-a} \int_a^b f(x)\d x.
\]
This leads us to our next definition:

\begin{definition}
  Let $f$ be continuous on $[a,b]$. The \dfn{average value} of $f$ on
  $[a,b]$ is given by
  \[
  \frac{1}{b-a}\int_a^b f(x)\d x.
  \]
\end{definition}

The average value of a function gives the height of a single rectangle whose area is equal to
\[
\int_a^b f(x) \d x
\]
\begin{image}
  \begin{tikzpicture}[
      declare function = {f(\x) = -sin(deg(\x)) + 3;}]
    \begin{axis}[  
        domain=-.2:7, xmin =-.2,xmax=7,ymax=5,ymin=-.7,
        width=6in,
        height=3in,xtick={1,6},
        xticklabels={$a$,$b$},
        ytick style={draw=none},
        yticklabels={},
        axis lines=center, xlabel=$x$, ylabel=$y$,
        every axis y label/.style={at=(current axis.above origin),anchor=south},
        every axis x label/.style={at=(current axis.right of origin),anchor=west},
        axis on top,
      ]
      \foreach \rectnumber in {1}
               {
                 \addplot [draw=penColor,fill=fillp] plot coordinates
                          {({1+(\rectnumber - 1) * 5},{3.08})
                            ({1+(\rectnumber) * 5},{3.08})} \closedcycle;
               };
               \addplot [very thick,penColor, smooth] {f(x)};
               \node[black] at (axis cs:3.5,1.5) {\huge$\int_a^b f(x) \d x$};
    \end{axis}
  \end{tikzpicture}
\end{image}

An application of this definition is given in the next example.


\begin{example}
An object moves back and forth along a straight line with a velocity
given by $v(t) = (t-1)^2$ on $[0,3]$, where $t$ is measured in seconds
and $v(t)$ is measured in ft/s.

What is the average velocity of the object?
\begin{explanation}
By our definition, the average velocity is:
\begin{align*}
\frac{1}{3-0}\int_0^3 (t-1)^2\d t &=\frac13 \int_0^3 t^2-2t+1\d t\\
&= \frac13\eval{\frac{t^3}{3}-t^2+t}_0^3\\
&= 1\text{ ft/s}.
\end{align*}
\end{explanation}
\end{example}



When we take the average of a finite set of values, it does not matter
how we order those values.  When we are taking the average value of a
function, however, we need to be more careful.

For instance, there are at least two different ways to make sense of a
vague phrase like ``The average height of a point on the unit semi
circle''
\begin{image}
  \begin{tikzpicture}[
      declare function = {f(\x) = sin(deg(\x));}]
    \begin{axis}[  
        domain=-3:3, xmin =-1.2,xmax=1.2,ymax=1.2,ymin=-.2,
        width=6in,
        height=3in,xtick={-1,1},
        xticklabels={$-1$, $1$},
        ytick style={draw=none},
        yticklabels={},
        axis lines=center, xlabel=$x$, ylabel=$y$,
        every axis y label/.style={at=(current axis.above origin),anchor=south},
        every axis x label/.style={at=(current axis.right of origin),anchor=west},
        axis on top,
      ]
              \addplot [penColor, very thick, smooth, domain=(0:180)] ({cos(x)},{sin(x)});
    \end{axis}
  \end{tikzpicture}
\end{image}
One way we can make sense of ``The average height of a point on the
unit semi circle'' is to compute the average value of the function
\[
f(x) =\sqrt{1-x^2}
\]
on the interval $[-1,1]$.

\begin{example}
  Compute the average value of the function
  \[
  f(x) =\sqrt{1-x^2}
  \]
  on the interval $[-1,1]$.
  \begin{explanation}
    By definition, we wish to compute
    \[
    \frac{1}{2}\int_{-1}^1 \sqrt{1-x^2} \d x.
    \]
    Computing this integral geometrically, we find the average value
    to be $\answer[given]{\frac{\pi}{4}}$.
  \end{explanation}
\end{example}
Another way we can make sense of ``The average height of a point on the
unit semi circle'' is the average value of the function
\[
g(\theta) =\sin(\theta)
\]
on $[0,\pi]$, since $\sin(\theta)$ is the height of the point on the
unit circle at the angle $\theta$.
\begin{example}
  Compute the average value of the function
  \[
  g(\theta) =\sin(\theta)
  \]
  on the interval $[0,\pi]$.
  \begin{explanation}
    By definition, we wish to compute
    \[
    \frac{1}{\pi} \int_0^\pi\sin(\theta) \d \theta.
    \]
    Computing this integral geometrically, we find the average value
    to be $\answer[given]{\frac{2}{\pi}}$.
  \end{explanation}
\end{example}

See if you can understand intuitively why the average using $f$ should
be larger than the average using $g$.  %% It is amusing to note that
%% since
%% \[
%% g(x) < f(x)
%% \]this
%% shows that $\pi^2>8$.



\section{Mean value theorem for integrals}

Just as we have a Mean Value Theorem for Derivatives, we also have a
Mean Value Theorem for Integrals.


\begin{theorem}[The Mean Value Theorem for integrals]
Let $f$ be continuous on $[a,b]$. There exists a value $c$ in $[a,b]$
such that
\[
\int_a^bf(x)\d x = f(c)(b-a).
\]
\end{theorem}

This is an \emph{existential} statement. The Mean Value Theorem for
Integrals tells us:
\begin{quote}
\textbf{The average value of a continuous function is in the range of
  the function.}
\end{quote}

We demonstrate the principles involved in this version of the Mean
Value Theorem in the following example.

\begin{example}%%%BADBAD we state that this is an existential thm, then we ask for the value!
  Consider $\int_0^\pi \sin x \d x$. Find a value $c$ guaranteed by the Mean Value Theorem.
  \begin{explanation}
    We first need to evaluate $\int_0^\pi \sin x\d x$.
    \[
    \int_0^\pi\sin x\d x =\eval{-\cos x}_0^\pi = 2.
    \]
    Thus we seek a value $c$ in $[0,\pi]$ such that $\pi\sin c =2$. 
    \[
    \pi\sin c = 2\ \ \Rightarrow\ \ \sin c = 2/\pi\ \ \Rightarrow\ \ c = \arcsin(2/\pi) \approx 0.69.
    \]
    
A graph of $\sin x$ is sketched along with a rectangle with height
$\sin (0.69)$ are pictured below. The area of the rectangle is the
same as the area under $\sin x$ on $[0,\pi]$.
\begin{image}
  \begin{tikzpicture}[
      declare function = {f(\x) = sin(deg(\x));}]
    \begin{axis}[  
        domain=-.2:3.3, xmin =-.2,xmax=3.3,ymax=1.2,ymin=-.2,
        width=6in,
        height=3in,xtick={3.141,6.282},
        xticklabels={$\pi$, $2\pi$},
        ytick style={draw=none},
        yticklabels={},
        axis lines=center, xlabel=$x$, ylabel=$y$,
        every axis y label/.style={at=(current axis.above origin),anchor=south},
        every axis x label/.style={at=(current axis.right of origin),anchor=west},
        axis on top,
      ]
	\addplot [draw=penColor,fill=fillp] plot coordinates
                          {({0},{2/pi})
                            ({pi},{2/pi})} \closedcycle;
               \addplot [very thick,penColor, smooth] {f(x)};

    \end{axis}
  \end{tikzpicture}
\end{image}
  \end{explanation}
\end{example}
\end{document}
